### 2017-01-16 README.Rmd
initial

---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](https://github.com/benmarwick/olympicdamboundaries/commits/master) 
[![minimal R version](https://img.shields.io/badge/R%3E%3D-`r as.character(getRversion())`-brightgreen.svg)](https://cran.r-project.org/)
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/) 
[![Travis-CI Build Status](https://travis-ci.org/benmarwick/olympicdamboundaries.png?branch=master)](https://travis-ci.org/benmarwick/olympicdamboundaries) 
[![Circle CI](https://circleci.com/gh/benmarwick/olympicdamboundaries.svg?style=shield&circle-token=:circle-token)](https://circleci.com/gh/benmarwick/olympicdamboundaries)
[![codecov.io](https://codecov.io/github/benmarwick/olympicdamboundaries/coverage.svg?branch=master)](https://codecov.io/github/benmarwick/olympicdamboundaries?branch=master) [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0001--7879--4531-green.svg)](http://orcid.org/0000-0001-7879-4531)

## Research compendium for a report on xxxx

### Compendium DOI:

http://dx.doi.org/xxxxxxx

The files at the URL above will generate the results as found in the publication. The files hosted at github.com/benmarwick/olympicdamboundaries are the development versions and may have changed since the report was published

### Authors  of this repository:

Ben Marwick (benmarwick@gmail.com)

### Published in:

Marwick, B,  xxxxx

### Overview of contents

This repository is our research compendium for our analysis of xxxx. The compendium contains all data, code, and text associated with the publication. The `Rmd` files in the `analysis/paper/` directory contain details of how all the analyses reported in the paper were conducted, as well as instructions on how to rerun the analysis to reproduce the results. The `data/` directory in the `analysis/` directory contains all the raw data.

### The supplementary files

The `analysis/` directory contains:

- the manuscript as submitted (in MS Word format) and it's Rmd source file
- all the data files (in CSV format, in the `data/` directory)    
- supplementary information source files (in R markdown format) and executed versions
- all the figures that are included in the paper (in the `figures/` directory)

### The R package

This repository is organized as an R package. There are no actual R functions in this package - all the R code is in the Rmd file. I simply used the R package structure to help manage dependencies, to take advantage of continuous integration for automated code testing, and so I didn't have to think too much about how to organise the files.

To download the package source as you see it on GitHub, for offline browsing, use this line at the shell prompt (assuming you have Git installed on your computer):

```{r eval=FALSE}
git clone https://github.com/benmarwick/olympicdamboundaries.git
```

Once the download is complete, open the `olympicdamboundaries.Rproj` in RStudio to begin working with the package and compendium files.

The package has a number of dependencies on other R packages, and programs outside of R. These are listed at the bottom of this README. Installing these can be time-consuming and complicated, so we've done two things to simpify access to the compendium. First is the packrat directory, which contains the source code for all the packages we depend on. If all works well, these will be installed on your computer when you open `olympicdamboundaries.Rproj` in RStudio. Second is our Docker image that includes all the necessary software, code and data to run our analysis. The Docker image may give a quicker entry point to the project, and is more self-contained, so might save some fiddling with installing things.

### The Docker image

A Docker image is a lightweight GNU/Linux virtual computer that can be run as a piece of software on Windows and OSX (and other Linux systems). To capture the complete computational environment used for this project we have a Dockerfile that specifies how to make the Docker image that we developed this project in. The Docker image includes all of the software dependencies needed to run the code in this project, as well as the R package and other compendium files. To launch the Docker image for this project, first, [install Docker](https://docs.docker.com/installation/) on your computer. At the Docker prompt, enter:

    docker run -dp 8787:8787 benmarwick/olympicdamboundaries

This will start a server instance of RStudio. Then open your web browser at localhost:8787 or or run `docker-machine ip default` in the shell to find the correct IP address, and log in with rstudio/rstudio.

Once logged in, use the Files pane (bottom right) to navigate to `/` (the root directory), then open the folder for this project, and open the `.Rproj` file for this project. Once that's open, you'll see the `analysis/paper` directory in the Files pane where you can find the R markdown document, and knit them to produce the results in the paper. More information about using RStudio in Docker is avaiable at the [Rocker](https://github.com/rocker-org) [wiki](https://github.com/rocker-org/rocker/wiki/Using-the-RStudio-image) pages.

We developed and tested the package on this Docker container, so this is the only platform that We're confident it works on, and so recommend to anyone wanting to use this package to generate the vignette, etc.

### Licenses

Manuscript:  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

Code: [MIT](http://opensource.org/licenses/MIT) year: `r format(Sys.Date(), "%Y")`, copyright holder: Ben Marwick

Data: [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Dependencies

I used [RStudio](http://www.rstudio.com/products/rstudio/) on Ubuntu 16.04 and Windows 7. See the colophon section of the docx file in `analysis/paper` for a full list of the packages that this project depends on.

### Contact

Ben Marwick, Associate Professor, Department of Anthropology
Denny Hall 117, Box 353100, University of Washington
Seattle, WA 98195-3100 USA

t. (+1) 206.552.9450   e. bmarwick@uw.edu
f. (+1) 206.543.3285   w. http://faculty.washington.edu/bmarwick/

### 2017-01-16 tests/testthat.R

@@ -1,0 +1,4 @@
+library(testthat)
+library(olympicdamboundaries)
+
+test_check("olympicdamboundaries")

### 2017-01-16 analysis/paper/paper.Rmd
bring over code from JFA

---
title: 'The title'
author:
- Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: journal-of-archaeological-science.csl
abstract: |
  This is the abstract.
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

In Table \@ref(tab:carstab) we can see some data about the relationship between pressure and tempurature. We can reference the table by the ID of the code chunk that produces the table, like this: `\@ref(tab:carstab)`

```{r carstab}
library(knitr)
kable(head(mtcars), caption = "Data about cars")
```

## Including Plots

You can also embed plots, for example:

```{r pressureplot, echo=FALSE, fig.cap="Plot of car data"}
plot(pressure)
```

In Figure \@ref(fig:pressureplot) we can see some data about pressure. We can cross-reference the figure like this: `\@ref(fig:pressureplot)`, using the chunk ID as the key to link the cross-reference to the figure.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Citations and References

We can have a citation, using a `.bib` file that holds all the details. To get this: [@Marwick2016repro] we type `[@Marwick2016repro]`. The text after the `@` is the bibtex key that links the in-text citation to the full details in the `.bib` file.

All of the usual variations on in-text citation formatting are possible in markdown, and listed for reference here: <http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html>

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
# which R packages and versions?
kable(data.frame(Setting = names(devtools::session_info()$platform),
                 Value = sapply(devtools::session_info()$platform, `[`, 1),
                 stringsAsFactors = FALSE,
                 row.names = NULL),
      caption = "R session information")
```

```{r}
kable(devtools::session_info()$packages,
      caption = "Packages that this report depends on")

# what commit is this file at?
library(git2r)
repo <- repository(path = "../..")
last_commit <- commits(repo)[[1]]
```

The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`. The current commit message is "`r last_commit@summary`".

##### 
<!-- This is a comment, to note that the five has symbols above will create a page break in the word document that is output from this document. This comment will not be visible in the word doc -->

## References

```{r eval = FALSE}

###########################################################
# checking P&M's edits from 14 May 2012
#
# clear the workspace
rm(list = ls())
#
# load data
load("E:/My Documents/My Papers/Journal of Field Archaeology/.RData")
# 
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
library(foreign)
ODX_allsites_2010 <- read.dbf("Join_Extract_ArcSite.dbf")
# 
# now looking at my analysis journal from 
# F:\My Documents\My Employment\HUGHES\Clause 7 Report\5thPreliminaryAnalysis
# 
# sites per TP
sites.per.tp <- data.frame(table(na.omit(ODX_allsites_2010$GREGLANDF)))
# find proportion of all sites in each TP
sites.per.tp$prop <- sites.per.tp$Freq/sum(sites.per.tp$Freq)*100
#
# get counts from density squares
ODX_allsites_2010$DENCOUNT <- with(ODX_allsites_2010, (as.numeric(levels(DENSQ1)[DENSQ1]))+(as.numeric(levels(DENSQ2)[DENSQ2]))+(as.numeric(levels(DENSQ3)[DENSQ3]))+(as.numeric(levels(DENSQ4)[DENSQ4]))+(as.numeric(levels(DENSQ5)[DENSQ5]))+(as.numeric(levels(DENSQ6)[DENSQ6]))+(as.numeric(levels(DENSQ7)[DENSQ7]))+(as.numeric(levels(DENSQ8)[DENSQ8])+(as.numeric(levels(DENSQ9)[DENSQ9]))/9)*((as.numeric(levels(DIMSIT1)[DIMSIT1]))*(as.numeric(levels(DIMSIT2)[DIMSIT2]))))
# check if all sites with density square data were processed
# should return TRUE
identical(length(na.omit(ODX_allsites_2010$DENSQ1)), length(na.omit(ODX_allsites_2010$DENCOUNT)))
# check how many sites with density square data
length(na.omit(ODX_allsites_2010$DENCOUNT))
#
# get counts when all artefacts counted
ODX_allsites_2010$COUNTONLY <- with( ODX_allsites_2010, as.numeric(levels(TOTART)[TOTART]) )
# check how many are ok
length(na.omit(ODX_allsites_2010$TOTART)) # should have
length(na.omit(ODX_allsites_2010$COUNTONLY)) # did get
#
# get counts from when only a VISTEST was taken:
ODX_allsites_2010$VISCOUNT<-with(ODX_allsites_2010, (as.numeric(levels(VISEST)[VISEST]))*((as.numeric(levels(DIMSIT1)[DIMSIT1]))*(as.numeric(levels(DIMSIT2)[DIMSIT2]))))
# check how many are ok
length(na.omit(ODX_allsites_2010$VISEST)) # should have
length(na.omit(ODX_allsites_2010$VISCOUNT)) # did get
#
ODX_allsites_2010$artefacts <- as.numeric(with(ODX_allsites_2010, 
      ifelse(COUNTONLY > 1, COUNTONLY,
      ifelse(VISCOUNT  > 1 ,  VISCOUNT, 
      ifelse(DENCOUNT  > 1 , DENCOUNT, "XX")))))
#                                          
# have a bit of a look 
# ODX_allsites_2010[ODX_allsites_2010$VISCOUNT  != 0 ,]
# ODX_allsites_2010[ODX_allsites_2010$DENCOUNT  != 0 ,]
# ODX_allsites_2010[ODX_allsites_2010$COUNTONLY != 0 ,]
#
# summary(ODX_allsites_2010$VISCOUNT)
# summary(ODX_allsites_2010$DENCOUNT)
# summary(ODX_allsites_2010$COUNTONLY)
#  
# check how many are ok
length(na.omit(ODX_allsites_2010$artefacts))
# 
# check how many sites in total
# summary( ODX_allsites_2010[,1:10] )
# 10636 sites in data, 133 are hearths so 10503 should 
# have artefact data... in fact 9859 do, so some 644 sites
# have problems with artefact data, that's only 0.6% so that's
# not too bad.
#
# get total number of artefacts in each TP
artefacts.per.tp <- aggregate(data = ODX_allsites_2010, artefacts ~ GREGLANDF, FUN = sum)
#
# check output
identical(
  sum(na.omit(ODX_allsites_2010$artefacts)),
  sum(artefacts.per.tp$artefacts))
#
# how many artefacts missing from summary?
sum(na.omit(ODX_allsites_2010$artefacts)) - sum(artefacts.per.tp$artefacts)
# 5926 artefacts missing... why are they missing? not sure
#
#
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
library(foreign)
ODX_TPs_2010 <- read.dbf("CLipped_ODO_terrain_patterns_2010_Ja.dbf")
# 
# Calculate the areas of the sampling polygons in ArcMap:
# 1.  Open the attribute table of the layer you want to edit. 
# 2.	Right-click the field heading for area (if there is no field for area values, you can add a new field for area by clicking the Options button and selecting the new field option). This must be formatted as a Double not a short integer.
# 3.	Click Calculate Values. 
# 4.	Checkmark the Advanced box (this is important - don't forget this!) 
# 5.	Type the following VBA statement in the first text box (you should be able to copy and paste): 
# Dim dblArea as double 
# Dim pArea as IArea 
# Set pArea = [shape] 
# dblArea = pArea.area 
# 6.	Type the variable  dblArea =in the second text box (near the bottom of the dialog box) directly under the area field name. 
#
# check areas and proportion of total areas for each TP
library(data.table)
ODX_TPs_2010.dt <- data.table(ODX_TPs_2010)
setkey(ODX_TPs_2010.dt,GREGLANDF)
ODX_TP_areas <- ODX_TPs_2010.dt[,sum(Clip_Area_), by = GREGLANDF]
ODX_TP_areas$prop <- (ODX_TP_areas$V1/sum(ODX_TP_areas$V1))*100
# check that proportions add to 100%
sum(ODX_TP_areas$prop)
# check of total area of TPs
sum(ODX_TP_areas$V1)/(1000*1000)
#
# repeat checks for landforms
ODX_LF_areas <- ODX_TPs_2010.dt[,sum(Clip_Area_), by = LANDFORM]
ODX_LF_areas$prop <- (ODX_LF_areas$V1/sum(ODX_LF_areas$V1))*100
# check that proportions add to 100%
sum(ODX_TP_areas$prop)
#
# Plot of TP areas (figure 4 of JFA paper)
install.packages("ggplot2")
library(ggplot2)
ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
  geom_bar(fill = "white", colour = "black") +
  theme_bw(base_size = 17) +
  opts(axis.text.x = theme_text(size = 17)) + 
  opts(axis.text.y = theme_text(size = 17)) +
  opts(axis.title.x = theme_text(size = 17)) + xlab("Terrain pattern") +
  opts(axis.title.y = theme_text(size = 17, angle = 90)) + ylab("Percentage of study area")
#
# check of sites per sq km for each TP
sites.per.sqkm <- merge(sites.per.tp, ODX_TP_areas, by.x = "Var1", by.y = "GREGLANDF")
sites.per.sqkm$sites.per <- with(sites.per.sqkm, Freq/V1*1000*1000)
#
# check of artefacts per sq km for each TP
artefacts.per.sqkm <- merge(artefacts.per.tp, ODX_TP_areas)
artefacts.per.sqkm$artefacts.per <- with(artefacts.per.sqkm, artefacts/V1*1000*1000)

# For looking a distance values from features in ArcMap
# First, polygon features have to be converted to line features.data management tools, features, feature to line.
# 1.  Open spatial analyst extension
# 2.	Define extent (LTA_areas is a good one)
# 3.	Spatial analyst - distance- straight line
# 4.	Then go to ArcToolbox - Spatial analyst tools - extraction 
# 5.	'Extract values to points'
# 6. Then arctoolbox, spatial analysis tools, extraction, extract values to points. point layer is sites, values come from raster, then RASTERVALU is distance field. 
# 7 .Then join to TP polygon to get TP values in sites dbf.
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
library(foreign)
ODX_allsites_2010 <- read.dbf("Join_Extract_ArcSite.dbf")
ODX_allsites_2010 <- ODX_allsites_2010[ !is.na(ODX_allsites_2010$GREGLANDF), ]
ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
# order the levels of GREGLANDF for plotting
# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
levels(ODX_allsites_2010.s$GREGLANDF)
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])
# plot
install.packages("ggplot2")
library(ggplot2)
ggplot(ODX_allsites_2010.s, aes(GREGLANDF, RASTERVALU)) + 
  geom_boxplot() + 
  scale_y_log10(breaks = seq(0,1400,200)) + 
  theme_bw(base_size = 17) +
  opts(axis.text.x = theme_text(size = 17)) + 
  opts(axis.text.y = theme_text(size = 12)) +
  opts(axis.title.x = theme_text(size = 17)) + xlab("Terrain pattern") +
  opts(axis.title.y = theme_text(size = 17, angle = 90)) + ylab("Distance of site to boundary (m)")

# anova to see if sig diff between distances...
fit <- aov(data = ODX_allsites_2010.s, RASTERVALU ~ GREGLANDF)
plot(TukeyHSD(fit))
# plot anova
install.packages("granovaGG", dependencies = TRUE)
library("granovaGG")
granovagg.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)
install.packages("granova")
library("granova")
granova.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)

setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology\\outer_buffers")
# start with OUTER buffers
bd_files <- list.files()
# list in, dataframe out
library(plyr)
library(foreign)
bd.data <- ldply(bd_files, read.dbf)
bd.data$SiteDens <- with(bd.data, Count_/buf_area * (1000*1000))
# exclude buffers with less than n sites since they 
# skew the density calculations 
n = 15
bd.data <- bd.data[bd.data$Count_ > n, ]
unique(bd.data$BUFF_DIST)
bd.data$Buf <- as.factor(with(bd.data, ifelse(BUFF_DIST == 100, '000-100', ifelse(BUFF_DIST == 200, '100-200', ifelse(BUFF_DIST == 300, '200-300', ifelse(BUFF_DIST == 400, '300-400', ifelse(BUFF_DIST == 500, '400-500','NA')))))))

#  how many sites in each set?
sum(na.omit(subset(bd.data, GREGLANDF=='Czs4')$Count_))
sum(na.omit(subset(bd.data, GREGLANDF=='Czs5')$Count_))

# now check the boundary densities for INNER buffers
setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology\\inner_buffers")
inbd.files <- list.files()
# list in, dataframe out
library(plyr)
library(foreign)
inbd.data <- ldply(inbd.files, read.dbf)
inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
unique(inbd.data$BUFF_DIST)
inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))

#  how many sites in each set?
sum(as.numeric((subset(inbd.data[as.numeric(inbd.data$Join_Count),], GREGLANDF=='Czs4'))$Join_Count))
sum(as.numeric((subset(inbd.data[as.numeric(inbd.data$Join_Count),], GREGLANDF=='Czs5'))$Join_Count))

# combine inner with outer buffers
inner <- subset(inbd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')[,c('GREGLANDF','SiteDens','Buf')]
outer <- subset(bd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')  [,c('GREGLANDF','SiteDens','Buf')]
# colnames(outer)[2] <-  'mean_SiteDens'
inner_outer <- na.omit(rbind(inner, outer))
# put the boundary distances in order
levels(inner_outer$Buf)
inner_outer$Buf <- factor(inner_outer$Buf, 
  levels(inner_outer$Buf)[c(5,4,3,2,1,6,7,8,9,10)])
# remove rows with zero for SiteDens
inner_outer <- inner_outer[inner_outer[, 'SiteDens']!=0, ]  
# plot inner and outer together
library(ggplot2)
library(grid)
ggplot(inner_outer,  aes(Buf, SiteDens)) + 
  geom_point(size = 2) + facet_grid(GREGLANDF ~ .) + 
  scale_y_log10() + 
  geom_smooth(aes(group = 1), fill = "grey80", colour = "grey") +
  theme_bw() + 
  opts(axis.title.x = theme_text(vjust = -0.5, size = 15)) + 
  opts(axis.title.y=theme_text(size = 15, angle=90)) + 
  opts(plot.margin = unit(c(1,1,2,2), "lines")) + 
  xlab("Distance from boundary in meters (negative values are interior distances, positive values are exterior distances)") + 
  ylab("Sites per sq km")

# significance tesing for changes in site density over buffers
# classic ANOVA...
summary(fit <- aov(data = inner_outer[inner_outer$GREGLANDF == 'Czs5' ,], SiteDens ~ Buf))
# yes for Csz5, no for Csz4
# inspect to see if assumptions are ok
qqnorm(resid(fit), main="Normal Q-Q Plot")
qqline(resid(fit), col="red")
# a bit wacky...

# try permutation ANOVA...
# install.packages("lmPerm")
# library(lmPerm)
# summary(fitp <- aovp(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, perm="Exact"))
# yes for Csz5, no for Csz4

library(coin) # library needed for more permutation tests
# We'll use Monte Carlo simulations to get the p value
oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
# this is a test of whether 
# two or more samples from normal distributions have the 
# same means, variances not assumed equal
kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
# yes for Csz5, no for Csz4

# In each of the occurrences of those TPs create buffer zones A (1m to -100m), B (-100m to -200m), C (-200m to -300m).  Then work out the density of sites in each buffer unit and in the whole TP.
# 
# Than plot the ratios for each Czs4 and Czs5 Terrain Pattern of:
# site densities in A/site densities in whole of each TP
# site densities in B/site densities in whole of each TP
# site densities in C/site densities in whole of each TP

inner # site densities per buffer area
sites.per.sqkm # densities per TP

# Czs5
# site densities in 0-100 m inner buffer divided by total density of TP
Czs5.inner.100 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-100-000",])
Czs5.inner.100.ratio <- mean(Czs5.inner.100$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs5",])$sites.per
# site densities in 100-200 m inner buffer divided by total density of TP
Czs5.inner.200 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-200-100",])
Czs5.inner.200.ratio <- mean(Czs5.inner.200$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs5",])$sites.per
# site densities in 200-300 m inner buffer divided by total density of TP
Czs5.inner.300 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-300-200",])
Czs5.inner.300.ratio <- mean(Czs5.inner.300$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs5",])$sites.per

# Czs4
# site densities in 0-100 m inner buffer divided by total density of TP
Czs4.inner.100 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-100-000",])
Czs4.inner.100.ratio <- mean(Czs4.inner.100$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs4",])$sites.per
# site densities in 100-200 m inner buffer divided by total density of TP
Czs4.inner.200 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-200-100",])
Czs4.inner.200.ratio <- mean(Czs4.inner.200$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs4",])$sites.per
# site densities in 200-300 m inner buffer divided by total density of TP
Czs4.inner.300 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-300-200",])
Czs4.inner.300.ratio <- mean(Czs4.inner.300$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs4",])$sites.per

# combine both for plotting
Czs4 <- c(Czs4.inner.100.ratio, Czs4.inner.200.ratio, Czs4.inner.300.ratio)
Czs5 <- c(Czs5.inner.100.ratio, Czs5.inner.200.ratio, Czs5.inner.300.ratio)
bound <- data.frame(cbind(Czs4, Czs5))
bound$dist <- c("0-100", "100-200", "200-300")
library(reshape2)
bound.m <- melt(bound)

# plot
library(ggplot2)
ggplot(bound.m, aes(dist,value, colour = variable)) + geom_line(aes(group = variable))

# convert to z-scores....
Czs4.sc <-  scale(Czs4)
Czs5.sc <-  scale(Czs5)
bound.sc <- data.frame(cbind(Czs4.sc, Czs5.sc))
bound.sc$dist <- c("0-100", "100-200", "200-300")
colnames(bound.sc) <- c("Czs4", "Czs5", "distance")
library(reshape2)
bound.sc.m <- melt(bound.sc)
colnames(bound.sc.m) <- c("distance", "Terrain Pattern", "density of sites")

# plot scales
library(ggplot2)
ggplot(bound.sc.m, aes(x=bound.sc.m[,1], y=bound.sc.m[,3])) + 
  geom_line(aes(group = bound.sc.m[,2], linetype = bound.sc.m[,2]), size = 1)  +
  theme_bw() +
  opts(axis.text.x = theme_text(size = 17)) + 
  opts(axis.text.y = theme_text(size = 17)) +
  opts(axis.title.x = theme_text(size = 17)) +
  opts(axis.title.y = theme_text(size = 17, angle = 90)) + 
  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
  ylab("z-score of site density in buffer area") +
  scale_linetype_manual(name="Terrain Pattern", values = c(1,2) ) +
  opts(legend.text=theme_text(size=14)) +
  opts(legend.title=theme_text(size=14))

# checking benefit of mobile GIS
dat <- read.table(text = "
Year Pre.2007	2007	2008	2009
Area.surveyed.(km2)	35	95	275	145
Rate.of.survey.(km2/person/day)	0	0.3	0.3	0.3
Total.no.persondays	0 317	917	483
Approximate.number.of.sites.recorded	665	3900	6600	5500
No.sites.recorded/person/day	 3.5	12.3	7.2	11.3
Frequency.of.occurrence.of.sites/km2	19	41	24	38", row.names = 1, header = TRUE)

dat_t <- data.frame(t(dat))
# quick look
pairs(dat_t)

# correlation of sites recorded by density of sites
cor.test(dat_t[,5], dat_t[,6], method = 'pearson')
# linear model for sites recorded by density of sites
summary(lm(dat_t[,5] ~ dat_t[,6]))

# subset 2007-2009 data
dat_t_s <- dat_t[2:4,]
# linear model for sites recorded by density of sites, 2007-2009
summary(lm(dat_t_s[,5] ~ dat_t_s[,6]))

# predict given 19 sites/km2
19 * 0.297  + 0.043

```

### 2017-01-17 analysis/paper/paper.Rmd
all text & code now in Rmd

@@
 ---
-title: 'The title'
+title: 'Landform boundary effects on Holocene forager landscape use in arid South Australia'
 author:
-- Ben Marwick
+- Ben Marwick (corresponding author)
+- Peter Hiscock 
+- Philip Hughes 
+- Marjorie Sullivan
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
     reference_docx: templates/template.docx
-bibliography: references.bib
-csl: journal-of-archaeological-science.csl
 abstract: |
-  This is the abstract.
+  Landscapes throughout a region vary in the resources they contain.
+We investigate how Holocene forager populations adapted to this variation in the linear sand dune desert of arid South Australia.
+We use data from surface scatters of stone artefacts collected during pedestrian survey to compare behaviors at landscape boundaries to behaviours at the centers of landforms.
+We propose a model of human use of the landscape that predicts that the prehistoric occupants of the study are were sensitive to the different economic potential of dissimilar landscapes.
+In evaluating the model we find that there are different densities of archaeological sites in each landscape type.
+We also find indications of a boundary effect resulting from people having used marginal areas of each landscape type in response to the resource characteristics of adjoining landforms.
+In addition, we make some observations on our field data collection methods, identifying the general conditions where mobile GIS may be optimally efficient for archaeological survey
 ---
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+library(knitr)
+opts_chunk$set(echo = FALSE)
 ```
 
-# Introduction
+Affiliations:
 
-## R Markdown
+Ben Marwick (corresponding author)
+University of Washington
+Denny Hall 117, Box 353100
+Seattle, WA 98195-3100 USA
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+t. (+1) 206.552.9450 
+f. (+1) 206.543.3285
+bmarwick@uw.edu
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+Peter Hiscock   
+The University of Sydney, Sydney NSW, Australia
 
-In Table \@ref(tab:carstab) we can see some data about the relationship between pressure and tempurature.
-We can reference the table by the ID of the code chunk that produces the table, like this: `\@ref(tab:carstab)`
+Philip Hughes    
+Department of Archaeology, Flinders University, GPO Box 2100, Adelaide, SA, 5001
 
-```{r carstab}
-library(knitr)
-kable(head(mtcars), caption = "Data about cars")
-```
+Marjorie Sullivan    
+The Australian National University, Canberra ACT 0200, Australia and 
+Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
 
-## Including Plots
+##### . 
+<!--- page break --->
 
-You can also embed plots, for example:
+## Introduction
 
-```{r pressureplot, echo=FALSE, fig.cap="Plot of car data"}
-plot(pressure)
-```
+A key step in understanding past human use of landscapes is investigating how people adapted their behaviours to the different landscapes they inhabited.
+Often this kind of analyses depends on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g.
+Barton 2003).
+Variations in human behavior at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences.
+Previous work on past human behaviors at ecological interfaces has focused on regions of highly contrasting biomes.
+For example, Epp (1985) found that archaeological sites were more concentrated in an ecotone (i.e.
+ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes.
+Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972 ).
+This increased variety and density of plants and animals, often described as an ‘edge effect’, was claimed by Harris et al.
+(1969) to be a causal factor in the emergence of agriculture.
+This edge effect was also extended to human culture, with Gummerman and Johnson (1971) arguing for increased human cultural diversity in ecotonal zones of central Arizona.
 
-In Figure \@ref(fig:pressureplot) we can see some data about pressure.
-We can cross-reference the figure like this: `\@ref(fig:pressureplot)`, using the chunk ID as the key to link the cross-reference to the figure.
+These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists have been overly simplistic in their use of many ecological concepts (Rhoades 1978, Hardesty 1980, King and Graham 1981).
+Problems with these previous approaches include ignoring controversies among biologists about the definition of ecotones, and unjustified assumptions about the uniformity of all ecotones, and about edge effects always increasing the amount of food species available to humans.
+Here we pursue a different approach by focusing on subtle variations in landscape characteristics within a biome, rather than between them.
+By focusing on continuous variation, rather than step-wise shifts in the archaeological record across ecological interfaces, we introduce a new approach to studying behaviors of prehistoric forager groups at interface zones.
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+In this study we demonstrate the use of gradients of change in archaeological materials over distance to explore forager choices of where to carry out activities in respect to environmental boundaries.
+We test a prediction that prehistoric foragers were sensitive to differences in resource structure as gradients across the landscape.
+This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability that make comparative landscape analysis more tractable in many contexts.
+The spatial scales of much of the previous work on prehistoric arid zone foragers has ranged from continental (Williams, Ulm et al.
+2013) to ecological regions (Holdaway, Douglass et al.
+2013).
+In this paper, we focus on human behavior in the arid zone at a novel scale and location: the interface between two types of subtly different landscapes.
+We show that prehistoric forager groups’ decisions about how they occupied the study area were sensitive to subtle differences in local landscape characteristics.
 
-## Citations and References
+Our work is motivated by previous archaeological work on cultural landscapes of foragers that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982, Veth 1993, Napton and Greathouse 1996, Holdaway 1998, Barton 2003, Smith 2006, Fanning, Holdaway et al.
+2008, Smith, Williams et al.
+2008, Veth, Smith et al.
+2008, Williams, Ulm et al.
+2013, Davies, Holdaway et al.
+2015).
+The archaeological signature of landuse typically has been the distribution and composition of surface stone artefact scatters, which are by far the most abundant type of site.
+In this context, sites are typically small, dense and discrete concentrations of stone artefacts that represent temporary, short-term activity areas.
+The spatially discrete quality of the majority of sites in these desert regions, such as our study area, means it is possible to distinguish between even quite close sites (<50 m apart) with confidence, unlike many other contexts where artifact distributions are more spatially continuous (Thomas 1975, Foley 1981, Dunnell and Dancey 1983, Kuna 2000).
 
-We can have a citation, using a `.bib` file that holds all the details.
-To get this: [@Marwick2016repro] we type `[@Marwick2016repro]`.
-The text after the `@` is the bibtex key that links the in-text citation to the full details in the `.bib` file.
+Our study area at Olympic Dam, South Australia, offers a unique opportunity to study landscape interfaces for two key reasons.
+First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (TABLE 1).
+The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes, making them convenient for spatial analysis of archaeological features.
+Landform boundary effects have previously been employed in studying the distribution of artefacts to infer culture group areas (Kimes, Haselgrove et al.
+1982).
+However, archaeological discussions of settlement patterns have typically focused on straight line distances from key resources such as raw materials for stone artefacts, as these are often easier to measure and interpret than landscape boundaries (Daniel 2001, Jones, Beck et al.
+2003).
+With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.
 
-All of the usual variations on in-text citation formatting are possible in markdown, and listed for reference here: <http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html>
+The second reason why our data is uniquely suited to investigating landscape boundary effects is that we have a large number of sites spread across several landscape types.
+The relatively high density of sites provides an opportunity to investigate fine-grained gradients of archaeological change over multiple landscapes and their interfaces.
+More specifically, we can test whether a boundary effect exists at the edge of landscape units by examining how the size and density of surface scatters of stone artefacts changes with their distance from the boundaries.
+
+## Olympic Dam and investigation of its archaeology
+
+Olympic Dam is located on the southeastern edge of the driest part of the Australian arid zone.
+Archaeological material is found across the broader region and partly reflects the long period of human occupation, the wide range of locally available resources and the extremely low rates of sediment movement that might otherwise obscure surface artefacts.
+Optically stimulated luminescence dating of sediments underling buried artefacts at one typical site we investigated indicates a terminus post quem  of 5 k BP for the artifact bearing deposit (Sullivan, Field et al.
+2012).
+Such an age is consistent with the majority of dates for human occupation of sand deserts elsewhere in Australia, where evidence has suggested a substantial expansion of human activity in the last few millennia (Veth, Smith et al.
+2001, Smith 2006).
+A single dated site is insufficient to draw conclusions beyond the local level, however the distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact scatters recorded during the survey (Hughes and Hiscock 2005).
+This mid to late Holocene period was also a time of broad-scale dune stability, so site visibility is not strongly controlled by dune movement.
+In 2013 we returned to sites that had been recorded 25 years earlier and found them to be in the same state as originally recorded, indicating that site visibility has changed little over 25 years.
+
+The Olympic Dam area has been the target of mineral exploration and mining since the 1980s.
+The extensive records of archaeological sites at Olympic Dam have been created during research-orientated consultancy work related to this exploration and mining activity.
+In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the proposed Olympic Dam mining project.
+They used terrain patterns based on combining categories of different landforms and surface geology (FIG 2) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982).
+The landscape consists of ancient stony plateau (or tableland) surfaces partly overlain by fields of longitudinal east-west dunes.
+Many of the interdune corridors (or swales) and extensive flat to gently sloping plateau surfaces without dunes are covered by a lag of gravels and cobbles referred to as gibber.
+Water sources include ephemeral small lakes, claypans and canegrass-vegetated swamps which are subject to high evaporation rates, with no free water available for most of the year.
+In this region vegetation patterns were also determined by the presence or absence of sand, gibber surfaces and water bodies (Badman 1999).
+This means that there are long-term relationships between the landscape units employed in the analysis of Hughes and Hiscock (1982), and plant and animal resources exploited by forager communities in the past.
+
+Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al.
+2011; Hughes and Hiscock 1982; Hughes and Sullivan 1984).
+In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts.
+The model results in a series of predictions about the archaeological record expected for each of six landform types (TABLE 1) and five geological regimes (TABLE 2).
+These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.
+
+These descriptions were initially tested in the 1980s and 1990s by archaeological surveys of the mining area and infrastructure corridors.
+The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites  (Hiscock and Hughes 1983, Hiscock 1989).
+In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009 (FIG 2).
+The data collected during this expansion phase is the basis for this study.
+
+## Methods and Materials
+
+Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 16,000 archaeological sites.
+We achieved close to 90% coverage of 353 km2.
+The sites consist of stone artefact scatters (87%), artefact scatters with knapping floors (7%), knapping floors (5.5%) and quarries (0.5%) (Hughes, Hiscock et al.
+2011).
+Extended descriptions of site definitions and survey methods have previously been presented in Hughes et al.
+(2011).
+
+To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information.
+Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003, Bevan and Conolly 2004, Tripcevich 2004, 2004, Given and Hyla 2006, Wagtendonk and De Jeu 2007, de la Vega and Agulla 2010, Tripcevich and Wernke 2010, Fei 2011, Scianna and Villa 2011, Traviglia 2011, Fee, Pettegrew et al.
+2013, Newhard, Levine et al.
+2013, Cascalheira, Gonçalves et al.
+2014, Banning and Hitchings 2015).
+A detailed description of our specific hardware, software, and justifications for our choices is presented in our SOM.
+
+We investigated behaviors at landform boundaries by analyzing the contrast between archaeological sites close to the boundary of a landform and the archaeology deeper within the landform.
+Our test of boundary effects focused on the archaeological evidence from resource-rich landform types/terrain patterns, such as Czs4, bordered by landscapes with lower resource levels, such as landform types 5 or 6.
+Because Czs4 and Czs5 dominate the 353 km2 sample area and contain the majority of sites, the analyses focused on comparing these two terrain patterns.
+
+Two approaches were used to diagnose the presence and character of boundary effects.
+In the first approach, we determined the straight-line distance from every site to the nearest boundary of the terrain pattern in which it was contained.
+Descriptive statistics for these distances allows testing of the prediction that sites tend to be closer to the edge of some landforms than other landforms.
+In the second approach, we measured the density of sites at 100 m intervals, or buffer zones, inwards from the boundary of the terrain patterns Csz4 and Csz5, as the best represented terrain patterns.
+These data were used to test the prediction that site densities tends to be higher towards the center of Csz4, as a resource-rich landform, compared to Csz5, as a resource-poor landform.
+We created interior buffer polygons at 0-100 m, 101-200 m and 201-300 m from the boundary of the terrain patterns and the number of sites per square kilometer in each buffer was measured.
+The density of sites in each buffer zone was then divided by the average density for the entire terrain pattern to standardize the density value for the terrain pattern.
+Finally, z-scores were calculated to put the differences in site density on a comparable scale between the two terrain patterns.
+Z-scores are a transformation of the data that show how many standard deviations the site density in each buffer polygon varies from the mean density (z = 0) of the buffers in each terrain pattern.
+This second approach provided a way to minimize the effect of differences in the sizes of landscape units being compared.
+It does this by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns.
+
+# Results
+
+### Model validation
+
+The data from Olympic Dam allow a preliminary test of the predictions of the landuse model developed for the region in the early 1980s (Hughes, Hiscock et al.
+2011).
+The first prediction tested here is that there was differential use of localities within dunefields.
+More specifically, areas of widely-spaced longitudinal sand dunes were predicted to have been intensely occupied when compared with areas containing more closely-spaced dunes.
+This difference is anticipated because of the greater range and density of animal foods and water sources in places where dunes are more widely separated.
+
+A wide variety of landforms and geological regimes were recorded, displayed in FIG 4 as the percentage coverage of each terrain pattern.
+In total these areas contained about 10,400 archaeological sites.
+Most of that sample area consists of widely spaced dunes – landform type 4 (46%), moderately spaced dunes – landform type 5 (35%) and closely spaced dunes – landform type 6 (11%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P).
+Other terrain patterns make up less than 10% of the 353 km2 study area (FIG 4).
+
+About 44% of all sites were recorded in these two terrain patterns (Czs4: 33.8%, Czs5: 10.7%) with no other terrain pattern having more than 20% of all sites.
+When considered as homogenous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
+As predicted (TABLE 1), sites are more frequent in fields of widely spaced dunes (Czs4) with 53 sites/km2 compared with more closely spaced dunes (Czs5) which have only 21 sites/km2 (t = 5.29, df = 8.6, p-value = 0.0005).
+Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at 35,789 artefacts/km2, compared with closely spaced dunes Czs5 at 2,702 artefacts/km2.
+
+### Landscape boundary effects
+
+FIG 5 depicts median distances for site distances from terrain pattern boundaries where the dunefields are on average more than 1 km across (FIG 2).
+The general pattern is clear, with distances being substantially further in widely spaced dunefields such landform type 4 than in more closely spaced dunefields such as landform types 5 and 6.
+An examination of FIG 2 shows that this relationship is not dependent simply on the size or shape of the terrain patterns.
+
+FIG 6 shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform.
+These data give support to the prediction that resource-rich landscapes such as Csz4 have sites located away from their margin and far inside the terrain pattern areas while landscapes with lower resource levels such as Csz5 commonly have sites closer to their margins and fewer sites further in their interiors.
+
+## Discussion
+
+Confirmation of the model predictions were facilitated by two survey characteristics.
+One is that the extensive survey area, the intensive level of survey and the resulting large sample size of archaeological sites, meant that observation of human choices and behaviour could be obtained at a very high spatial resolution.
+The second is that objective and robust environmental classifications, creating well-defined terrain patterns, made it possible to quantify landscape diversity in the study area at a high spatial resolution and reduced false associations that could result from less accurate environmental mapping.
+Reprise testing of model predictions…
+Reprise testing of boundary effect predictions…
+
+This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to the different economic potential of dissimilar terrain patterns.
+This claim is axiomatic to human behavioural ecology in which different landscapes are commonly compared as homogenous units to improve analytical tractability.
+The novel element of the data presented here is that the distribution of human activities varies substantially and predictably within as well as between terrain patterns.
+It is also noteworthy that a relatively subtle geographic differences between landscapes – the different spacing of sand dunes between Czs4 and Czs5 – has a marked visible effect on the distribution of archaeological sites.
+
+### Efficiency of a mobile GIS compared to paper systems for field survey
+
+Our survey activity in the Olympic Dam region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
+Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM).
+TABLE 5 shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS.
+Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day.
+This metric is useful because it is easy to measure without burdening field workers with additional data logging tasks.
+However, we need to take into account possible confounding factors, such as travel time to the day’s survey area, the size and density of the sites in the day’s survey area, skill and efficiency of the individual archaeologist and the sampling strategy.
+
+Our pre-2007 data were collected by more experienced archaeologists than the 2007-2009 data, using a recording form with less variables than our mobile GIS recording forms, and the pre-2007 survey areas were all closer to the camp, with less travel time compared to the 2007-2009 data.
+Further, the average site size for the pre-2007 data is less than the 2007-2009 data.
+TABLE 5 shows that prior to adopting the mobile GIS in 2007 we recorded 3.5 sites per person per day.
+With the mobile GIS our rate increased substantially to an average of 10.2 sites per person per day for 2007-2009.
+
+While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS because the rate of recording is strongly correlated with the density of sites (r = 0.97, p = 0.003).
+To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data (rate = site density * 0.297 (0.004)+ 0.062139 (0.146),  r2 = 0.999, p = 0.0002, standard errors in parentheses) and predicted the site recording rate given the density of 19 sites km-2 observed in the pre-2007 data.
+The resulting rate is 5.7 sites/person/day, which is what we would expect if a mobile GIS had been used to record sites in an area with 19 sites km-2.
+The difference of 2.2 sites/person/day between the actual pre-2007 average rate of 3.5 and the extrapolated average rate of 5.7 is most likely due to the use of mobile GIS.
+This represents a sixty percent increase in the site recording rate.
+This demonstrates an advantage for using mobile GIS in a context where data collection was intensive, a highly structured data collection form could be used, and the start-up and training costs could be amortized over multiple years.
+
+## Conclusion
+
+The substantive archaeological contribution of this study, resulting from the use of mobile GIS to record archaeological data, has been to show that when choosing habitations prehistoric people in the Olympic Dam area were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
+The implication of this finding is that even in relatively homogenous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites (cf.
+Attwell and Fletcher 1987).
+These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007, Kowalewski 2008: 233).
+Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with simple behavioural ecological models.
+
+## Acknowledgments
+
+This work was undertaken under a contract with BHP Billiton that includes research support.
+Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301) and a grant from the Herbert H.
+Gowen International Studies Endowment (UW Office of Global Affairs).
+
+##### 
+<!--- page break --->
+
+## References
+
+Attwell, M. R. and M. Fletcher (1987). "An analytical technique for investigating spatial relationships." Journal of Archaeological Science 14(1): 1-11.
+
+Badman, F. J. (1999). "The Lake Eyre South Study: Vegetation." Lake Eyre South Monograph Series 2: 176-200.
+
+Banning, E. B. and P. Hitchings (2015). "Digital Archaeological Survey: Using iPads In Archaeological Survey In Wadi Quseiba, Northern Jordan." The SAA Archaeological Record 15(4): 31-36.
+
+Barton, H. (2003). "The Thin Film of Human Action: interpretations of Arid Zone Archaeology." Australian Archaeology 57: 32-41.
+
+Bevan, A. and J. Conolly (2004). "GIS, Archaeological Survey, and Landscape Archaeology on the Island of Kythera, Greece." Journal of Field Archaeology 29(1-2): 123-138.
+
+Cascalheira, J., C. Gonçalves and N. Bicho (2014). "Smartphones and the use of customized Apps in archaeological projects." The SAA Archaeological Record 14(5): 20-25.
+
+Daniel, I. R., Jr. (2001). "Stone Raw Material Availability and Early Archaic Settlement in the Southeastern United States." American Antiquity 66(2): 237-265.
+
+Davies, B., S.
+J.
+Holdaway and P.
+C.
+Fanning (2015).
+"Modelling the palimpsest: An exploratory agent-based model of surface archaeological deposit formation in a fluvial arid Australian landscape." The Holocene: 0959683615609754.
+
+de la Vega, S.
+G.-D.
+and S.
+O.
+Agulla (2010).
+Using mobile GIS.
+for Field Digital Data Acquisition (FDA) in archaeological field survey.
+Geoinformation Technologies for Geo-Cultural Landscapes: European Perspectives.
+A.
+Vassilopoulos, N.
+Evelpidou, O.
+Bender and A.
+Krek: 35-46.
+Dunnell, R. C. and W. S. Dancey (1983). "The Siteless Survey: A Regional Scale Data Collection Strategy." Advances in Archaeological Method and Theory 6: 267-287.
+
+Epp, H. T. (1985). "Ecological Edges and Archaeological Site Location in Saskatchewan, Canada." North American Archaeologist 5(4): 323-336.
+
+Fanning, P.
+C., S.
+J.
+Holdaway and E.
+J.
+Rhodes (2008).
+"A new geoarchaeology of Aboriginal artefact deposits in western NSW, Australia: establishing spatial and temporal geomorphic controls on the surface archaeological record." Geomorphology 101(3): 524-532.
+
+Fee, S. B., D. K. Pettegrew and W. R. Caraher (2013). "Taking Mobile Computing to the Field." Near Eastern Archaeology (NEA) 76(1): 50-55.
+
+Fei, G. (2011). "Data Acquisition of Field Archaeological Supported by PDA+ GPS and ArcPad." Journal of Hefei University (Natural Sciences) 2: 011.
+
+Foley, R.
+(1981).
+Off-site archaeology: an alternative approach for the short-sited.
+Pattern of the past: studies in honour of David Clarke.
+I.
+Hodder, G.
+Isaac and N.
+Hammond.
+Cambridge, Cambridge University Press: 157-183.
+
+Given, O. and S. Hyla (2006). "Forging Partnerships in Laos." Expedition 48(1): 38-39.
+Glassow, M. A. and M. Leone (1972 ). Changes in the Adaptations of Southwestern Basketmakers: a Systems Perspective. Contemporary Archaeology. M. Leone, Southern Illinois University Press: 289-302.
+
+Gumerman, G.
+J.
+and R.
+R.
+Johnson (1971).
+"Prehistoric human population distribution in a biological transition zone." The distribution of prehistoric population aggregates," edited by George J.
+Gumerman: 83-102.
+
+Hardesty, D. L. (1980). "The Use of General Ecological Principles in Archaeology." Advances in Archaeological Method and Theory 3: 157-187.
+
+Harris, D. R., P. J. Ucko and G. Dimbleby (1969). "Agricultural systems, ecosystems and the origins of agriculture." The domestication and exploitation of plants and animals.: 3-15.
+
+Hiscock, P.
+(1989).
+Artefact recording in the field.
+Sites and Bytes.
+Recording Aboriginal places in Australia.
+J.
+Flood, I.
+Johnson and S.
+Sullivan.
+Canberra, Australian Government Publishing Service: .20-38.
+
+Hiscock, P. and P. J. Hughes (1983). "One method of recording scatters of stone artefacts during site surveys." Australian Archaeology 17: 87-98.
+
+Holdaway, S. (1998). New approaches to open site spatial archaeology in Sturt National Park, New South Wales, Australia. Archaeology in Oceania.
+
+Holdaway, S.
+J., M.
+J.
+Douglass and P.
+C.
+Fanning (2013).
+"A new ecological framework for understanding human environment interactions in arid Australia." Archaeology in environment and technology intersections and transformations: 51-68.
+
+Hughes, P.
+and P.
+Hiscock (2005).
+The archaeology of the Lake Eyre South area.
+The archaeology of the Lake Eyre South Region.
+W.
+J.
+H.
+Slaytor.
+Adelaide, Royal Geographical Society of South Australia.
+6: 1-20.
+
+Hughes, P., P.
+Hiscock, M.
+Sullivan and B.
+Marwick (2011).
+"An outline of archaeological investigations for the Olympic Dam Expansion in arid northeast South Australia." Journal of the Anthropological Society of South Australia 34: 21-37
+
+Hughes, P. J. and P. Hiscock (1982). Annex 5. Olympic Dam Draft Environmental Impact Statement. Unpublished Report to Roxby Management Services Pty Ltd through Kinhill Stearns Roger Pty Ltd, Adelaide.
+
+Jones, G. T., C. Beck, E. E. Jones and R. E. Hughes (2003). "Lithic source use and paleoarchaic foraging territories in the Great Basin." American Antiquity 68(1): 5-38.
+
+Kimes, T., C. Haselgrove and I. Hodder (1982). "A method for the identification of the location of regional cultural boundaries." Journal of Anthropological Archaeology 1(2): 113-131.
+
+King, F. B. and R. W. Graham (1981). "Effects of Ecological and Paleoecological Patterns on Subsistence and Paleoenvironmental Reconstructions." American Antiquity 46(1): 128-142.
+
+Kowalewski, S. A. (2008). "Regional settlement pattern studies." Journal of Archaeological Research 16(3): 225-285.
+
+Kuna, M.
+(2000).
+Surface artefact studies in the Czech Republic.
+The Future of Surface Artefact Survey in Europe, Sheffield Academic Press, Sheffield, UK.
+J.
+Bintliff, M.
+Kuna and N.
+Venclova.
+Sheffield, Sheffield Archaeological Monographs: 29-44.
+
+McPherron, S. and H. Dibble (2003). "Using computers In adverse field conditions: Tales from the Egyptian desert." The SAA Archaeological Record 3(5): 28-32.
+
+Napton, L.
+K.
+and E.
+A.
+Greathouse (1996).
+Archaeological investigations in the Simpson Desert, Northern Territory.
+Archaeology of Northern Australia.
+P.
+Veth and P.
+Hiscock.
+Brisbane, University of Queensland.
+TEMPUS Volume 4: 106-136.
+
+Newhard, J.
+M.
+L., N.
+S.
+Levine, A.
+D.
+Phebus, S.
+Craft and J.
+D.
+Littlefield (2013).
+"A geoinformatic approach to the collection of archaeological survey data." Cartography and Geographic Information Science 40(1): 3-17.
+
+Rhoades, R. E. (1978). "Archaeological Use and Abuse of Ecological Concepts and Studies: The Ecotone Example." American Antiquity 43(4): 608-614.
+
+Scianna, A. and B. Villa (2011). "GIS applications in archaeology." Archeologia e Calcolatori 22: 337-363.
+
+Smith, M., A.
+Williams, C.
+Turney and M.
+Cupper (2008).
+"Human—environment interactions in Australian drylands: exploratory time-series analysis of archaeological records." The Holocene 18(3): 389-401.
+
+Smith, M.
+A.
+(2006).
+"Characterizing late Pleistocene and Holocene stone artefact assemblages from Puritjarra rock shelter: a long sequence from the Australian desert." Records of the Australian Museum 58: 371-410.
+
+Struever, S. (1968). Woodland subsistence-settlement systems in the lower Illinois Valley. New perspectives in archaeology. S. R. Binford and L. R. Binford. Chicago, Aldine: 285-312.
+
+Sullivan, M., T.
+L.
+Field, P.
+Hughes, B.
+Marwick, P.
+Przystupa and J.
+K.
+Feathers (2012).
+"OSL ages that inform late phases of dune formation and human occupation near Olympic Dam in Northeastern South Australia." Quaternary Australasia 29(1): 4-11.
+
+Thomas, D. H. (1975). Nonsite sampling in archaeology: up the creek without a site. Sampling in archaeology. J. W. Mueller. Tucson, AZ, University of Arizona Press: 61-81.
+Traviglia, A. (2011). Integrated Archaeological Investigations for the Study of the Greater Aquileia Area. The New Technologies for Aquileia, Aquileia, Italy.
+
+Tripcevich, N. (2004). "Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey." Cartography and Geographic Information Science 31(3): 137-151.
+
+Tripcevich, N. (2004). "Mobile GIS in archaeological survey." The SAA Archaeological Record(May).
+
+Tripcevich, N. and S. A. Wernke (2010). "On-site recording of excavation data using mobile GIS." Journal of Field Archaeology 35(4): 380-397.
+
+Veth, P. (1993). "Islands in the Interior: A Model for the Colonisation of Australia's Arid Zone." Archaeology in Oceania 24: 81-92.
+
+Veth, P., M. Smith and M. Haley (2001). "Kaalpi: The archaeology of an outlying range in the dunefields of the Western Desert." Australian Archaeology(52): 9-17.
+
+Veth, P., M. Smith and P. Hiscock (2008). Desert peoples: archaeological perspectives, John Wiley & Sons.
+
+Wagtendonk, A. J. and R. A. M. De Jeu (2007). "Sensible field computing: Evaluating the use of mobile GIS methods in scientific fieldwork." Photogrammetric Engineering and Remote Sensing 73(6): 651.
+
+Williams, A.
+N., S.
+Ulm, A.
+R.
+Cook, M.
+C.
+Langley and M.
+Collard (2013).
+"Human refugia in Australia during the Last Glacial Maximum and terminal Pleistocene: A geospatial analysis of the 25–12 ka Australian archaeological record." Journal of Archaeological Science 40(12): 4612-4625.
+
+Wilson, L. (2007). "Understanding prehistoric lithic raw material selection: application of a gravity model." Journal of Archaeological Method and Theory 14(4): 388-411.
+
+##### .
 
@@
 The current commit message is "`r last_commit@summary`".
-
-##### 
-<!-- This is a comment, to note that the five has symbols above will create a page break in the word document that is output from this document. This comment will not be visible in the word doc -->
-
-## References
 

### 2017-01-23 analysis/paper/paper.Rmd
initial

@@
 # load data
-load("E:/My Documents/My Papers/Journal of Field Archaeology/.RData")
+load("../data/.RData")
 #
-setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
+
 library(foreign)
-ODX_allsites_2010 <- read.dbf("Join_Extract_ArcSite.dbf")
+ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
 # 
@@
 #
-setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
 library(foreign)
-ODX_TPs_2010 <- read.dbf("CLipped_ODO_terrain_patterns_2010_Ja.dbf")
+ODX_TPs_2010 <- read.dbf("../data/CLipped_ODO_terrain_patterns_2010_Ja.dbf")
 # 
@@
 # Plot of TP areas (figure 4 of JFA paper)
-install.packages("ggplot2")
+
 library(ggplot2)
 ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
-  geom_bar(fill = "white", colour = "black") +
+  geom_bar(stat = "identity",
+           fill = "white", colour = "black") +
   theme_bw(base_size = 17) +
-  opts(axis.text.x = theme_text(size = 17)) + 
-  opts(axis.text.y = theme_text(size = 17)) +
-  opts(axis.title.x = theme_text(size = 17)) + xlab("Terrain pattern") +
-  opts(axis.title.y = theme_text(size = 17, angle = 90)) + ylab("Percentage of study area")
+  theme(axis.text.x = element_text(size = 17),
+        axis.text.y = element_text(size = 17),
+        axis.title.x = element_text(size = 17),
+        axis.title.y = element_text(size = 17, angle = 90)) + 
+  ylab("Percentage of study area") +
+  xlab("Terrain pattern") 
 #
@@
 # 7 .Then join to TP polygon to get TP values in sites dbf.
-setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology")
+
 library(foreign)
-ODX_allsites_2010 <- read.dbf("Join_Extract_ArcSite.dbf")
+ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
 ODX_allsites_2010 <- ODX_allsites_2010[ !is.na(ODX_allsites_2010$GREGLANDF), ]
@@
   levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])
+
 # plot
-install.packages("ggplot2")
 library(ggplot2)
@@
   theme_bw(base_size = 17) +
-  opts(axis.text.x = theme_text(size = 17)) + 
-  opts(axis.text.y = theme_text(size = 12)) +
-  opts(axis.title.x = theme_text(size = 17)) + xlab("Terrain pattern") +
-  opts(axis.title.y = theme_text(size = 17, angle = 90)) + ylab("Distance of site to boundary (m)")
+  theme(axis.text.x =  element_text(size = 17),
+        axis.text.y =  element_text(size = 12),
+        axis.title.x = element_text(size = 17),
+        axis.title.y = element_text(size = 17, 
+                                    angle = 90))  +
+  xlab("Terrain pattern") +
+  ylab("Distance of site to boundary (m)")
 
@@
 plot(TukeyHSD(fit))
+
 # plot anova
-install.packages("granovaGG", dependencies = TRUE)
+# install.packages("granovaGG", dependencies = TRUE)
 library("granovaGG")
 granovagg.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)
-install.packages("granova")
+# install.packages("granova")
 library("granova")
@@
 
-setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology\\outer_buffers")
 # start with OUTER buffers
-bd_files <- list.files()
+bd_files <- list.files("../data/outer_buffers", full.names = TRUE)
 # list in, dataframe out
@@
 # now check the boundary densities for INNER buffers
-setwd("E:\\My Documents\\My Papers\\Journal of Field Archaeology\\inner_buffers")
-inbd.files <- list.files()
+inbd.files <- list.files("../data/inner_buffers", full.names = TRUE)
 # list in, dataframe out
@@
 library(grid)
-ggplot(inner_outer,  aes(Buf, SiteDens)) + 
-  geom_point(size = 2) + facet_grid(GREGLANDF ~ .) + 
+library(scales)
+ggplot(inner_outer,  
+       aes(Buf, 
+           SiteDens)) + 
+  geom_point(size = 2) + 
+  facet_grid(GREGLANDF ~ .) + 
   scale_y_log10() + 
-  geom_smooth(aes(group = 1), fill = "grey80", colour = "grey") +
+  geom_smooth(aes(group = 1), 
+              fill = "grey80", 
+              colour = "grey") +
   theme_bw() + 
-  opts(axis.title.x = theme_text(vjust = -0.5, size = 15)) + 
-  opts(axis.title.y=theme_text(size = 15, angle=90)) + 
-  opts(plot.margin = unit(c(1,1,2,2), "lines")) + 
-  xlab("Distance from boundary in meters (negative values are interior distances, positive values are exterior distances)") + 
+  theme(
+  axis.title.x = element_text(vjust = -0.5, 
+                              size = 15),
+  axis.title.y=  element_text(size = 15, 
+                              angle=90), 
+  plot.margin = unit(c(1,1,2,2), "lines")) + 
+  xlab("Distance from boundary in meters \n(negative values are interior distances, \npositive values are exterior distances)") + 
   ylab("Sites per sq km")
@@
 oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
+oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, distribution=approximate(B=10000))
 # this is a test of whether 
@@
 kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
+kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, distribution=approximate(B=10000))
 # yes for Csz5, no for Csz4
@@
 library(ggplot2)
-ggplot(bound.m, aes(dist,value, colour = variable)) + geom_line(aes(group = variable))
+ggplot(bound.m, 
+       aes(dist,
+           value, 
+           colour = variable)) + 
+  geom_line(aes(group = variable))
 
@@
 library(ggplot2)
-ggplot(bound.sc.m, aes(x=bound.sc.m[,1], y=bound.sc.m[,3])) + 
-  geom_line(aes(group = bound.sc.m[,2], linetype = bound.sc.m[,2]), size = 1)  +
+ggplot(bound.sc.m, 
+       aes(x=bound.sc.m[,1], 
+           y=bound.sc.m[,3])) + 
+  geom_line(aes(group = bound.sc.m[,2], 
+                linetype = bound.sc.m[,2]), 
+            size = 1)  +
   theme_bw() +
-  opts(axis.text.x = theme_text(size = 17)) + 
-  opts(axis.text.y = theme_text(size = 17)) +
-  opts(axis.title.x = theme_text(size = 17)) +
-  opts(axis.title.y = theme_text(size = 17, angle = 90)) + 
+  theme(
+  axis.text.x =  element_text(size = 17), 
+  axis.text.y =  element_text(size = 17),
+  axis.title.x = element_text(size = 17),
+  axis.title.y = element_text(size = 17,
+                              angle = 90),
+  legend.text =  element_text(size=14),
+  legend.title = element_text(size=14)) + 
   xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
   ylab("z-score of site density in buffer area") +
-  scale_linetype_manual(name="Terrain Pattern", values = c(1,2) ) +
-  opts(legend.text=theme_text(size=14)) +
-  opts(legend.title=theme_text(size=14))
+  scale_linetype_manual(name="Terrain Pattern", values = c(1,2) )
 

### 2017-01-23 README.Rmd
get some plots and maps in there

@@
 [![Circle CI](https://circleci.com/gh/benmarwick/olympicdamboundaries.svg?style=shield&circle-token=:circle-token)](https://circleci.com/gh/benmarwick/olympicdamboundaries)
-[![codecov.io](https://codecov.io/github/benmarwick/olympicdamboundaries/coverage.svg?branch=master)](https://codecov.io/github/benmarwick/olympicdamboundaries?branch=master) [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0001--7879--4531-green.svg)](http://orcid.org/0000-0001-7879-4531)
+[![ORCiD](https://img.shields.io/badge/ORCiD-0000--0001--7879--4531-green.svg)](http://orcid.org/0000-0001-7879-4531)
 

### 2017-01-23 analysis/paper/paper.Rmd
get some plots and maps in there

@@
 opts_chunk$set(echo = FALSE)
+options(warn=-1)
+# image width should be 90 mm for 1 col, 140 mm for 1.5 col, 190mm for two cols
 ```
@@
 
-Olympic Dam is located on the southeastern edge of the driest part of the Australian arid zone.
+```{r overviewmap, fig.cap="Map of region including the study area", fig.width=9}
+
+knitr::include_graphics("../figures/regional_map.png")
+
+```
+
+Olympic Dam is located on the southeastern edge of the driest part of the Australian arid zone (Figure \@ref(fig:overviewmap)).
 Archaeological material is found across the broader region and partly reflects the long period of human occupation, the wide range of locally available resources and the extremely low rates of sediment movement that might otherwise obscure surface artefacts.
@@
 
+```{r mapofstudyarea1, fig.cap="Map of study area with landforms and terrain patterns", fig.width=9}
+
+knitr::include_graphics("../figures/landforms_only.png")
+
+```
+
+```{r mapofstudyarea2, fig.cap="Map of study area with landforms and terrain patterns", fig.width=9}
+
+knitr::include_graphics("../figures/landforms_and_sites.png")
+
+```
+
 The Olympic Dam area has been the target of mineral exploration and mining since the 1980s.
@@
 In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the proposed Olympic Dam mining project.
-They used terrain patterns based on combining categories of different landforms and surface geology (FIG 2) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982).
+They used terrain patterns based on combining categories of different landforms and surface geology (Figure \@ref(fig:mapofstudyarea)) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982).
 The landscape consists of ancient stony plateau (or tableland) surfaces partly overlain by fields of longitudinal east-west dunes.
@@
 The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites  (Hiscock and Hughes 1983, Hiscock 1989).
-In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009 (FIG 2).
+In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009 (Figure \@ref(fig:mapofstudyarea)).
 The data collected during this expansion phase is the basis for this study.
@@
 
+```{r sitetypetally}
+library(foreign)
+library(data.table)
+ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
+ODX_allsites_2010 <- data.table(ODX_allsites_2010)
+# percentages of each site type
+ODX_site_type_tally <- ODX_allsites_2010[  , by = .(SITTYP), .(.N ), ]
+ODX_site_type_tally$prop <- with(ODX_site_type_tally, round(N / sum(N),3) *100)
+
+artefact_scatters <- with(ODX_site_type_tally, 
+                          ODX_site_type_tally[SITTYP == "Artefact Scatter (AS)", 
+                                              prop])
+as_kf           <- with(ODX_site_type_tally, 
+                          ODX_site_type_tally[SITTYP == "AS + Knapping Floor (KF)", 
+                                              prop])
+knapping_floors  <- with(ODX_site_type_tally, 
+                          ODX_site_type_tally[SITTYP == "Single KF", 
+                                              prop])
+quarries         <- with(ODX_site_type_tally, 
+                          ODX_site_type_tally[SITTYP == "Quarry", 
+                                              prop])
+
+```
+
 Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 16,000 archaeological sites.
 We achieved close to 90% coverage of 353 km2.
-The sites consist of stone artefact scatters (87%), artefact scatters with knapping floors (7%), knapping floors (5.5%) and quarries (0.5%) (Hughes, Hiscock et al.
+The sites consist of stone artefact scatters (`r artefact_scatters`%), artefact scatters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al.
 2011).
@@
 
-A wide variety of landforms and geological regimes were recorded, displayed in FIG 4 as the percentage coverage of each terrain pattern.
-In total these areas contained about 10,400 archaeological sites.
-Most of that sample area consists of widely spaced dunes – landform type 4 (46%), moderately spaced dunes – landform type 5 (35%) and closely spaced dunes – landform type 6 (11%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P).
-Other terrain patterns make up less than 10% of the 353 km2 study area (FIG 4).
+```{r percentcoverageofeachterrainpattern, fig.cap="Terrain pattern percentage coverage of the 2007-2008 study area", fig.width=9}
+# Terrain pattern percentage coverage of the 2007-2008 study area in Figure map
+library(foreign)
+ODX_TPs_2010 <- read.dbf("../data/CLipped_ODO_terrain_patterns_2010_Ja.dbf")
+
+# check areas and proportion of total areas for each TP
+library(data.table)
+ODX_TPs_2010.dt <- data.table(ODX_TPs_2010)
+setkey(ODX_TPs_2010.dt,GREGLANDF)
+ODX_TP_areas <- ODX_TPs_2010.dt[,sum(Clip_Area_), by = GREGLANDF]
+ODX_TP_areas$prop <- (ODX_TP_areas$V1/sum(ODX_TP_areas$V1))*100
+# check that proportions add to 100%
+# sum(ODX_TP_areas$prop)
+# check of total area of TPs
+# sum(ODX_TP_areas$V1)/(1000*1000)
+#
+# repeat checks for landforms
+ODX_LF_areas <- ODX_TPs_2010.dt[,sum(Clip_Area_), by = LANDFORM]
+ODX_LF_areas$prop <- (ODX_LF_areas$V1/sum(ODX_LF_areas$V1))*100
+# check that proportions add to 100%
+# sum(ODX_TP_areas$prop)
+# 
+# get values to use in text
+LF4 <- round(with(ODX_LF_areas, ODX_LF_areas[LANDFORM == "4", prop]),1)
+LF5 <- round(with(ODX_LF_areas, ODX_LF_areas[LANDFORM == "5", prop]),1)
+LF6 <- round(with(ODX_LF_areas, ODX_LF_areas[LANDFORM == "6", prop]),1)
+#
+# Plot of TP areas (figure 4 of JFA paper)
+text_size <- 13
+library(ggplot2)
+ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
+  geom_bar(stat = "identity",
+           fill = "white", 
+           colour = "black") +
+  theme_bw(base_size = text_size) +
+  theme(axis.text.x = element_text(size = text_size),
+        axis.text.y = element_text(size = text_size),
+        axis.title.x = element_text(size = text_size),
+        axis.title.y = element_text(size = text_size, angle = 90)) + 
+  ylab("Percentage of study area") +
+  xlab("Terrain pattern") +
+  coord_fixed(ratio =  0.618/2)
+
+# count of all sites per TP
+all_sites_in_study_area <- prettyNum(sum(table(na.omit(ODX_allsites_2010$GREGLANDF))),
+                                     big.mark = ",")
+```
+
+A wide variety of landforms and geological regimes were recorded, displayed in figure \@ref(fig:percentcoverageofeachterrainpattern) as the percentage coverage of each terrain pattern.
+In total these areas contained about `r all_sites_in_study_area` archaeological sites.
+Most of the study area consists of widely spaced dunes – landform type 4 (`r LF4`%), moderately spaced dunes – landform type 5 (`r LF5`%) and closely spaced dunes – landform type 6 (`r LF6`%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P).
+Other terrain patterns make up less than 10% of the 353 km2 study area (Figure \@ref(fig:percentcoverageofeachterrainpattern)).
 
@@
 #
-# clear the workspace
-rm(list = ls())
-#
-# load data
-load("../data/.RData")
-#
-
 library(foreign)
 ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
-# 
-# now looking at my analysis journal from 
-# F:\My Documents\My Employment\HUGHES\Clause 7 Report\5thPreliminaryAnalysis
-# 
 # sites per TP
@@
 # should return TRUE
-identical(length(na.omit(ODX_allsites_2010$DENSQ1)), length(na.omit(ODX_allsites_2010$DENCOUNT)))
-# check how many sites with density square data
-length(na.omit(ODX_allsites_2010$DENCOUNT))
-#
+# identical(length(na.omit(ODX_allsites_2010$DENSQ1)), 
+# length(na.omit(ODX_allsites_2010$DENCOUNT)))
+# # check how many sites with density square data
+# length(na.omit(ODX_allsites_2010$DENCOUNT))
+# #
 # get counts when all artefacts counted
@@
 # check how many are ok
-length(na.omit(ODX_allsites_2010$TOTART)) # should have
-length(na.omit(ODX_allsites_2010$COUNTONLY)) # did get
+# length(na.omit(ODX_allsites_2010$TOTART)) # should have
+# length(na.omit(ODX_allsites_2010$COUNTONLY)) # did get
 #
@@
 # check how many are ok
-length(na.omit(ODX_allsites_2010$VISEST)) # should have
-length(na.omit(ODX_allsites_2010$VISCOUNT)) # did get
+# length(na.omit(ODX_allsites_2010$VISEST)) # should have
+# length(na.omit(ODX_allsites_2010$VISCOUNT)) # did get
 #
@@
 # check how many are ok
-length(na.omit(ODX_allsites_2010$artefacts))
+# length(na.omit(ODX_allsites_2010$artefacts))
 # 

### 2017-01-25 analysis/paper/paper.Rmd
site and artefact density

@@
 ```{r setup, include=FALSE}
+# image width should be 90 mm for 1 col, 140 mm for 1.5 col, 190mm for two cols
+fig_width_two_cols <- 190/25 # inch
+fig_width_one_col <- 90/25 # inch
+
 library(knitr)
-opts_chunk$set(echo = FALSE)
-options(warn=-1)
-# image width should be 90 mm for 1 col, 140 mm for 1.5 col, 190mm for two cols
+opts_chunk$set(echo = FALSE,
+               warning = FALSE,
+               message = FALSE,
+               dpi = 600)
+options(warn = -1)
+
 ```
@@
 
-In this study we demonstrate the use of gradients of change in archaeological materials over distance to explore forager choices of where to carry out activities in respect to environmental boundaries.
-We test a prediction that prehistoric foragers were sensitive to differences in resource structure as gradients across the landscape.
+In this study we ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of where to carry out activities in respect to environmental boundaries.
+We predict that prehistoric foragers were sensitive to differences in resource structure as gradients across the landscape.
 This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability that make comparative landscape analysis more tractable in many contexts.
@@
 
-```{r overviewmap, fig.cap="Map of region including the study area", fig.width=9}
+```{r overviewmap, fig.cap="Map of region including the study area", fig.width=fig_width_one_col}
 
@@
 
-```{r mapofstudyarea1, fig.cap="Map of study area with landforms and terrain patterns", fig.width=9}
+```{r mapofstudyarea1, fig.cap="Map of study area with landforms and terrain patterns", fig.width=fig_width_two_cols}
 
@@
 
-```{r mapofstudyarea2, fig.cap="Map of study area with landforms and terrain patterns", fig.width=9}
+```{r mapofstudyarea2, fig.cap="Map of study area showing the distribution of archaeological sites used in this study", fig.width=fig_width_two_cols}
 
@@
 In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the proposed Olympic Dam mining project.
-They used terrain patterns based on combining categories of different landforms and surface geology (Figure \@ref(fig:mapofstudyarea)) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982).
+They used terrain patterns based on combining categories of different landforms and surface geology (Figure \@ref(fig:mapofstudyarea1)) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982).
 The landscape consists of ancient stony plateau (or tableland) surfaces partly overlain by fields of longitudinal east-west dunes.
@@
 The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites  (Hiscock and Hughes 1983, Hiscock 1989).
-In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009 (Figure \@ref(fig:mapofstudyarea)).
-The data collected during this expansion phase is the basis for this study.
+In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009.
+The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).
 
@@
 
-```{r percentcoverageofeachterrainpattern, fig.cap="Terrain pattern percentage coverage of the 2007-2008 study area", fig.width=9}
+```{r percentcoverageofeachterrainpattern, fig.cap="Terrain pattern percentage coverage of the 2007-2008 study area", fig.width=fig_width_two_cols}
+
 # Terrain pattern percentage coverage of the 2007-2008 study area in Figure map
@@
 library(ggplot2)
-ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
+ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, -prop), prop)) + 
   geom_bar(stat = "identity",
@@
 
-About 44% of all sites were recorded in these two terrain patterns (Czs4: 33.8%, Czs5: 10.7%) with no other terrain pattern having more than 20% of all sites.
+```{r sitesinterrainpatterns, fig.cap="Site density in each terrain pattern.
+Annotation shows the result of a t-test to compare site densities in the two largest terrain patters, Czs4 and Czs5", fig.width=fig_width_two_cols}
+## sites in each TP
+sitesinterrainpatterns <- ODX_allsites_2010[, .N, by = GREGLANDF][, prop := N/sum(N) * 100]
+
+# prop of sites in Czs4
+sites_in_Czs4 <-  round(with(sitesinterrainpatterns, sitesinterrainpatterns[GREGLANDF == "Czs4", prop]), 1)
+
+# Prop of sites in Czs5 
+sites_in_Czs5 <- round(with(sitesinterrainpatterns, sitesinterrainpatterns[GREGLANDF == "Czs5", prop]), 1)
+
+# no other TP has more than N % of sites...
+sites_in_other_TPs <- round(max(with(sitesinterrainpatterns, sitesinterrainpatterns[!GREGLANDF %in% c("Czs4","Czs5") , prop])),1)
+
+## site density per TP
+
+# join TP areas to site counts per TP
+sites_in_terrain_patterns_and_areas <- ODX_TP_areas[sitesinterrainpatterns]
+# add column of DENSITY of sites per sq km 
+sites_in_terrain_patterns_and_areas <- 
+sites_in_terrain_patterns_and_areas[ , DENSITY := (N / V1) * (1000 * 1000), ]
+
+# test for difference in sites/km2 in Czs4 and Czs5
+# aggregate to get one row per TP_NO
+ODX_allsites_2010_per_TP_NO <- 
+ODX_allsites_2010[, list(.N, unique(AREA_M2)), by = TP_NO][, sites_per_sqkm := (N/V2) ]
+# split TP_NO col to get TPs in their own col
+ODX_allsites_2010_per_TP_NO <- ODX_allsites_2010_per_TP_NO[, c("TP", "NO") := tstrsplit(TP_NO, "/") ]
+# subset Czs4 and Czs5
+ODX_allsites_2010_per_TP_NO_Czs4_and_Czs5 <- ODX_allsites_2010_per_TP_NO[ TP %in% c("Czs4", "Czs5") , ,  ]
+# t-test for each TP
+t_test_for_two_TPs <-  with(ODX_allsites_2010_per_TP_NO_Czs4_and_Czs5, 
+                            t.test( sites_per_sqkm ~  as.factor(TP)))
+t_test_for_two_TPs_p_value <- round(t_test_for_two_TPs$p.value, 3)
+t_test_for_two_TPs_t_value <- round(t_test_for_two_TPs$statistic, 3)
+t_test_for_two_TPs_df_value <- round(t_test_for_two_TPs$parameter, 3)
+# summary values for
+
+# density of sites in Czs4
+site_density_in_Czs4 <-  round(mean(with(ODX_allsites_2010_per_TP_NO_Czs4_and_Czs5, ODX_allsites_2010_per_TP_NO_Czs4_and_Czs5[TP == "Czs4", sites_per_sqkm]), 1))
+
+# density of sites in Czs5 
+site_density_in_Czs5 <-  round(mean(with(ODX_allsites_2010_per_TP_NO_Czs4_and_Czs5, ODX_allsites_2010_per_TP_NO_Czs4_and_Czs5[TP == "Czs5", sites_per_sqkm]), 1))
+
+## artefact density per TP
+
+## plot site densities and artefact densities per TP
+g <- ggplot(ODX_allsites_2010_per_TP_NO,
+       aes(reorder(TP,
+           -sites_per_sqkm),
+           sites_per_sqkm)) +
+  geom_boxplot() +
+  theme_bw(base_size = text_size) +
+  theme(axis.text.x = element_text(size = text_size),
+        axis.text.y = element_text(size = text_size),
+        axis.title.x = element_text(size = text_size),
+        axis.title.y = element_text(size = text_size, angle = 90)) + 
+  ylab("Sites per square kilometer") +
+  xlab("Terrain pattern") +
+  coord_fixed(ratio =  0.04)
+
+# add horizontal bars
+hline <- 150
+g <-  g + geom_segment(aes(x = 2, # Czs4
+                            y = hline, 
+                            xend = 10,  # Czs5
+                            yend = hline))
+
+# add vertical bars
+g <-   g + geom_segment(aes(x = c(2), # Czs4
+                            y = c(hline), 
+                            xend = c(2),  # Czs5
+                            yend = c(90)))
+g <-   g + geom_segment(aes(x = c(10), # Czs4
+                            y = c(hline), 
+                            xend = c(10),  # Czs5
+                            yend = c(40)))
+# add annotation
+g <- g + annotate("text", 
+                  x = 6,
+                  y = 160, 
+                  label = paste0("t(", t_test_for_two_TPs_df_value, ") = ",
+                                 t_test_for_two_TPs_t_value, ", p = ",
+                                 t_test_for_two_TPs_p_value))
+# show plot
+g
+```
+
+```{r artefactdensity}
+# get counts from density squares
+ODX_allsites_2010$DENCOUNT <- with(ODX_allsites_2010, (as.numeric(levels(DENSQ1)[DENSQ1]))+(as.numeric(levels(DENSQ2)[DENSQ2]))+(as.numeric(levels(DENSQ3)[DENSQ3]))+(as.numeric(levels(DENSQ4)[DENSQ4]))+(as.numeric(levels(DENSQ5)[DENSQ5]))+(as.numeric(levels(DENSQ6)[DENSQ6]))+(as.numeric(levels(DENSQ7)[DENSQ7]))+(as.numeric(levels(DENSQ8)[DENSQ8])+(as.numeric(levels(DENSQ9)[DENSQ9]))/9)*((as.numeric(levels(DIMSIT1)[DIMSIT1]))*(as.numeric(levels(DIMSIT2)[DIMSIT2]))))
+# check if all sites with density square data were processed
+# should return TRUE
+# identical(length(na.omit(ODX_allsites_2010$DENSQ1)), 
+# length(na.omit(ODX_allsites_2010$DENCOUNT)))
+# # check how many sites with density square data
+# length(na.omit(ODX_allsites_2010$DENCOUNT))
+# #
+# get counts when all artefacts counted
+ODX_allsites_2010$COUNTONLY <- with( ODX_allsites_2010, as.numeric(levels(TOTART)[TOTART]) )
+# check how many are ok
+# length(na.omit(ODX_allsites_2010$TOTART)) # should have
+# length(na.omit(ODX_allsites_2010$COUNTONLY)) # did get
+#
+# get counts from when only a VISTEST was taken:
+ODX_allsites_2010$VISCOUNT<-with(ODX_allsites_2010, (as.numeric(levels(VISEST)[VISEST]))*((as.numeric(levels(DIMSIT1)[DIMSIT1]))*(as.numeric(levels(DIMSIT2)[DIMSIT2]))))
+# check how many are ok
+# length(na.omit(ODX_allsites_2010$VISEST)) # should have
+# length(na.omit(ODX_allsites_2010$VISCOUNT)) # did get
+#
+ODX_allsites_2010$artefacts <- as.numeric(with(ODX_allsites_2010, 
+      ifelse(COUNTONLY > 1, COUNTONLY,
+      ifelse(VISCOUNT  > 1 ,  VISCOUNT, 
+      ifelse(DENCOUNT  > 1 , DENCOUNT, "XX")))))
+#                                          
+# have a bit of a look 
+# ODX_allsites_2010[ODX_allsites_2010$VISCOUNT  != 0 ,]
+# ODX_allsites_2010[ODX_allsites_2010$DENCOUNT  != 0 ,]
+# ODX_allsites_2010[ODX_allsites_2010$COUNTONLY != 0 ,]
+#
+# summary(ODX_allsites_2010$VISCOUNT)
+# summary(ODX_allsites_2010$DENCOUNT)
+# summary(ODX_allsites_2010$COUNTONLY)
+#  
+# check how many are ok
+# length(na.omit(ODX_allsites_2010$artefacts))
+# 
+# check how many sites in total
+# summary( ODX_allsites_2010[,1:10] )
+# 10636 sites in data, 133 are hearths so 10503 should 
+# have artefact data... in fact 9859 do, so some 644 sites
+# have problems with artefact data, that's only 0.6% so that's
+# not too bad.
+#
+# get total number of artefacts in each TP
+artefacts.per.tp <- aggregate(data = ODX_allsites_2010, artefacts ~ GREGLANDF, FUN = sum)
+
+# check of artefacts per sq km for each TP
+artefacts.per.sqkm <- merge(artefacts.per.tp, ODX_TP_areas)
+artefacts.per.sqkm$artefacts.per <- with(artefacts.per.sqkm, artefacts/V1*1000*1000)
+
+# extract values for text 
+artefacts_per_sqkm_Czs4 <- 
+prettyNum(round(with(artefacts.per.sqkm, artefacts.per.sqkm[GREGLANDF == "Czs4", 'artefacts.per']), 0), big.mark = ",")
+
+artefacts_per_sqkm_Czs5 <- 
+prettyNum(round(with(artefacts.per.sqkm, artefacts.per.sqkm[GREGLANDF == "Czs5", 'artefacts.per']), 0), big.mark = ",")
+
+```
+
+About 44% of all sites were recorded in these two terrain patterns (Czs4: `r sites_in_Czs4`%, Czs5: `r sites_in_Czs5`%) with no other terrain pattern having more than `r sites_in_other_TPs`% of all sites.
 When considered as homogenous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
-As predicted (TABLE 1), sites are more frequent in fields of widely spaced dunes (Czs4) with 53 sites/km2 compared with more closely spaced dunes (Czs5) which have only 21 sites/km2 (t = 5.29, df = 8.6, p-value = 0.0005).
-Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at 35,789 artefacts/km2, compared with closely spaced dunes Czs5 at 2,702 artefacts/km2.
+As predicted (TABLE 1), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km2 compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km2 (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns).
+Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km2, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km2.
 

### 2017-01-26 analysis/paper/paper.Rmd
z-scores and sand dune spacing

@@
 Archaeological material is found across the broader region and partly reflects the long period of human occupation, the wide range of locally available resources and the extremely low rates of sediment movement that might otherwise obscure surface artefacts.
-Optically stimulated luminescence dating of sediments underling buried artefacts at one typical site we investigated indicates a terminus post quem  of 5 k BP for the artifact bearing deposit (Sullivan, Field et al.
+Optically stimulated luminescence (OSL) dating of sediments underling buried artefacts at one typical site we investigated indicates a terminus post quem of 5 k BP for the artifact bearing deposit (Sullivan, Field et al.
 2012).
-Such an age is consistent with the majority of dates for human occupation of sand deserts elsewhere in Australia, where evidence has suggested a substantial expansion of human activity in the last few millennia (Veth, Smith et al.
+Excavations at a second site produced five OSL ages for an open-air artefact quarrying/knapping location.
+These ages indicate that the site was used over relatively short time period, at about 2.3–2.4 ka, and again within the last 1 ka (Sullivan et al.
+2014).
+No organic materials have been found at any of the open sites or in the excavated deposits.
+Such ages are consistent with the majority of dates for human occupation of sand deserts elsewhere in Australia, where evidence has suggested a substantial expansion of human activity in the last few millennia (Veth, Smith et al.
 2001, Smith 2006).
-A single dated site is insufficient to draw conclusions beyond the local level, however the distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact scatters recorded during the survey (Hughes and Hiscock 2005).
-This mid to late Holocene period was also a time of broad-scale dune stability, so site visibility is not strongly controlled by dune movement.
+A pair of dated sites are insufficient to draw conclusions beyond the local level, however the distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact scatters recorded during the survey (Hughes and Hiscock 2005).
+A study of 78 OSL ages from geological contexts in the study area has shown that the mid to late Holocene period was also a time of broad-scale dune stability (Hughes et al.
+2014).
+This means that site preservation and visibility is not strongly controlled by dune movement.
 In 2013 we returned to sites that had been recorded 25 years earlier and found them to be in the same state as originally recorded, indicating that site visibility has changed little over 25 years.
@@
 ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
+ODX_allsites_2010 <- ODX_allsites_2010[ !is.na(ODX_allsites_2010$GREGLANDF), ]
 ODX_allsites_2010 <- data.table(ODX_allsites_2010)
@@
 #
-# Plot of TP areas (figure 4 of JFA paper)
+# Plot of TP areas 
 text_size <- 13
@@
 
-About 44% of all sites were recorded in these two terrain patterns (Czs4: `r sites_in_Czs4`%, Czs5: `r sites_in_Czs5`%) with no other terrain pattern having more than `r sites_in_other_TPs`% of all sites.
+About `r round(sites_in_Czs4 + sites_in_Czs5,0)`% of all sites were recorded in these two terrain patterns (Czs4: `r sites_in_Czs4`%, Czs5: `r sites_in_Czs5`%) with no other terrain pattern having more than `r sites_in_other_TPs`% of all sites.
 When considered as homogenous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
@@
 
-FIG 5 depicts median distances for site distances from terrain pattern boundaries where the dunefields are on average more than 1 km across (FIG 2).
-The general pattern is clear, with distances being substantially further in widely spaced dunefields such landform type 4 than in more closely spaced dunefields such as landform types 5 and 6.
-An examination of FIG 2 shows that this relationship is not dependent simply on the size or shape of the terrain patterns.
-
-FIG 6 shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform.
-These data give support to the prediction that resource-rich landscapes such as Csz4 have sites located away from their margin and far inside the terrain pattern areas while landscapes with lower resource levels such as Csz5 commonly have sites closer to their margins and fewer sites further in their interiors.
+```{r distanceofsitetoboundary, fig.cap="Distribution of distances of archaeological sites to terrain pattern bounaries.
+Data points show distance values for individual sites.", fig.width=fig_width_one_col}
+
+ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
+ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
+# order the levels of GREGLANDF for plotting
+# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
+levels(ODX_allsites_2010.s$GREGLANDF)
+ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
+  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])
+
+# plot
+library(ggplot2)
+library(ggforce)
+ggplot(ODX_allsites_2010.s, 
+       aes(reorder(GREGLANDF, -RASTERVALU),
+           RASTERVALU)) + 
+  geom_boxplot() + 
+  scale_y_log10(breaks = seq(0,1400,200)) + 
+  theme_bw(base_size = text_size-1) +
+  theme(axis.title.y = element_text(angle = 90))  +
+  geom_sina(alpha = 0.03) +
+  xlab("Terrain pattern") +
+  ylab("Distance of site to boundary (m)")
+```
+
+Figure \@ref(fig:distanceofsitetoboundary) depicts median distances for site distances from terrain pattern boundaries where the dunefields are on average more than 1 km across (cf.
+Figure \@ref(fig:mapofstudyarea1)).
+The general pattern is clear, with distances being substantially further in widely spaced dunefields such landform type 4 (e.g.
+A4, Czs4 and Q4) than in more closely spaced dunefields such as landform types 5 and 6 (e.g.
+Czs5, Q5and Q5).
+Visual inspection of Figure \@ref(fig:mapofstudyarea1) shows that this relationship is not dependent simply on the size or shape of the terrain patterns.
+
+```{r sitesinbuffers, fig.cap="Site density by distance to the boundary of the terrain pattern for Czs4 and Czs5.
+Data points show site density values for each buffer zone.", fig.width=fig_width_two_cols}
+
+# start with OUTER buffers
+bd_files <- list.files("../data/outer_buffers", full.names = TRUE)
+# list in, dataframe out
+library(plyr)
+library(foreign)
+bd.data <- ldply(bd_files, read.dbf)
+bd.data$SiteDens <- with(bd.data, Count_/buf_area * (1000*1000))
+# exclude buffers with less than n sites since they 
+# skew the density calculations 
+n = 15
+bd.data <- bd.data[bd.data$Count_ > n, ]
+# unique(bd.data$BUFF_DIST)
+bd.data$Buf <- as.factor(with(bd.data, ifelse(BUFF_DIST == 100, '000-100', ifelse(BUFF_DIST == 200, '100-200', ifelse(BUFF_DIST == 300, '200-300', ifelse(BUFF_DIST == 400, '300-400', ifelse(BUFF_DIST == 500, '400-500','NA')))))))
+
+#  how many sites in each set?
+# sum(na.omit(subset(bd.data, GREGLANDF=='Czs4')$Count_))
+# sum(na.omit(subset(bd.data, GREGLANDF=='Czs5')$Count_))
+
+# now check the boundary densities for INNER buffers
+inbd.files <- list.files("../data/inner_buffers", full.names = TRUE)
+# list in, dataframe out
+library(plyr)
+library(foreign)
+inbd.data <- ldply(inbd.files, read.dbf)
+inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
+inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
+unique(inbd.data$BUFF_DIST)
+inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))
+
+#  how many sites in each set?
+# sum(as.numeric((subset(inbd.data[as.numeric(inbd.data$Join_Count),], GREGLANDF=='Czs4'))$Join_Count))
+# sum(as.numeric((subset(inbd.data[as.numeric(inbd.data$Join_Count),], GREGLANDF=='Czs5'))$Join_Count))
+
+# combine inner with outer buffers
+inner <- subset(inbd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')[,c('GREGLANDF','SiteDens','Buf')]
+outer <- subset(bd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')  [,c('GREGLANDF','SiteDens','Buf')]
+# colnames(outer)[2] <-  'mean_SiteDens'
+inner_outer <- na.omit(rbind(inner, outer))
+# put the boundary distances in order
+# levels(inner_outer$Buf)
+inner_outer$Buf <- factor(inner_outer$Buf, 
+  levels(inner_outer$Buf)[c(5,4,3,2,1,6,7,8,9,10)])
+# remove rows with zero for SiteDens
+inner_outer <- inner_outer[inner_outer[, 'SiteDens']!=0, ]  
+# plot inner and outer together
+library(ggplot2)
+library(grid)
+library(scales)
+ggplot(inner_outer,  
+       aes(Buf, 
+           SiteDens)) + 
+  geom_point(size = 2) + 
+  facet_grid(GREGLANDF ~ .) + 
+  scale_y_log10() + 
+  geom_smooth(aes(group = 1), 
+              fill = "grey80", 
+              colour = "grey") +
+  theme_bw() + 
+  theme(
+  axis.title.x = element_text(vjust = -0.5, 
+                              size = 15),
+  axis.title.y=  element_text(size = 15, 
+                              angle=90)) + 
+  xlab("Distance from boundary in meters \n(negative values are interior distances, \npositive values are exterior distances)") + 
+  ylab("Sites per square km")
+
+# two-way ANOVA
+inner_outer$Buf <- as.character(inner_outer$Buf)
+two_way_anova_data <- inner_outer[inner_outer$GREGLANDF %in% c('Czs5','Czs4') &
+                                      !inner_outer$Buf %in% c("-100-000", "000-100"), ]
+two_way_anova_summary <- summary(fit <- aov(data = two_way_anova_data, 
+                   SiteDens ~  GREGLANDF * Buf))[[1]]
+two_way_anova_F  <- round(two_way_anova_summary$`F value`[3],3)
+two_way_anova_df  <- paste0(round(two_way_anova_summary$Df[c(1,2)],0), collapse = ", ")
+two_way_anova_p <- round(two_way_anova_summary$`Pr(>F)`[3],3)
+
+```
+
+```{r zscoresonsitedensitybyzones, fig.cap="Z-scores for site density in interior buffer zones for Czs4 and Czs5", fig.width=fig_width_one_col}
+
+# Czs5
+# site densities in 0-100 m inner buffer divided by total density of TP
+Czs5.inner.100 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-100-000",])
+Czs5.inner.100.ratio <- mean(Czs5.inner.100$SiteDens)/with(sites_in_terrain_patterns_and_areas, sites_in_terrain_patterns_and_areas[GREGLANDF == "Czs5",])$DENSITY
+# site densities in 100-200 m inner buffer divided by total density of TP
+Czs5.inner.200 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-200-100",])
+Czs5.inner.200.ratio <- mean(Czs5.inner.200$SiteDens)/with(sites_in_terrain_patterns_and_areas, sites_in_terrain_patterns_and_areas[GREGLANDF == "Czs5",])$DENSITY
+# site densities in 200-300 m inner buffer divided by total density of TP
+Czs5.inner.300 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-300-200",])
+Czs5.inner.300.ratio <- mean(Czs5.inner.300$SiteDens)/with(sites_in_terrain_patterns_and_areas, sites_in_terrain_patterns_and_areas[GREGLANDF == "Czs5",])$DENSITY
+
+# Czs4
+# site densities in 0-100 m inner buffer divided by total density of TP
+Czs4.inner.100 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-100-000",])
+Czs4.inner.100.ratio <- mean(Czs4.inner.100$SiteDens)/with(sites_in_terrain_patterns_and_areas, sites_in_terrain_patterns_and_areas[GREGLANDF == "Czs4",])$DENSITY
+# site densities in 100-200 m inner buffer divided by total density of TP
+Czs4.inner.200 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-200-100",])
+Czs4.inner.200.ratio <- mean(Czs4.inner.200$SiteDens)/with(sites_in_terrain_patterns_and_areas, sites_in_terrain_patterns_and_areas[GREGLANDF == "Czs4",])$DENSITY
+# site densities in 200-300 m inner buffer divided by total density of TP
+Czs4.inner.300 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-300-200",])
+Czs4.inner.300.ratio <- mean(Czs4.inner.300$SiteDens)/with(sites_in_terrain_patterns_and_areas, sites_in_terrain_patterns_and_areas[GREGLANDF == "Czs4",])$DENSITY
+
+# combine both for plotting
+Czs4 <- c(Czs4.inner.100.ratio, Czs4.inner.200.ratio, Czs4.inner.300.ratio)
+Czs5 <- c(Czs5.inner.100.ratio, Czs5.inner.200.ratio, Czs5.inner.300.ratio)
+bound <- data.frame(cbind(Czs4, Czs5))
+bound$dist <- c("0-100", "100-200", "200-300")
+library(reshape2)
+bound.m <- melt(bound)
+
+# convert to z-scores....
+Czs4.sc <-  scale(Czs4)
+Czs5.sc <-  scale(Czs5)
+bound.sc <- data.frame(cbind(Czs4.sc, Czs5.sc))
+bound.sc$dist <- c("0-100", "100-200", "200-300")
+colnames(bound.sc) <- c("Czs4", "Czs5", "distance")
+library(reshape2)
+bound.sc.m <- melt(bound.sc)
+colnames(bound.sc.m) <- c("distance", "Terrain Pattern", "density of sites")
+
+# dot-plot
+library(viridis)
+ggplot(bound.sc.m, 
+       aes(x=distance, 
+           y=`density of sites`,
+           shape = `Terrain Pattern`,
+           colour = `Terrain Pattern`)) + 
+  geom_point(size = 5)  +
+  geom_line(aes(group = `Terrain Pattern`)) +
+  theme_bw(base_size = text_size) +
+  coord_flip() +
+  scale_color_viridis(discrete = TRUE) +
+  theme(axis.title.y = element_text(angle = 90)) + 
+  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
+  ylab("z-score of site density in buffer area")  +
+  labs(shape="Terrain\nPattern", colour="Terrain\nPattern")
+
+```
+
+Figure \@ref(fig:sitesinbuffers) shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform.
+The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding the 0-100 m zone).
+When we use z-scores to standardize for differences in site density between two terrain patterns, we see that the pattern is even more stark.
+Figure \@ref(fig:zscoresonsitedensitybyzones) shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform.
+This also minimizes the effects of differences in the sizes of landscape units by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns.
+These analyses give support to the prediction that resource-rich landscapes such as Csz4 have sites located away from their margin and far inside the terrain pattern areas while landscapes with lower resource levels such as Csz5 commonly have sites closer to their margins and fewer sites further in their interiors.
 
@@
 
-Confirmation of the model predictions were facilitated by two survey characteristics.
+In our measurements of site location and site densities, we observed that subtle differences in physical features of the landscape had significant effects on where people were active on the study area during the Holocene.
+Testing of our model's predictions were facilitated by two characteristics that are distinctive to this project.
 One is that the extensive survey area, the intensive level of survey and the resulting large sample size of archaeological sites, meant that observation of human choices and behaviour could be obtained at a very high spatial resolution.
 The second is that objective and robust environmental classifications, creating well-defined terrain patterns, made it possible to quantify landscape diversity in the study area at a high spatial resolution and reduced false associations that could result from less accurate environmental mapping.
-Reprise testing of model predictions…
-Reprise testing of boundary effect predictions…
-
-This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to the different economic potential of dissimilar terrain patterns.
+
+We found that in there were both more archaeological sites per square kilometer, and more artefacts per square kilometer, in landforms with more widely spaced dunes.
+This is consistent with our model, and our field observations that more and denser sites tended to occur occur near exposures of good-quality flakeable stone that are often found in interdunal swales.
+In areas where the sand dunes are more closely spaced, such as landform types 5 and 6, the dune flanks flow into each other, leaving few or no swales between the dunes where surface outcrops of flakable stone might be exposed.
+This differential availability of flakable stone may be an important factor that influenced ancient peoples' use of the landscape.
+
+Our analysis of boundary effects shows people were senstive to subtle gradients of change in the landscape.
+We found that sites were generally located futher from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
+This indicates that people preferred to be located close to the center of areas with widely-spared dunes, and on the edges of the areas with closely spaced dunes.
+Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.
+This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns.
 This claim is axiomatic to human behavioural ecology in which different landscapes are commonly compared as homogenous units to improve analytical tractability.
@@
 It is also noteworthy that a relatively subtle geographic differences between landscapes – the different spacing of sand dunes between Czs4 and Czs5 – has a marked visible effect on the distribution of archaeological sites.
+We speculate that this is due to the relationship between dune spacing and the availability of flakable stone.
+The causes of variation in dune spacing are poorly understood.
+Previous work has associated sand transport rates and directions, wind regime characteristics, grain size patterns, topography, and substrates with dune spacing, however there are no simple relationships (Lancaster 1988; Wasson and Hyde 1983; Wasson et al.
+1988).
 
@@
 Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day.
-This metric is useful because it is easy to measure without burdening field workers with additional data logging tasks.
+This metric is useful because it is easy to measure without burdening field workers with additional data-logging tasks.
 However, we need to take into account possible confounding factors, such as travel time to the day’s survey area, the size and density of the sites in the day’s survey area, skill and efficiency of the individual archaeologist and the sampling strategy.
+
+```{r mobilegisefficiency}
+
+# checking benefit of mobile GIS
+dat <- read.table(text = "
+Year Pre.2007	2007	2008	2009
+Area.surveyed.(km2)	35	95	275	145
+Rate.of.survey.(km2/person/day)	0	0.3	0.3	0.3
+Total.no.persondays	0 317	917	483
+Approximate.number.of.sites.recorded	665	3900	6600	5500
+No.sites.recorded/person/day	 3.5	12.3	7.2	11.3
+Frequency.of.occurrence.of.sites/km2	19	41	24	38", row.names = 1, header = TRUE)
+
+dat_t <- data.frame(t(dat))
+
+# linear model for sites recorded by density of sites
+mobile_gis_efficiency_lm <- summary(lm(dat_t[,5] ~ dat_t[,6]))
+
+# subset 2007-2009 data
+dat_t_s <- dat_t[2:4,]
+# linear model for sites recorded by density of sites, 2007-2009
+mobile_gis_efficiency_lm_2007_2009 <- summary(lm(dat_t_s[,5] ~ dat_t_s[,6]))
+
+# predict given 19 sites/km2
+19 * 0.297  + 0.043 
+```
 
@@
 Further, the average site size for the pre-2007 data is less than the 2007-2009 data.
+These conditions should favor a higher rate of recording for the 2007-2009 data.
 TABLE 5 shows that prior to adopting the mobile GIS in 2007 we recorded 3.5 sites per person per day.
@@
 
+Hughes, P., M.
+Sullivan, F.
+Williams, N.
+Spooner and D.
+Questiaux (2014).	Sand dune formation over the last 140 ka in the Roxby dunefield, arid northern South Australia.
+Quaternary Australasia 31(1):11.
+
 Hughes, P., P.
@@
 Sheffield, Sheffield Archaeological Monographs: 29-44.
+
+Lancaster, N. (1988). Controls of eolian dune size and spacing. Geology, 16(11), 972-975.
 
@@
 
+Sullivan, M., P. Hughes, A. Mosig Way and N. Spooner (2014).	Prehistoric mining at Olympic Dam in arid South Australia. Archaeology in Oceania 49(1):43-55.
+
 Sullivan, M., T.
@@
 Wagtendonk, A. J. and R. A. M. De Jeu (2007). "Sensible field computing: Evaluating the use of mobile GIS methods in scientific fieldwork." Photogrammetric Engineering and Remote Sensing 73(6): 651.
+
+Wasson, R.
+J., Fitchett, K., Mackey, B., & Hyde, R.
+(1988).
+Large‐scale patterns of dune type, spacing and orientation in the Australian continental dunefield.
+The Australian Geographer, 19(1), 89-104.
+
+Wasson, R. J., & Hyde, R. (1983). Factors determining desert dune type. Nature, 304(5924), 337-339.
 
@@
   axis.title.y=  element_text(size = 15, 
-                              angle=90), 
-  plot.margin = unit(c(1,1,2,2), "lines")) + 
+                              angle=90)) + 
   xlab("Distance from boundary in meters \n(negative values are interior distances, \npositive values are exterior distances)") + 
@@
 # a bit wacky...
+
+# two-way ANOVA
+inner_outer$Buf <- as.character(inner_outer$Buf)
+two_way_anova_data <- inner_outer[inner_outer$GREGLANDF %in% c('Czs5','Czs4') &
+                                      !inner_outer$Buf %in% c("-100-000", "000-100"), ]
+summary(fit <- aov(data = two_way_anova_data, 
+                   SiteDens ~  GREGLANDF * Buf))
 
@@
 
+# dot-plot
+library(viridis)
+ggplot(bound.sc.m, 
+       aes(x=distance, 
+           y=`density of sites`,
+           shape = `Terrain Pattern`,
+           colour = `Terrain Pattern`)) + 
+  geom_point(size = 5)  +
+  geom_line(aes(group = `Terrain Pattern`)) +
+  theme_bw(base_size = text_size) +
+  coord_flip() +
+  scale_color_viridis(discrete = TRUE) +
+  theme(axis.title.y = element_text(angle = 90)) + 
+  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
+  ylab("z-score of site density in buffer area")  +
+  labs(shape="Terrain\nPattern", colour="Terrain\nPattern")
+
 # checking benefit of mobile GIS

### 2017-01-26 analysis/paper/SOM.Rmd
got the text tables in

---
title: 'Supplementary Online Material for Landform boundary effects on Holocene forager landscape use in arid South Australia'
author:
- Ben Marwick (corresponding author)
- Peter Hiscock 
- Philip Hughes 
- Marjorie Sullivan
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
---

```{r, echo = FALSE}
# extract code from main manuscript file into a new R file
# so we can run select chunks from that in this file
library(knitr)
opts_chunk$set(echo = FALSE, 
               messages = FALSE, 
               warnings = FALSE, 
               dpi = 600)
               # fig.width = 12
               #fig.show = 'hide'

invisible(purl("paper.Rmd",  
               output="temp",
               quiet=TRUE))
read_chunk("temp")
```

Affiliations:

Ben Marwick (corresponding author)
University of Washington
Denny Hall 117, Box 353100
Seattle, WA 98195-3100 USA

t. (+1) 206.552.9450 
f. (+1) 206.543.3285
bmarwick@uw.edu

Peter Hiscock   
The University of Sydney, Sydney NSW, Australia

Philip Hughes    
Department of Archaeology, Flinders University, GPO Box 2100, Adelaide, SA, 5001

Marjorie Sullivan    
The Australian National University, Canberra ACT 0200, Australia and 
Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia

##### . 
<!--- page break --->

## Introduction

In this supplementary file we describe our decision to use a mobile GIS, and our field data collection workflow, and provide diagnostics for the linear models presented in the main text.

### Mobile GIS and field data collection workflow

The decision to use a mobile GIS was based on four specific advantages over paper-based systems of the numerous benefits listed by Tripcevich (2004a; 2004b), Clarke (2004) and Wagtendonk and De Jeu (2007):

1.	Mobile GIS results in substantial efficiencies when data collection is intensive, for example when many sites will be recorded per person per day. Survey planning indicated that an average of 10 sites per person per day would be recorded and up to 100 sites per person per day would be possible. If a paper-based recording system had been adopted, only a fraction of the more than 16,000 sites located would have been able to be recorded in the time and with the resources available. An additional related advantage is that the data are stored in a form that makes them immediately available for spatial or statistical analyses, and interim reporting.

2.	A highly structured digital recording system would minimize recording errors and eliminate the potential for errors and the time-costs of transcribing data from field notes or recording sheets into a computer (Schneiderman-Fox & Pappalardo 1996). Given the large number of sites expected, the time saved by eliminating transcription was expected to be substantial.

3.	There were other digital datasets, including fence and road locations, previously surveyed areas and recorded sites, and terrain units that covered all of the survey area. Over 95% of the survey was covered by good quality recent air photography. Some of the thematic maps had been digitized, other data were able to be digitized and stored as shapefiles by the mining company, BHP Billiton. While paper maps can provide this information, and may be easier to read than a small screen in bright sun, a mobile GIS gives more immediate context awareness (Pascoe et al. 1999). In particular the mobile GIS enables a user to determine location relative to other features and automates calculations relating to these features such as distance, direction and area.

4.	Mobile GIS is most effective on large multi-year projects where the time and expense of setup and training is offset by fieldworkers becoming increasingly efficient with the technology over time. The mining project has multiple stakeholders, including indigenous groups, company executives, government authorities as well as the survey team members, and mobile GIS data are collected in a format that can be shared easily from a server-computer GIS.

The basic hardware and software used were similar to those described in previous studies (McPherron & Dibble 2003; Tripcevich 2004b; Given & Hyla 2006), including ruggedised handheld computers (TDS Recon and Nomad, described by Clegg et al. (2006) with a touchscreen interface and built-in GPS (Holux GM-270U and SiRFStar III). These PDAs run Windows Mobile 6 operating system and ESRI ArcPad software (Figure \@ref(fig:arcpad)). All of our configuration files, recording forms, and on-the-fly analysis scripts are archived online at <http://doi.org/10.5281/zenodo.11833>. Since our project began we have noted that newer models of the handheld computers have appeared, with improvements in processing speed, display quality, battery life, and a reduction in cost.

```{r arcpad, fig.cap="View of one page of the recording form showing typical data entry controls such as checkboxes, dropdown fields and text and number fields.", echo=FALSE}

knitr::include_graphics("../figures/arcpad.png")

```

ESRI’s suite of programs is ubiquitous in the private sector and in higher education (http://www.gisjobs.com/survey_all/), and extensive support is available from other users in freely accessible online fora. Free and open source alternatives are relatively new and have a much smaller user and developer community and fewer online resources (Stankovic et al. 2004; Steiniger & Hay 2009). While the use of proprietary software made the cost of the system relatively high, future mobile GIS projects could be run at a lower cost, for example using free open source data collection software designed for smartphones such as the Open Data Kit for Android (Anokwa et al. 2009). At the time we were planning our fieldwork there was no open source tool-chain that provided equivalent functionality to the ESRI software.

Although there is a variety of software for mobile GIS (Ancona et al. 1999; Ryan et al. 1999; Hall & Gray 2004; Döner & Yomralıoğlu 2008; Sobotkova 2013), ArcPad has four important features that made it preferable for this use:

1.	Real-time capture and display of GPS data in NMEA format. In addition to position, ArcPad can display detailed telemetry (e.g. speed, direction) and information about the quality of the data (number of satellites, position error). This is useful for navigation and coverage calculations during the survey.

2.	ArcPad can display the user’s position in context with other datasets. ArcPad uses the standard GIS convention of data layers to produce maps displaying multiple datasets. This is useful for showing the user custom datasets such as the location of tracks, fences, previously recorded and surveyed areas and detailed aerial photography.

3.	ArcPad allows rapid input of field data into a database. Data files created during the survey can be appended to a master record at the end of each day for rapid analysis, reporting and safe archiving.

4.	The extensive use of ArcPad in other industries and disciplines means there are numerous examples of customized ArcPad applications, for example urban maintenance, ecological survey purposes and disaster management, available at online support fora that were useful to adapt for the archaeological survey (cf. Tripcevich 2004b: 19).

The ArcPad recording forms, and Access database for storing and querying the data collected by the mobile GIS, were designed and tested by the authors over a total of two months prior to fieldwork. This startup phase required specialized skills in creating XML files that determined the user interface of the recording forms, writing VBScript that controls the actions that occurred in the forms (such as validation of input and on-the-fly calculations), and writing SQL queries to summarize field data in ArcGIS and Access on daily, weekly and other schedules for quality control and reporting. The main issues encountered during the startup period were the learning curves of these three languages (none of us had any prior experience), programming of input validation scripts to minimize invalid data entry (ie. preventing accidental attempts by a user to enter improbably large or small values for site attributes), and optimizing the user interface to combine readability of the ArcPad recording forms with a logical and efficient layout of the data collection controls (ie. text boxes, drop-down menus, radio buttons, etc.).

The original paper recording forms that had been used in the region (Hiscock & Hughes 1983) were revised extensively and translated into XML to design ArcPad forms using ESRI’s ArcPad Studio program and a free text editor (Notepad++). The XML format has the advantages of being widely used, well-documented and editable with the most basic text editing software (e.g. the forms can be edited directly on the PDAs in the field). As much as possible the data input options on the user interface were constrained using drop-down menus, check boxes and radio buttons rather than allowing flexible but error-prone data input using text boxes (FIG 3).

With more than 20 archaeologists using this system, most of whom had no prior experience with GIS, these input constraints helped speed data entry, standardize data collection, avoid invalid data and minimize data cleaning time. XML was also used for ArcPad configuration files to set communication variables between the GPS and the recording form, and to maximize the onscreen map area by hiding rarely used toolbar buttons. These configuration files were prepared during the startup period and afterwards only rarely altered.

In addition to the custom ArcPad recording form, scripts were written to calculate certain fields automatically based on entered data, check data input on the fly, prompt the user if certain conditions were not met and provide information to the user about what to do when recording a site. ArcPad forms were scripted using VBScript, which like XML, is widely used, well-documented and editable using any text editor. TABLE 3 describes some of the ways VBScript was used to control data input. VBScript and SQL were used to move the data from the PDA to a desktop computer, append data to a master database, update the mining company's heritage dataset to ensure newly recorded sites could be protected, check for duplicates and assign unique identifiers to sites. GPS tracklog data were recorded constantly during survey and archived with ArcGIS to prevent resurveying of the same area, to show survey coverage, and hence demonstrate the absence of sites from some parts of the landscape.

A typical workflow during survey would be for an archaeologist to use ArcGIS to select the area to be surveyed, export the GIS data for that selection (ie. aerial photograph raster file, previously recorded archaeological sites polygon shapefile, and road and fences polyline shapefile) from the desktop to the handheld computers, conduct the survey and enter data onto the mobile GIS when sites are recorded, then return to the desktop computer and download the day’s site and tracklog data into the Access database, validate the recorded sites by running prepared queries that summarize the day’s finds and visualize the tracklogs and recorded sites in ArcGIS, and finally backup the day’s data to three separate locations.

Sites located during the 2007-2009 surveys were recorded as single points into a point shapefile that was stored as a flat table. The co-ordinates for each point were averaged from ten positions captured by the GPS to improve the reliability of site location data to give an estimated horizontal accuracy of 2 m. Using this system, information about site location, environmental context, site size, number of artefacts, artefact raw materials, the presence and abundance of certain implement types (e.g. backed artefacts, tulas, grinding stones, hammerstones), the presence of hearths and attributes of knapping floors was collected. In more focused surveys since 2009 in areas outside the original 515 km2 survey area polygon shapefiles have been used to map the boundaries of sites. When recording polygon features a minimum of five positions are averaged at each vertex to improve location accuracy. Although ArcPad allows for direct capture of digital images and other data input (e.g. measurements of stone artefacts from digital calipers), these were not relevant to the survey goals and these features were not utilized.

Data collection with a mobile GIS enabled our survey team to collect the massive amount of data collected during survey fieldwork and substantially reduce transcription time and error (Fitzpatrick 2002; McPherron & Dibble 2003; Tripcevich 2004a; 2004b). Our experience is comparable to that of Döner and Yomralıoğlu (2008), who recorded the time spent on field data collection and transferring data using both mobile GIS and a paper and GPS system for a project mapping transport networks for rural settlements in Turkey. They observed that the most substantial time savings were gained during data transfer. The pattern they document is a three-fold increase in the rate of site recording with mobile GIS. The same general pattern of time saving was observed in our history of research at Olympic Dam. Our pre-2007 data was transferred from field notes to a computer in 8-10 days after the survey concluded, while our 2007-2009 data was transferred from the mobile GIS to the desktop computer in less than five minutes per fieldwork day.

The Olympic Dam sites fall into a narrow range of site types and stone artefacts within them are patterned in highly repeated ways that suited a systematic recording protocol. An important consideration in evaluating this use of mobile GIS and traditional methods is the regularity of the archaeological materials that were recorded. Projects that encounter high levels of variation, or cannot anticipate the kinds of data to be recorded, might benefit more from freeform data capture rather than the tightly controlled data entry fields that were an important advantage for this survey. Of course the value of more flexible data entry and subsequent problems for analysis would need to be balanced against the enhanced uniformity and accuracy of data when entry is tightly controlled and subject to automatic error checking and defined value limits.

As a final observation, it is noted that the disruption to fieldwork when the mobile GIS system fails may be underrepresented in the previous literature on mobile GIS in archaeology (McPherron & Dibble 2003; Tripcevich 2004b; but see McCaffrey et al. 2005). Mobile GIS is a complex and flexible setup that replaces many pieces of equipment (including tape measures, compass, notebooks) and if the handheld computer fails during a paperless survey then fieldworkers are limited in the tasks they can perform. This was a more serious concern in the early implementation of mobile GIS and more recent equipment and software are mostly reliable and robust. Breakdowns can never be completely avoided however and during the Olympic Dam survey their impacts were mitigated with multiple nightly data backups and keeping redundant functioning equipment on hand for immediate replacement of broken units, to minimize delays caused by waiting for repairs.  When the sites database was 'cleaned' at the end of the survey stage, serious errors (ie. fields containing data outside of a credible range or containing a mix of integers and non-integers, mostly likely due to accidents during data entry) and omissions (fields left blank or with data that is incomplete) were found to have occurred in fewer than 0.5% of the sites. Sites with problematic data were re-recorded and the unclean recorded were marked so they could be excluded from analysis.

Undoubtedly McCoy and Ladefoged (2009) were correct in forecasting that mobile GIS and related spatial technologies will continue to change archaeology and contribute to its progress, despite the steep learning curve for current systems. As hardware becomes more affordable it will become more widely used and better supported by communities of users, lowering the barriers to use by archaeologists. Similarly as archaeology reaches into new directions and new methods, such as information science (cf. Llobera 2010), mobile GIS is likely to expand beyond a replacement of paper forms to include the capture and in-field peer-to-peer collaboration with large and more complex datasets including imagery, video and remote sensing data.

### References cited

Ancona M, Dodero G, Mongiardino M and Traverso A 1999 Taking Digital Notes in the Field: the Archeo tool-set. In: JA Barceló, I Briz and A Vila (eds). Computer Applications and Quantitative Methods in Archaeology (BAR International Series 757) 117-121. Archaeopress, Oxford.

Anokwa Y, Hartung C, Brunette W, Borriello G and Lerer A 2009. Open source data collection in the developing world. Computer 42(10), 97-99.

Clarke K, C. 2004. Mobile Mapping and Geographic Information Systems. Cartography and Geographic Information Science 31, 131-136.

Clegg P, Bruciatelli L, Domingos F, Jones RR, De Donatis M and Wilson RW 2006. Digital geological mapping with tablet PC and PDA: A comparison. Computers & Geosciences 32(10), 1682-1698.

Doner F and Yomralioglu T 2008. Examination and comparison of mobile GIS technology for real time geo-data acquisition in the field. Survey Review 40(309), 221-234.

Fitzpatrick T 2002. Using ArcPad forms to streamline data collection. ArcUser July-Sept, 56-58.

Given O and Hyla S 2006. Forging Partnerships in Laos. Expedition 48(1), 38-39.

Hall M and Gray P 2004 Mobile Support for Team-Based Field Surveys. In: S Brewster and M Dunlop (eds). Mobile Human-Computer Interaction – MobileHCI 2004 105-217. Springer Berlin / Heidelberg.

Hiscock P and Hughes PJ 1983. One method of recording scatters of stone artefacts during site surveys. Australian Archaeology 17, 87-98.

Llobera M 2010. Archaeological Visualization: Towards an Archaeological Information Science (AISc). Journal of Archaeological Method and Theory, 1-31.

McCaffrey K, Jones R, Holdsworth R, Wilson R, Clegg P, Imber J, Holliman N and Trinks I 2005. Unlocking the spatial dimension: digital technologies and the future of geoscience fieldwork. Journal of the Geological Society 162(6), 927.

McCoy MD and Ladefoged TN 2009. New Developments in the Use of Spatial Technology in Archaeology. Journal of Archaeological Research 17(3), 263-295.

McPherron S and Dibble H 2003. Using computers In adverse field conditions: Tales from the Egyptian desert. The SAA Archaeological Record 3(5), 28-32.

Pascoe J, Ryan N and Morse D 1999. Issues in developing context-aware computing. Handheld and ubiquitous computing 1707/1999, 208-221.

Ryan N, Pascoe J and Morse DR 1999 FieldNote: extending a GIS into the field. In: JA Barcelo, I Briz and A Vila (eds). New Techniques for Old Times: Computer Applications in Archaeology, 1998 (BAR S755). Archaeopress, Oxford

Schneiderman-Fox F and Pappalardo AM 1996. A paperless approach toward field data collection: An example from the Bronx. SAA Bulletin 14(1), 1-18.

Sobotkova A 2013. "Android GIS – not ready for prime time (poster)." Computer Applications and Quantitative Methods in Archaeology  Retrieved 22 Sept 2014, from https://www.fedarch.org/wordpress/wp-content/uploads/2013/03/AdelaPoster.pdf.

Stankovic J, Neteler M and Flor R 2004. Mobile Wireless GRASS GIS for Handheld Computers Running GNU/Linux. Transactions in GIS 8(2), 225-233.

Steiniger S and Hay GJ 2009. Free and open source geographic information tools for landscape ecology. Ecological Informatics 4(4), 183-195.

Tripcevich N 2004a. Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey. Cartography and Geographic Information Science 31(3), 137-151.

Tripcevich N 2004b. Mobile GIS in archaeological survey. The SAA Archaeological Record(May).

Wagtendonk AJ and De Jeu RAM 2007. Sensible field computing: Evaluating the use of mobile GIS methods in scientific fieldwork. Photogrammetric Engineering and Remote Sensing 73(6), 651.

### Diagnostics for the linear models presented in the main text

```{r, ref.label="mobilegisefficiency", results = "hide"}
```

```{r}
library(ggplot2)
library(ggfortify)
```

The plots below show the diagnostic plots for the linear model predicting the number of sites recorded per person, per day from the frequencies of occurrence of sites km^2^ for all field seasons (pre-2007, 2007-2009)

```{r fig.width=10, fig.height=8}
autoplot(mobile_gis_efficiency_lm_out, 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10)
```

The plots below show the diagnostic plots for the linear model predicting the number of sites recorded per person, per day from the frequencies of occurrence of sites km^2^ for the 2007-2009 field seasons only:

```{r fig.width=10, fig.height=8}
autoplot(mobile_gis_efficiency_lm_2007_2009_out, 
         which = 1:6, 
         ncol = 2, 
         label.size = 3) + 
  theme_bw(base_size = 10)
```

### 2017-01-26 analysis/paper/paper.Rmd
got the text tables in

@@
 Philip Hughes    
-Department of Archaeology, Flinders University, GPO Box 2100, Adelaide, SA, 5001
+Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
 
 Marjorie Sullivan    
-The Australian National University, Canberra ACT 0200, Australia and 
 Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
@@
 
-A key step in understanding past human use of landscapes is investigating how people adapted their behaviours to the different landscapes they inhabited.
+A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the different landscapes they inhabited.
 Often this kind of analyses depends on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g.
@@
 ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes.
-Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972 ).
+Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972).
 This increased variety and density of plants and animals, often described as an ‘edge effect’, was claimed by Harris et al.
@@
 
+##### . 
+<!-- page break -->
+
+```{r table1}
+table1 <- read.csv("../data/table_1.csv")
+names(table1) <- c("Landform type", "Archaeological prediction")
+kable(table1, caption = "Landform types and predicted archaeological signature. Adapted from Hughes et al. (2011), Kinhill-Stearns Roger (1982) and Hughes and Sullivan (1984).")
+```
+
 Our study area at Olympic Dam, South Australia, offers a unique opportunity to study landscape interfaces for two key reasons.
-First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (TABLE 1).
+First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (Table \@ref(tab:table1)).
 The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes, making them convenient for spatial analysis of archaeological features.
@@
 
+##### . 
+<!-- page break -->
+
+```{r table2}
+table2 <- read.csv("../data/table_2.csv")
+names(table2) <- c("Geological regime", "Period", "Description", "Materials for artefact manufacture")
+kable(table2, caption = "Geological regimes in the study area and predicted archaeological importance. Adapted from Hughes et al. (2011).")
+```
+
 Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al.
@@
 In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts.
-The model results in a series of predictions about the archaeological record expected for each of six landform types (TABLE 1) and five geological regimes (TABLE 2).
+The model results in a series of predictions about the archaeological record expected for each of six landform types (Table \@ref(tab:table1)) and five geological regimes (Table \@ref(tab:table2)).
 These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.
@@
 The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites  (Hiscock and Hughes 1983, Hiscock 1989).
-In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009.
+In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009.
 The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).
@@
 Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 16,000 archaeological sites.
-We achieved close to 90% coverage of 353 km2.
+We achieved close to 90% coverage of 353 km^2^.
 The sites consist of stone artefact scatters (`r artefact_scatters`%), artefact scatters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al.
@@
 2014, Banning and Hitchings 2015).
+Our mobile GIS configuration files, recording forms, and scripts are archived online at <http://doi.org/10.5281/zenodo.11833>.
 A detailed description of our specific hardware, software, and justifications for our choices is presented in our SOM.
+
+### Investigating boundary effects
 
@@
 Our test of boundary effects focused on the archaeological evidence from resource-rich landform types/terrain patterns, such as Czs4, bordered by landscapes with lower resource levels, such as landform types 5 or 6.
-Because Czs4 and Czs5 dominate the 353 km2 sample area and contain the majority of sites, the analyses focused on comparing these two terrain patterns.
+Because Czs4 and Czs5 dominate the 353 km^2^ sample area and contain the majority of sites, the analyses focused on comparing these two terrain patterns.
 
@@
 The density of sites in each buffer zone was then divided by the average density for the entire terrain pattern to standardize the density value for the terrain pattern.
+
 Finally, z-scores were calculated to put the differences in site density on a comparable scale between the two terrain patterns.
@@
 
-# Results
+### Reproducibility and open source materials
+
+To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/xxx/xxx>.
+Also in this version-controlled compendium are the raw data for all the tests reported here, as well as additional regression diagnostics and power tests.
+All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository.
+In our SOM our code is released under the MIT licence, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).
+
+## Results
 
@@
 Most of the study area consists of widely spaced dunes – landform type 4 (`r LF4`%), moderately spaced dunes – landform type 5 (`r LF5`%) and closely spaced dunes – landform type 6 (`r LF6`%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P).
-Other terrain patterns make up less than 10% of the 353 km2 study area (Figure \@ref(fig:percentcoverageofeachterrainpattern)).
+Other terrain patterns make up less than 10% of the 353 km^2^ study area (Figure \@ref(fig:percentcoverageofeachterrainpattern)).
 
@@
 About `r round(sites_in_Czs4 + sites_in_Czs5,0)`% of all sites were recorded in these two terrain patterns (Czs4: `r sites_in_Czs4`%, Czs5: `r sites_in_Czs5`%) with no other terrain pattern having more than `r sites_in_other_TPs`% of all sites.
-When considered as homogenous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
-As predicted (TABLE 1), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km2 compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km2 (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns).
-Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km2, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km2.
+When considered as homogeneous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
+As predicted (Table \@ref(tab:table1)), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km^2^ compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km^2^ (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns).
+Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km^2^, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km^2^.
 
@@
 # this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
-levels(ODX_allsites_2010.s$GREGLANDF)
+# levels(ODX_allsites_2010.s$GREGLANDF)
 ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
@@
 inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
-unique(inbd.data$BUFF_DIST)
+# unique(inbd.data$BUFF_DIST)
 inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))
@@
 
-Figure \@ref(fig:sitesinbuffers) shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform.
+Figure \@ref(fig:sitesinbuffers) shows that in Czs4 landscapes the density of sites increases toward the center of the landform while in Czs5 landforms site density decreases towards the center of the landform.
 The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding the 0-100 m zone).
 When we use z-scores to standardize for differences in site density between two terrain patterns, we see that the pattern is even more stark.
-Figure \@ref(fig:zscoresonsitedensitybyzones) shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform.
+Figure \@ref(fig:zscoresonsitedensitybyzones) shows that in Czs4 landscapes the density of sites increases toward the center of the landform while in Czs5 landforms site density decreases towards the center of the landform.
 This also minimizes the effects of differences in the sizes of landscape units by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns.
@@
 
-Our analysis of boundary effects shows people were senstive to subtle gradients of change in the landscape.
-We found that sites were generally located futher from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
+Our analysis of boundary effects shows people were sensitive to subtle gradients of change in the landscape.
+We found that sites were generally located further from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
 This indicates that people preferred to be located close to the center of areas with widely-spared dunes, and on the edges of the areas with closely spaced dunes.
 Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.
+
 This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns.
-This claim is axiomatic to human behavioural ecology in which different landscapes are commonly compared as homogenous units to improve analytical tractability.
+This claim is axiomatic to human behavioral ecology in which different landscapes are commonly compared as homogeneous units to improve analytical tractability.
 The novel element of the data presented here is that the distribution of human activities varies substantially and predictably within as well as between terrain patterns.
@@
 Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM).
-TABLE 5 shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS.
+Table \@ref(tab:mobilegisefficiency) shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS.
 Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day.
@@
 
+display_table <- dat
+display_table$"" <- row.names(display_table)
+row.names(display_table) <- NULL
+display_table <- display_table[ , c(5, 1:4)]
+names(display_table) <- c("", "pre-2007", "2007", "2008", "2009")
+display_table[ , 1] <- gsub("\\.", " ", display_table[ , 1])
+display_table[ , 1] <- gsub("km2", "km^2^", display_table[ , 1])
+display_table[ , 1] <- gsub("persondays", "person-days", display_table[ , 1])
+
+# print table
+kable(display_table, caption = "Summary of archaeological survey productivity at Olympic Dam.
+The data for sites recorded per person per day for before 2007 come from a survey undertaken in 1982 by Philip Hughes and Peter Hiscock that recorded 176 sites in 50 person days.")
+
+# modelling these data
+
 dat_t <- data.frame(t(dat))
 
+# rate for pre-2007 and 2007-2009
+rate_for_pre_2007 <- dat_t$No.sites.recorded.person.day[1]  
+rate_for_2007_thru_2009 <-  round(mean(dat_t$No.sites.recorded.person.day[c(2:4)]),1)
+
+# the rate of recording is strongly correlated with the density of sites 
 # linear model for sites recorded by density of sites
-mobile_gis_efficiency_lm <- summary(lm(dat_t[,5] ~ dat_t[,6]))
-
+mobile_gis_efficiency_lm <- 
+  summary(mobile_gis_efficiency_lm_out <- lm(dat_t$No.sites.recorded.person.day ~ dat_t$Frequency.of.occurrence.of.sites.km2))
+mobile_gis_efficiency_lm_r_squared <- round(mobile_gis_efficiency_lm$adj.r.squared, 3)
+mobile_gis_efficiency_lm_p_value <- round(mobile_gis_efficiency_lm$coefficients[2, 4]  , 3)
+
+# a linear model for the 2007-2009 data
 # subset 2007-2009 data
@@
 # linear model for sites recorded by density of sites, 2007-2009
-mobile_gis_efficiency_lm_2007_2009 <- summary(lm(dat_t_s[,5] ~ dat_t_s[,6]))
-
-# predict given 19 sites/km2
-19 * 0.297  + 0.043 
+mobile_gis_efficiency_lm_2007_2009 <-
+ summary(mobile_gis_efficiency_lm_2007_2009_out <- 
+ lm(dat_t_s$No.sites.recorded.person.day ~ dat_t_s$Frequency.of.occurrence.of.sites.km2))
+
+# display model
+mobile_gis_efficiency_lm_2007_2009_est <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,1], 3)
+mobile_gis_efficiency_lm_2007_2009_est_error <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,2], 3)
+mobile_gis_efficiency_lm_2007_2009_inter <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,1], 3)
+mobile_gis_efficiency_lm_2007_2009_inter_error <-   round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,2], 3)
+mobile_gis_efficiency_lm_2007_2009_r_sq_adj <- round(mobile_gis_efficiency_lm_2007_2009$adj.r.squared, 3)
+mobile_gis_efficiency_lm_2007_2009_p_value <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2, 4], 3)
+
+# predict given 19 sites/km2 of pre-2007
+resulting_rate <- round(19 * mobile_gis_efficiency_lm_2007_2009_est + mobile_gis_efficiency_lm_2007_2009_inter, 2)
+
+# difference in rates, pre-2007 and 2007-2009
+difference_in_rates_pre_2007_and_2007_2009 <- 
+resulting_rate - rate_for_pre_2007
+
+# percentage increase
+percentage_increase <- round((resulting_rate - rate_for_pre_2007) / rate_for_pre_2007 * 100, 0)
 ```
@@
 These conditions should favor a higher rate of recording for the 2007-2009 data.
-TABLE 5 shows that prior to adopting the mobile GIS in 2007 we recorded 3.5 sites per person per day.
-With the mobile GIS our rate increased substantially to an average of 10.2 sites per person per day for 2007-2009.
-
-While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS because the rate of recording is strongly correlated with the density of sites (r = 0.97, p = 0.003).
-To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data (rate = site density * 0.297 (0.004)+ 0.062139 (0.146),  r2 = 0.999, p = 0.0002, standard errors in parentheses) and predicted the site recording rate given the density of 19 sites km-2 observed in the pre-2007 data.
-The resulting rate is 5.7 sites/person/day, which is what we would expect if a mobile GIS had been used to record sites in an area with 19 sites km-2.
-The difference of 2.2 sites/person/day between the actual pre-2007 average rate of 3.5 and the extrapolated average rate of 5.7 is most likely due to the use of mobile GIS.
-This represents a sixty percent increase in the site recording rate.
-This demonstrates an advantage for using mobile GIS in a context where data collection was intensive, a highly structured data collection form could be used, and the start-up and training costs could be amortized over multiple years.
+Table \@ref(tab:mobilegisefficiency) shows that prior to adopting the mobile GIS in 2007 we recorded `r rate_for_pre_2007` sites per person per day.
+With the mobile GIS our rate increased substantially to an average of `r rate_for_2007_thru_2009` sites per person per day for 2007-2009.
+
+While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS because the rate of recording is strongly predicted by the density of sites (R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`).
+To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data , which resulted in a rate of site recording that can by predicted by site density * `r mobile_gis_efficiency_lm_2007_2009_est` (`r mobile_gis_efficiency_lm_2007_2009_est_error`) + `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) (standard errors in parentheses).
+This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate given the density of 19 sites/km^2^ observed in the pre-2007 data.
+The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^.
+The difference of `r difference_in_rates_pre_2007_and_2007_2009` sites/person/day between the actual pre-2007 average rate of `r rate_for_pre_2007` and the predicted average rate of `r resulting_rate` is most likely due to the use of mobile GIS.
+This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.
+Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, a highly structured data collection form could be used, and the start-up and training costs could be amortized over multiple years.
 
@@
 
-The substantive archaeological contribution of this study, resulting from the use of mobile GIS to record archaeological data, has been to show that when choosing habitations prehistoric people in the Olympic Dam area were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
-The implication of this finding is that even in relatively homogenous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites (cf.
+The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Olympic Dam area were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
+The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites (cf.
 Attwell and Fletcher 1987).
 These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007, Kowalewski 2008: 233).
-Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with simple behavioural ecological models.
+Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with relatively simple behavioral ecological models.
+We hope that this data-intensive approach to using gradients of changes in archaeological site attributes across the landscape might suggest a new direction for investigating past human behaviour at ecological interfaces, avoiding the problematic assumptions of previous work on edge effects.
 
@@
 This work was undertaken under a contract with BHP Billiton that includes research support.
-Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301) and a grant from the Herbert H.
-Gowen International Studies Endowment (UW Office of Global Affairs).
+Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301), a grant from the Herbert H.
+Gowen International Studies Endowment (UW Office of Global Affairs), and an Australian Research Council Future Fellowship (FT140100101).
 
@@
 Given, O. and S. Hyla (2006). "Forging Partnerships in Laos." Expedition 48(1): 38-39.
+
 Glassow, M. A. and M. Leone (1972 ). Changes in the Adaptations of Southwestern Basketmakers: a Systems Perspective. Contemporary Archaeology. M. Leone, Southern Illinois University Press: 289-302.
@@
 The current commit message is "`r last_commit@summary`".
-
-```{r eval = FALSE}
-
-###########################################################
-# checking P&M's edits from 14 May 2012
-#
-library(foreign)
-ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
-# sites per TP
-sites.per.tp <- data.frame(table(na.omit(ODX_allsites_2010$GREGLANDF)))
-# find proportion of all sites in each TP
-sites.per.tp$prop <- sites.per.tp$Freq/sum(sites.per.tp$Freq)*100
-#
-# get counts from density squares
-ODX_allsites_2010$DENCOUNT <- with(ODX_allsites_2010, (as.numeric(levels(DENSQ1)[DENSQ1]))+(as.numeric(levels(DENSQ2)[DENSQ2]))+(as.numeric(levels(DENSQ3)[DENSQ3]))+(as.numeric(levels(DENSQ4)[DENSQ4]))+(as.numeric(levels(DENSQ5)[DENSQ5]))+(as.numeric(levels(DENSQ6)[DENSQ6]))+(as.numeric(levels(DENSQ7)[DENSQ7]))+(as.numeric(levels(DENSQ8)[DENSQ8])+(as.numeric(levels(DENSQ9)[DENSQ9]))/9)*((as.numeric(levels(DIMSIT1)[DIMSIT1]))*(as.numeric(levels(DIMSIT2)[DIMSIT2]))))
-# check if all sites with density square data were processed
-# should return TRUE
-# identical(length(na.omit(ODX_allsites_2010$DENSQ1)), 
-# length(na.omit(ODX_allsites_2010$DENCOUNT)))
-# # check how many sites with density square data
-# length(na.omit(ODX_allsites_2010$DENCOUNT))
-# #
-# get counts when all artefacts counted
-ODX_allsites_2010$COUNTONLY <- with( ODX_allsites_2010, as.numeric(levels(TOTART)[TOTART]) )
-# check how many are ok
-# length(na.omit(ODX_allsites_2010$TOTART)) # should have
-# length(na.omit(ODX_allsites_2010$COUNTONLY)) # did get
-#
-# get counts from when only a VISTEST was taken:
-ODX_allsites_2010$VISCOUNT<-with(ODX_allsites_2010, (as.numeric(levels(VISEST)[VISEST]))*((as.numeric(levels(DIMSIT1)[DIMSIT1]))*(as.numeric(levels(DIMSIT2)[DIMSIT2]))))
-# check how many are ok
-# length(na.omit(ODX_allsites_2010$VISEST)) # should have
-# length(na.omit(ODX_allsites_2010$VISCOUNT)) # did get
-#
-ODX_allsites_2010$artefacts <- as.numeric(with(ODX_allsites_2010, 
-      ifelse(COUNTONLY > 1, COUNTONLY,
-      ifelse(VISCOUNT  > 1 ,  VISCOUNT, 
-      ifelse(DENCOUNT  > 1 , DENCOUNT, "XX")))))
-#                                          
-# have a bit of a look 
-# ODX_allsites_2010[ODX_allsites_2010$VISCOUNT  != 0 ,]
-# ODX_allsites_2010[ODX_allsites_2010$DENCOUNT  != 0 ,]
-# ODX_allsites_2010[ODX_allsites_2010$COUNTONLY != 0 ,]
-#
-# summary(ODX_allsites_2010$VISCOUNT)
-# summary(ODX_allsites_2010$DENCOUNT)
-# summary(ODX_allsites_2010$COUNTONLY)
-#  
-# check how many are ok
-# length(na.omit(ODX_allsites_2010$artefacts))
-# 
-# check how many sites in total
-# summary( ODX_allsites_2010[,1:10] )
-# 10636 sites in data, 133 are hearths so 10503 should 
-# have artefact data... in fact 9859 do, so some 644 sites
-# have problems with artefact data, that's only 0.6% so that's
-# not too bad.
-#
-# get total number of artefacts in each TP
-artefacts.per.tp <- aggregate(data = ODX_allsites_2010, artefacts ~ GREGLANDF, FUN = sum)
-#
-# check output
-identical(
-  sum(na.omit(ODX_allsites_2010$artefacts)),
-  sum(artefacts.per.tp$artefacts))
-#
-# how many artefacts missing from summary?
-sum(na.omit(ODX_allsites_2010$artefacts)) - sum(artefacts.per.tp$artefacts)
-# 5926 artefacts missing... why are they missing? not sure
-#
-#
-library(foreign)
-ODX_TPs_2010 <- read.dbf("../data/CLipped_ODO_terrain_patterns_2010_Ja.dbf")
-# 
-# Calculate the areas of the sampling polygons in ArcMap:
-# 1.  Open the attribute table of the layer you want to edit. 
-# 2.	Right-click the field heading for area (if there is no field for area values, you can add a new field for area by clicking the Options button and selecting the new field option).
-This must be formatted as a Double not a short integer.
-# 3.	Click Calculate Values. 
-# 4.	Checkmark the Advanced box (this is important - don't forget this!) 
-# 5.	Type the following VBA statement in the first text box (you should be able to copy and paste): 
-# Dim dblArea as double 
-# Dim pArea as IArea 
-# Set pArea = [shape] 
-# dblArea = pArea.area 
-# 6.	Type the variable  dblArea =in the second text box (near the bottom of the dialog box) directly under the area field name. 
-#
-# check areas and proportion of total areas for each TP
-library(data.table)
-ODX_TPs_2010.dt <- data.table(ODX_TPs_2010)
-setkey(ODX_TPs_2010.dt,GREGLANDF)
-ODX_TP_areas <- ODX_TPs_2010.dt[,sum(Clip_Area_), by = GREGLANDF]
-ODX_TP_areas$prop <- (ODX_TP_areas$V1/sum(ODX_TP_areas$V1))*100
-# check that proportions add to 100%
-sum(ODX_TP_areas$prop)
-# check of total area of TPs
-sum(ODX_TP_areas$V1)/(1000*1000)
-#
-# repeat checks for landforms
-ODX_LF_areas <- ODX_TPs_2010.dt[,sum(Clip_Area_), by = LANDFORM]
-ODX_LF_areas$prop <- (ODX_LF_areas$V1/sum(ODX_LF_areas$V1))*100
-# check that proportions add to 100%
-sum(ODX_TP_areas$prop)
-#
-# Plot of TP areas (figure 4 of JFA paper)
-
-library(ggplot2)
-ggplot(ODX_TP_areas, aes(reorder(GREGLANDF, prop), prop)) + 
-  geom_bar(stat = "identity",
-           fill = "white", colour = "black") +
-  theme_bw(base_size = 17) +
-  theme(axis.text.x = element_text(size = 17),
-        axis.text.y = element_text(size = 17),
-        axis.title.x = element_text(size = 17),
-        axis.title.y = element_text(size = 17, angle = 90)) + 
-  ylab("Percentage of study area") +
-  xlab("Terrain pattern") 
-#
-# check of sites per sq km for each TP
-sites.per.sqkm <- merge(sites.per.tp, ODX_TP_areas, by.x = "Var1", by.y = "GREGLANDF")
-sites.per.sqkm$sites.per <- with(sites.per.sqkm, Freq/V1*1000*1000)
-#
-# check of artefacts per sq km for each TP
-artefacts.per.sqkm <- merge(artefacts.per.tp, ODX_TP_areas)
-artefacts.per.sqkm$artefacts.per <- with(artefacts.per.sqkm, artefacts/V1*1000*1000)
-
-# For looking a distance values from features in ArcMap
-# First, polygon features have to be converted to line features.data management tools, features, feature to line.
-# 1.  Open spatial analyst extension
-# 2.	Define extent (LTA_areas is a good one)
-# 3.	Spatial analyst - distance- straight line
-# 4.	Then go to ArcToolbox - Spatial analyst tools - extraction 
-# 5.	'Extract values to points'
-# 6. Then arctoolbox, spatial analysis tools, extraction, extract values to points. point layer is sites, values come from raster, then RASTERVALU is distance field. 
-# 7 .Then join to TP polygon to get TP values in sites dbf.
-
-library(foreign)
-ODX_allsites_2010 <- read.dbf("../data/Join_Extract_ArcSite.dbf")
-ODX_allsites_2010 <- ODX_allsites_2010[ !is.na(ODX_allsites_2010$GREGLANDF), ]
-ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
-ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
-# order the levels of GREGLANDF for plotting
-# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
-levels(ODX_allsites_2010.s$GREGLANDF)
-ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
-  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])
-
-# plot
-library(ggplot2)
-ggplot(ODX_allsites_2010.s, aes(GREGLANDF, RASTERVALU)) + 
-  geom_boxplot() + 
-  scale_y_log10(breaks = seq(0,1400,200)) + 
-  theme_bw(base_size = 17) +
-  theme(axis.text.x =  element_text(size = 17),
-        axis.text.y =  element_text(size = 12),
-        axis.title.x = element_text(size = 17),
-        axis.title.y = element_text(size = 17, 
-                                    angle = 90))  +
-  xlab("Terrain pattern") +
-  ylab("Distance of site to boundary (m)")
-
-# anova to see if sig diff between distances...
-fit <- aov(data = ODX_allsites_2010.s, RASTERVALU ~ GREGLANDF)
-plot(TukeyHSD(fit))
-
-# plot anova
-# install.packages("granovaGG", dependencies = TRUE)
-library("granovaGG")
-granovagg.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)
-# install.packages("granova")
-library("granova")
-granova.1w(data = ODX_allsites_2010.s$RASTERVALU, group = ODX_allsites_2010.s$GREGLANDF)
-
-# start with OUTER buffers
-bd_files <- list.files("../data/outer_buffers", full.names = TRUE)
-# list in, dataframe out
-library(plyr)
-library(foreign)
-bd.data <- ldply(bd_files, read.dbf)
-bd.data$SiteDens <- with(bd.data, Count_/buf_area * (1000*1000))
-# exclude buffers with less than n sites since they 
-# skew the density calculations 
-n = 15
-bd.data <- bd.data[bd.data$Count_ > n, ]
-unique(bd.data$BUFF_DIST)
-bd.data$Buf <- as.factor(with(bd.data, ifelse(BUFF_DIST == 100, '000-100', ifelse(BUFF_DIST == 200, '100-200', ifelse(BUFF_DIST == 300, '200-300', ifelse(BUFF_DIST == 400, '300-400', ifelse(BUFF_DIST == 500, '400-500','NA')))))))
-
-#  how many sites in each set?
-sum(na.omit(subset(bd.data, GREGLANDF=='Czs4')$Count_))
-sum(na.omit(subset(bd.data, GREGLANDF=='Czs5')$Count_))
-
-# now check the boundary densities for INNER buffers
-inbd.files <- list.files("../data/inner_buffers", full.names = TRUE)
-# list in, dataframe out
-library(plyr)
-library(foreign)
-inbd.data <- ldply(inbd.files, read.dbf)
-inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
-inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
-unique(inbd.data$BUFF_DIST)
-inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))
-
-#  how many sites in each set?
-sum(as.numeric((subset(inbd.data[as.numeric(inbd.data$Join_Count),], GREGLANDF=='Czs4'))$Join_Count))
-sum(as.numeric((subset(inbd.data[as.numeric(inbd.data$Join_Count),], GREGLANDF=='Czs5'))$Join_Count))
-
-# combine inner with outer buffers
-inner <- subset(inbd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')[,c('GREGLANDF','SiteDens','Buf')]
-outer <- subset(bd.data, GREGLANDF =='Czs5' | GREGLANDF =='Czs4')  [,c('GREGLANDF','SiteDens','Buf')]
-# colnames(outer)[2] <-  'mean_SiteDens'
-inner_outer <- na.omit(rbind(inner, outer))
-# put the boundary distances in order
-levels(inner_outer$Buf)
-inner_outer$Buf <- factor(inner_outer$Buf, 
-  levels(inner_outer$Buf)[c(5,4,3,2,1,6,7,8,9,10)])
-# remove rows with zero for SiteDens
-inner_outer <- inner_outer[inner_outer[, 'SiteDens']!=0, ]  
-# plot inner and outer together
-library(ggplot2)
-library(grid)
-library(scales)
-ggplot(inner_outer,  
-       aes(Buf, 
-           SiteDens)) + 
-  geom_point(size = 2) + 
-  facet_grid(GREGLANDF ~ .) + 
-  scale_y_log10() + 
-  geom_smooth(aes(group = 1), 
-              fill = "grey80", 
-              colour = "grey") +
-  theme_bw() + 
-  theme(
-  axis.title.x = element_text(vjust = -0.5, 
-                              size = 15),
-  axis.title.y=  element_text(size = 15, 
-                              angle=90)) + 
-  xlab("Distance from boundary in meters \n(negative values are interior distances, \npositive values are exterior distances)") + 
-  ylab("Sites per sq km")
-
-# significance tesing for changes in site density over buffers
-# classic ANOVA...
-summary(fit <- aov(data = inner_outer[inner_outer$GREGLANDF == 'Czs5' ,], SiteDens ~ Buf))
-# yes for Csz5, no for Csz4
-# inspect to see if assumptions are ok
-qqnorm(resid(fit), main="Normal Q-Q Plot")
-qqline(resid(fit), col="red")
-# a bit wacky...
-
-# two-way ANOVA
-inner_outer$Buf <- as.character(inner_outer$Buf)
-two_way_anova_data <- inner_outer[inner_outer$GREGLANDF %in% c('Czs5','Czs4') &
-                                      !inner_outer$Buf %in% c("-100-000", "000-100"), ]
-summary(fit <- aov(data = two_way_anova_data, 
-                   SiteDens ~  GREGLANDF * Buf))
-
-# try permutation ANOVA...
-# install.packages("lmPerm")
-# library(lmPerm)
-# summary(fitp <- aovp(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, perm="Exact"))
-# yes for Csz5, no for Csz4
-
-library(coin) # library needed for more permutation tests
-# We'll use Monte Carlo simulations to get the p value
-oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
-oneway_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, distribution=approximate(B=10000))
-# this is a test of whether 
-# two or more samples from normal distributions have the 
-# same means, variances not assumed equal
-kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs5',], SiteDens ~ Buf, distribution=approximate(B=10000))
-kruskal_test(data = inner_outer[inner_outer$GREGLANDF == 'Czs4',], SiteDens ~ Buf, distribution=approximate(B=10000))
-# yes for Csz5, no for Csz4
-
-# In each of the occurrences of those TPs create buffer zones A (1m to -100m), B (-100m to -200m), C (-200m to -300m).  Then work out the density of sites in each buffer unit and in the whole TP.
-# 
-# Than plot the ratios for each Czs4 and Czs5 Terrain Pattern of:
-# site densities in A/site densities in whole of each TP
-# site densities in B/site densities in whole of each TP
-# site densities in C/site densities in whole of each TP
-
-inner # site densities per buffer area
-sites.per.sqkm # densities per TP
-
-# Czs5
-# site densities in 0-100 m inner buffer divided by total density of TP
-Czs5.inner.100 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-100-000",])
-Czs5.inner.100.ratio <- mean(Czs5.inner.100$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs5",])$sites.per
-# site densities in 100-200 m inner buffer divided by total density of TP
-Czs5.inner.200 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-200-100",])
-Czs5.inner.200.ratio <- mean(Czs5.inner.200$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs5",])$sites.per
-# site densities in 200-300 m inner buffer divided by total density of TP
-Czs5.inner.300 <- with(inner, inner[GREGLANDF == "Czs5" & Buf == "-300-200",])
-Czs5.inner.300.ratio <- mean(Czs5.inner.300$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs5",])$sites.per
-
-# Czs4
-# site densities in 0-100 m inner buffer divided by total density of TP
-Czs4.inner.100 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-100-000",])
-Czs4.inner.100.ratio <- mean(Czs4.inner.100$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs4",])$sites.per
-# site densities in 100-200 m inner buffer divided by total density of TP
-Czs4.inner.200 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-200-100",])
-Czs4.inner.200.ratio <- mean(Czs4.inner.200$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs4",])$sites.per
-# site densities in 200-300 m inner buffer divided by total density of TP
-Czs4.inner.300 <- with(inner, inner[GREGLANDF == "Czs4" & Buf == "-300-200",])
-Czs4.inner.300.ratio <- mean(Czs4.inner.300$SiteDens)/with(sites.per.sqkm, sites.per.sqkm[Var1 == "Czs4",])$sites.per
-
-# combine both for plotting
-Czs4 <- c(Czs4.inner.100.ratio, Czs4.inner.200.ratio, Czs4.inner.300.ratio)
-Czs5 <- c(Czs5.inner.100.ratio, Czs5.inner.200.ratio, Czs5.inner.300.ratio)
-bound <- data.frame(cbind(Czs4, Czs5))
-bound$dist <- c("0-100", "100-200", "200-300")
-library(reshape2)
-bound.m <- melt(bound)
-
-# plot
-library(ggplot2)
-ggplot(bound.m, 
-       aes(dist,
-           value, 
-           colour = variable)) + 
-  geom_line(aes(group = variable))
-
-# convert to z-scores....
-Czs4.sc <-  scale(Czs4)
-Czs5.sc <-  scale(Czs5)
-bound.sc <- data.frame(cbind(Czs4.sc, Czs5.sc))
-bound.sc$dist <- c("0-100", "100-200", "200-300")
-colnames(bound.sc) <- c("Czs4", "Czs5", "distance")
-library(reshape2)
-bound.sc.m <- melt(bound.sc)
-colnames(bound.sc.m) <- c("distance", "Terrain Pattern", "density of sites")
-
-# plot scales
-library(ggplot2)
-ggplot(bound.sc.m, 
-       aes(x=bound.sc.m[,1], 
-           y=bound.sc.m[,3])) + 
-  geom_line(aes(group = bound.sc.m[,2], 
-                linetype = bound.sc.m[,2]), 
-            size = 1)  +
-  theme_bw() +
-  theme(
-  axis.text.x =  element_text(size = 17), 
-  axis.text.y =  element_text(size = 17),
-  axis.title.x = element_text(size = 17),
-  axis.title.y = element_text(size = 17,
-                              angle = 90),
-  legend.text =  element_text(size=14),
-  legend.title = element_text(size=14)) + 
-  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
-  ylab("z-score of site density in buffer area") +
-  scale_linetype_manual(name="Terrain Pattern", values = c(1,2) )
-
-# dot-plot
-library(viridis)
-ggplot(bound.sc.m, 
-       aes(x=distance, 
-           y=`density of sites`,
-           shape = `Terrain Pattern`,
-           colour = `Terrain Pattern`)) + 
-  geom_point(size = 5)  +
-  geom_line(aes(group = `Terrain Pattern`)) +
-  theme_bw(base_size = text_size) +
-  coord_flip() +
-  scale_color_viridis(discrete = TRUE) +
-  theme(axis.title.y = element_text(angle = 90)) + 
-  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
-  ylab("z-score of site density in buffer area")  +
-  labs(shape="Terrain\nPattern", colour="Terrain\nPattern")
-
-# checking benefit of mobile GIS
-dat <- read.table(text = "
-Year Pre.2007	2007	2008	2009
-Area.surveyed.(km2)	35	95	275	145
-Rate.of.survey.(km2/person/day)	0	0.3	0.3	0.3
-Total.no.persondays	0 317	917	483
-Approximate.number.of.sites.recorded	665	3900	6600	5500
-No.sites.recorded/person/day	 3.5	12.3	7.2	11.3
-Frequency.of.occurrence.of.sites/km2	19	41	24	38", row.names = 1, header = TRUE)
-
-dat_t <- data.frame(t(dat))
-# quick look
-pairs(dat_t)
-
-# correlation of sites recorded by density of sites
-cor.test(dat_t[,5], dat_t[,6], method = 'pearson')
-# linear model for sites recorded by density of sites
-summary(lm(dat_t[,5] ~ dat_t[,6]))
-
-# subset 2007-2009 data
-dat_t_s <- dat_t[2:4,]
-# linear model for sites recorded by density of sites, 2007-2009
-summary(lm(dat_t_s[,5] ~ dat_t_s[,6]))
-
-# predict given 19 sites/km2
-19 * 0.297  + 0.043
-
-```
+The repository is online at <https://github.com/benmarwick/olympicdamboundaries>

### 2017-01-26 R/hello.R

@@ -1,6 +1,6 @@
 # Hello, world!
 #
-# This is an example function named 'hello' 
+# This is an example function named 'hello'
 # which prints 'Hello, world!'.
 #
 # You can learn more about package authoring with RStudio at:
@@ -16,3 +16,8 @@
 hello <- function() {
   print("Hello, world!")
 }
+
+
+#  ------------------------------------------------------------------------
+
+

### 2017-01-27 analysis/paper/paper.Rmd
update DOI

@@
 
-To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/xxx/xxx>.
+To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/10.17605/OSF.IO/RMKGE>.
 Also in this version-controlled compendium are the raw data for all the tests reported here, as well as additional regression diagnostics and power tests.

### 2017-01-30 analysis/paper/paper.Rmd
send to PH, MS and PH for comment

@@
 We use data from surface scatters of stone artefacts collected during pedestrian survey to compare behaviors at landscape boundaries to behaviours at the centers of landforms.
-We propose a model of human use of the landscape that predicts that the prehistoric occupants of the study are were sensitive to the different economic potential of dissimilar landscapes.
+We propose a model of human use of the landscape that predicts the prehistoric occupants of the study are were sensitive to the different economic potential of subtly dissimilar landscapes.
 In evaluating the model we find that there are different densities of archaeological sites in each landscape type.
@@
                message = FALSE,
-               dpi = 600)
+               dpi = 900)
 options(warn = -1)
@@
 Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972).
-This increased variety and density of plants and animals, often described as an ‘edge effect’, was claimed by Harris et al.
+This increased variety and density of plants and animals, often described as an ‘edge effect’, has been a focus of archaeological research.
+For example, it was claimed by Harris et al.
 (1969) to be a causal factor in the emergence of agriculture.
@@
 
-These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists have been overly simplistic in their use of many ecological concepts (Rhoades 1978, Hardesty 1980, King and Graham 1981).
-Problems with these previous approaches include ignoring controversies among biologists about the definition of ecotones, and unjustified assumptions about the uniformity of all ecotones, and about edge effects always increasing the amount of food species available to humans.
-Here we pursue a different approach by focusing on subtle variations in landscape characteristics within a biome, rather than between them.
+These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists had been overly simplistic in their use of many ecological concepts (Rhoades 1978, Hardesty 1980, King and Graham 1981).
+Problems with these previous approaches include ignoring controversies among biologists about the definition of ecotones, unjustified assumptions about the uniformity of all ecotones, and unverified assumptions about edge effects always increasing the amount of food species available to humans.
+Alternative approaches to investigating landform boundary effects can be found in the work of Kimes et al.
+(1982), who used distribution of artefacts to see how culture group areas related to landform boundaries.
+However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these are often easier to measure and interpret than landscape boundaries (Daniel 2001; Jones, Beck et al.
+2003).
+Here we pursue a new approach to understanding boundaries by focusing on subtle variations in landscape characteristics within a biome, rather than between them.
 By focusing on continuous variation, rather than step-wise shifts in the archaeological record across ecological interfaces, we introduce a new approach to studying behaviors of prehistoric forager groups at interface zones.
 
-In this study we ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of where to carry out activities in respect to environmental boundaries.
-We predict that prehistoric foragers were sensitive to differences in resource structure as gradients across the landscape.
-This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability that make comparative landscape analysis more tractable in many contexts.
-The spatial scales of much of the previous work on prehistoric arid zone foragers has ranged from continental (Williams, Ulm et al.
+In this study we ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of the optimum locations for their activities in respect to environmental boundaries.
+We predict that prehistoric foragers were sensitive to subtle differences in resource structure as gradients across the landscape.
+This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability.
+The step-wise approaches are useful because they make comparative landscape analysis more tractable in many contexts and at different scales.
+For example, the spatial scales of much previous work on prehistoric arid zone foragers has ranged from continental (Williams, Ulm et al.
 2013) to ecological regions (Holdaway, Douglass et al.
@@
 
-Our work is motivated by previous archaeological work on cultural landscapes of foragers that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982, Veth 1993, Napton and Greathouse 1996, Holdaway 1998, Barton 2003, Smith 2006, Fanning, Holdaway et al.
+Our work is motivated by previous archaeological work on cultural landscapes of foragers in Australian arid and semi-arid zones that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982, Veth 1993, Napton and Greathouse 1996, Holdaway 1998, Barton 2003, Smith 2006, Fanning, Holdaway et al.
 2008, Smith, Williams et al.
@@
 2015).
-The archaeological signature of landuse typically has been the distribution and composition of surface stone artefact scatters, which are by far the most abundant type of site.
-In this context, sites are typically small, dense and discrete concentrations of stone artefacts that represent temporary, short-term activity areas.
+The archaeological signature of landuse on these landscapes has typically been the distribution and composition of surface stone artefact scatters, which are by far the most abundant type of site.
+Sites in these landscapes are typically small, dense and discrete concentrations of stone artefacts that represent temporary, short-term activity areas.
 The spatially discrete quality of the majority of sites in these desert regions, such as our study area, means it is possible to distinguish between even quite close sites (<50 m apart) with confidence, unlike many other contexts where artifact distributions are more spatially continuous (Thomas 1975, Foley 1981, Dunnell and Dancey 1983, Kuna 2000).
@@
 The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes, making them convenient for spatial analysis of archaeological features.
-Landform boundary effects have previously been employed in studying the distribution of artefacts to infer culture group areas (Kimes, Haselgrove et al.
-1982).
-However, archaeological discussions of settlement patterns have typically focused on straight line distances from key resources such as raw materials for stone artefacts, as these are often easier to measure and interpret than landscape boundaries (Daniel 2001, Jones, Beck et al.
-2003).
 With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.
@@
 A pair of dated sites are insufficient to draw conclusions beyond the local level, however the distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact scatters recorded during the survey (Hughes and Hiscock 2005).
-A study of 78 OSL ages from geological contexts in the study area has shown that the mid to late Holocene period was also a time of broad-scale dune stability (Hughes et al.
+A study of 78 OSL ages from geological contexts in the study area has shown that the mid to late Holocene period was a time of broad-scale dune stability (Hughes et al.
 2014).
@@
 
-Two approaches were used to diagnose the presence and character of boundary effects.
+```{r methodschematic, fig.cap="Schematic of two approaches used to investigate boundary effects at Olympic Dam, South Australia.
+A: the straight-line distances from sites to the boundaries of the terrain patterns that they are contained it was measured.
+B: the number of sites per square kilometer was measured in a series of buffer areas, an example of a buffer area is indicated here by a grey region at the exterior of the terrain pattern area.", fig.width=fig_width_two_cols}
+
+knitr::include_graphics("../figures/method_schematic.png")
+
+```
+
+We used two approaches to investigate the presence and character of boundary effects on the Olympic Dam landscape (Figure \@ref(fig:methodschematic)).
 In the first approach, we determined the straight-line distance from every site to the nearest boundary of the terrain pattern in which it was contained.
 Descriptive statistics for these distances allows testing of the prediction that sites tend to be closer to the edge of some landforms than other landforms.
-In the second approach, we measured the density of sites at 100 m intervals, or buffer zones, inwards from the boundary of the terrain patterns Csz4 and Csz5, as the best represented terrain patterns.
+
+In the second approach, we measured the density of sites per square kilometer at 100 m intervals, or buffer zones, inwards from the boundary of the terrain patterns Csz4 and Csz5, as the best represented terrain patterns.
 These data were used to test the prediction that site densities tends to be higher towards the center of Csz4, as a resource-rich landform, compared to Csz5, as a resource-poor landform.
@@
 The density of sites in each buffer zone was then divided by the average density for the entire terrain pattern to standardize the density value for the terrain pattern.
-
-Finally, z-scores were calculated to put the differences in site density on a comparable scale between the two terrain patterns.
-Z-scores are a transformation of the data that show how many standard deviations the site density in each buffer polygon varies from the mean density (z = 0) of the buffers in each terrain pattern.
+Z-scores were calculated to put the differences in site density on a comparable scale between the two terrain patterns.
+Z-scores are a transformation of the data that show how many standard deviations the site density in each buffer polygon varies from the mean density (z = 0) of all the buffers in each terrain pattern.
 This second approach provided a way to minimize the effect of differences in the sizes of landscape units being compared.
@@
 To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/10.17605/OSF.IO/RMKGE>.
-Also in this version-controlled compendium are the raw data for all the tests reported here, as well as additional regression diagnostics and power tests.
+Also in this version-controlled compendium are the raw data for all the tests and visualizations reported here, as well as additional regression diagnostics.
 All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository.
@@
 2011).
-The first prediction tested here is that there was differential use of localities within dunefields.
-More specifically, areas of widely-spaced longitudinal sand dunes were predicted to have been intensely occupied when compared with areas containing more closely-spaced dunes.
-This difference is anticipated because of the greater range and density of animal foods and water sources in places where dunes are more widely separated.
+The first prediction tested here is that there was differential use of terrain pattern areas within dunefields.
+More specifically, we predicted that areas of widely-spaced longitudinal sand dunes were intensely occupied, compared with areas containing more closely-spaced dunes.
+This difference is anticipated because of the greater range and density of animal foods, stone for making artefacts, and water sources in places where dunes are more widely separated.
 
@@
 ## plot site densities and artefact densities per TP
+library(ggforce)
 g <- ggplot(ODX_allsites_2010_per_TP_NO,
@@
   geom_boxplot() +
+  geom_sina(alpha = 0.03) +
   theme_bw(base_size = text_size) +
@@
 
-Figure \@ref(fig:distanceofsitetoboundary) depicts median distances for site distances from terrain pattern boundaries where the dunefields are on average more than 1 km across (cf.
+Figure \@ref(fig:distanceofsitetoboundary) depicts distances of sites from the boundaries of the terrain pattern that contains those sites.
+The sample of sites in this figure is limited to terrain pattern areas that are on average more than 1 km across (cf.
 Figure \@ref(fig:mapofstudyarea1)).
-The general pattern is clear, with distances being substantially further in widely spaced dunefields such landform type 4 (e.g.
+The general pattern is clear, with distances to the boundary being substantially further in widely spaced dunefields such landform type 4 (e.g.
 A4, Czs4 and Q4) than in more closely spaced dunefields such as landform types 5 and 6 (e.g.
 Czs5, Q5and Q5).
+This means that sites tend to be nearer to the centers of landform type 4, but nearer to the boundaries of landform type 5.
 Visual inspection of Figure \@ref(fig:mapofstudyarea1) shows that this relationship is not dependent simply on the size or shape of the terrain patterns.
@@
 The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding the 0-100 m zone).
+
 When we use z-scores to standardize for differences in site density between two terrain patterns, we see that the pattern is even more stark.
@@
 In our measurements of site location and site densities, we observed that subtle differences in physical features of the landscape had significant effects on where people were active on the study area during the Holocene.
-Testing of our model's predictions were facilitated by two characteristics that are distinctive to this project.
+Tests of our model's predictions were facilitated by two characteristics that are distinctive to this project.
 One is that the extensive survey area, the intensive level of survey and the resulting large sample size of archaeological sites, meant that observation of human choices and behaviour could be obtained at a very high spatial resolution.
-The second is that objective and robust environmental classifications, creating well-defined terrain patterns, made it possible to quantify landscape diversity in the study area at a high spatial resolution and reduced false associations that could result from less accurate environmental mapping.
+The second is that objective and robust environmental classifications, creating well-defined terrain patterns, made it possible to quantify landscape diversity in the study area at a high spatial resolution and reduce false associations that could result from less accurate environmental mapping.
 
@@
 We found that sites were generally located further from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
-This indicates that people preferred to be located close to the center of areas with widely-spared dunes, and on the edges of the areas with closely spaced dunes.
+This indicates that people preferred to be located close to the center of areas with widely-spared dunes, but only on the edges of the areas with closely spaced dunes.
 Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.
@@
 This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns.
-This claim is axiomatic to human behavioral ecology in which different landscapes are commonly compared as homogeneous units to improve analytical tractability.
 The novel element of the data presented here is that the distribution of human activities varies substantially and predictably within as well as between terrain patterns.
-It is also noteworthy that a relatively subtle geographic differences between landscapes – the different spacing of sand dunes between Czs4 and Czs5 – has a marked visible effect on the distribution of archaeological sites.
-We speculate that this is due to the relationship between dune spacing and the availability of flakable stone.
+It is also noteworthy that relatively subtle geographic differences between landscapes – the different spacing of sand dunes between Czs4 and Czs5 – has a marked effect on the distribution of archaeological sites.
+We speculate that this is due to the relationship between dune spacing and the availability of flakable stone, as well as other resources.
 The causes of variation in dune spacing are poorly understood.
@@
 
-Our survey activity in the Olympic Dam region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
+In addition to these observations about landscape use, our survey activity in the Olympic Dam region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
 Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM).
@@
 mobile_gis_efficiency_lm_p_value <- round(mobile_gis_efficiency_lm$coefficients[2, 4]  , 3)
+mobile_gis_efficiency_lm_df <- round(mobile_gis_efficiency_lm$df[2] , 3)
+mobile_gis_efficiency_lm_F <- round(mobile_gis_efficiency_lm$fstatistic[1] , 3)
 
@@
 
-While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS because the rate of recording is strongly predicted by the density of sites (R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`).
-To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data , which resulted in a rate of site recording that can by predicted by site density * `r mobile_gis_efficiency_lm_2007_2009_est` (`r mobile_gis_efficiency_lm_2007_2009_est_error`) + `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) (standard errors in parentheses).
-This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate given the density of 19 sites/km^2^ observed in the pre-2007 data.
+While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS.
+This is because the rate of recording is strongly predicted by the density of sites (F(1, `r mobile_gis_efficiency_lm_df`) = `r mobile_gis_efficiency_lm_F`, R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`).
+To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) + `r mobile_gis_efficiency_lm_2007_2009_est`(site density) (`r mobile_gis_efficiency_lm_2007_2009_est_error`) +  (standard errors in parentheses).
+This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate using a mobile GIS given the density of 19 sites/km^2^ observed in the pre-2007 data.
 The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^.
@@
 This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.
-Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, a highly structured data collection form could be used, and the start-up and training costs could be amortized over multiple years.
+Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.
 
@@
 The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Olympic Dam area were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
-The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites (cf.
+The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites the reveal how people adapted to the landscape (cf.
 Attwell and Fletcher 1987).

### 2017-02-01 analysis/paper/paper.Rmd
update acknowledgements

@@
 
-This work was undertaken under a contract with BHP Billiton that includes research support.
+The fieldwork component of this study was undertaken in accordance with the Olympic Dam Agreement between BHP Billiton and the Aboriginal native title claimant groups under a contract with BHP Billiton that included research support.
 Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301), a grant from the Herbert H.

### 2017-02-01 analysis/paper/SOM.Rmd
edits from PH, PH and MS to adapt the paper for the current NT conditions.

@@
 Marjorie Sullivan    
-The Australian National University, Canberra ACT 0200, Australia and 
-Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
+The University of Sydney, Sydney NSW, Australia
 
@@
 
-The decision to use a mobile GIS was based on four specific advantages over paper-based systems of the numerous benefits listed by Tripcevich (2004a; 2004b), Clarke (2004) and Wagtendonk and De Jeu (2007):
+The decision to use a mobile GIS was based on four specific advantages over paper-based systems of the numerous benefits listed by Tripcevich (2004a; 2004b), Clarke (2004) and Wagtendonk and De Jeu (2007), and previously summarised for this project by Sullivan et al.
+(2014):
 
@@
 Over 95% of the survey was covered by good quality recent air photography.
-Some of the thematic maps had been digitized, other data were able to be digitized and stored as shapefiles by the mining company, BHP Billiton.
+Some of the thematic maps had been digitized, other data were able to be digitized and stored as shapefiles by the mining company's land managers.
 While paper maps can provide this information, and may be easier to read than a small screen in bright sun, a mobile GIS gives more immediate context awareness (Pascoe et al.
@@
 The pattern they document is a three-fold increase in the rate of site recording with mobile GIS.
-The same general pattern of time saving was observed in our history of research at Olympic Dam.
+The same general pattern of time saving was observed in our history of research in the Roxby dunefield.
 Our pre-2007 data was transferred from field notes to a computer in 8-10 days after the survey concluded, while our 2007-2009 data was transferred from the mobile GIS to the desktop computer in less than five minutes per fieldwork day.
 
-The Olympic Dam sites fall into a narrow range of site types and stone artefacts within them are patterned in highly repeated ways that suited a systematic recording protocol.
+The Roxby dunefield sites fall into a narrow range of site types and stone artefacts within them are patterned in highly repeated ways that suited a systematic recording protocol.
 An important consideration in evaluating this use of mobile GIS and traditional methods is the regularity of the archaeological materials that were recorded.
@@
 This was a more serious concern in the early implementation of mobile GIS and more recent equipment and software are mostly reliable and robust.
-Breakdowns can never be completely avoided however and during the Olympic Dam survey their impacts were mitigated with multiple nightly data backups and keeping redundant functioning equipment on hand for immediate replacement of broken units, to minimize delays caused by waiting for repairs.
+Breakdowns can never be completely avoided however and during the field survey their impacts were mitigated with multiple nightly data backups and keeping redundant functioning equipment on hand for immediate replacement of broken units, to minimize delays caused by waiting for repairs.
 When the sites database was 'cleaned' at the end of the survey stage, serious errors (ie.
@@
 
+Sullivan, M., P.
+Hiscock, and P.
+Hughes (2014) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia.
+Journal of the Anthropological Society of South Australia 38: 85-107.
+
 Tripcevich N 2004a. Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey. Cartography and Geographic Information Science 31(3), 137-151.

### 2017-02-01 analysis/paper/paper.Rmd
edits from PH, PH and MS to adapt the paper for the current NT conditions.

--- 2017-01-30 analysis/paper/paper.Rmd
@@
 abstract: |
-  Landscapes throughout a region vary in the resources they contain.
+  Landscapes throughout any region vary in the resources they contain.
 We investigate how Holocene forager populations adapted to this variation in the linear sand dune desert of arid South Australia.
-We use data from surface scatters of stone artefacts collected during pedestrian survey to compare behaviors at landscape boundaries to behaviours at the centers of landforms.
+We use data from surface scatters of stone artefacts collected during pedestrian survey to compare behaviors at landform boundaries to behaviours at the centers of landforms.
 We propose a model of human use of the landscape that predicts the prehistoric occupants of the study are were sensitive to the different economic potential of subtly dissimilar landscapes.
@@
 Philip Hughes  
+Flinders University, Adelaide, South Australia, Australia
 Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
@@
 Marjorie Sullivan   
+The University of Sydney, Sydney NSW, Australia
 Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
@@
 
-A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the different landscapes they inhabited.
-Often this kind of analyses depends on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g.
+A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the environmental variations they encountered.
+When analysing the residues of these behaviours, archaeologists often depend on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g.
 Barton 2003).
@@
 This edge effect was also extended to human culture, with Gummerman and Johnson (1971) arguing for increased human cultural diversity in ecotonal zones of central Arizona.
+In Australia edge effects in landscape resources and their exploitation have also been discussed.
+For instance, Hughes and Hiscock (1982) noted apparent edge effects in the concentration of stone artefact scatters along landscape boundaries in arid northern South Australia.
 
@@
 (1982), who used distribution of artefacts to see how culture group areas related to landform boundaries.
-However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these are often easier to measure and interpret than landscape boundaries (Daniel 2001; Jones, Beck et al.
+However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these can be easier to measure and interpret than landscape boundaries (Daniel 2001; Jones, Beck et al.
 2003).
@@
 
-In this study we ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of the optimum locations for their activities in respect to environmental boundaries.
+In this study we use the results from a sample of hand-held GIS-enabled archaeological site recordings to ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of the optimum locations for their activities in respect to environmental boundaries.
 We predict that prehistoric foragers were sensitive to subtle differences in resource structure as gradients across the landscape.
@@
 For example, the spatial scales of much previous work on prehistoric arid zone foragers has ranged from continental (Williams, Ulm et al.
-2013) to ecological regions (Holdaway, Douglass et al.
+2013; Bird et al 2016) to ecological regions (Holdaway, Douglass et al.
 2013).
-In this paper, we focus on human behavior in the arid zone at a novel scale and location: the interface between two types of subtly different landscapes.
-We show that prehistoric forager groups’ decisions about how they occupied the study area were sensitive to subtle differences in local landscape characteristics.
+In this paper, we introduce a novel focus on human behavior in the arid zone: the interface between two types of subtly different landscapes.
+We show that prehistoric forager groups' decisions about how they occupied the study area were sensitive to subtle differences in local landscape characteristics.
 
@@
 2013, Davies, Holdaway et al.
-2015).
+2015; Gould and Saggers 1985).
 The archaeological signature of landuse on these landscapes has typically been the distribution and composition of surface stone artefact scatters, which are by far the most abundant type of site.
 Sites in these landscapes are typically small, dense and discrete concentrations of stone artefacts that represent temporary, short-term activity areas.
-The spatially discrete quality of the majority of sites in these desert regions, such as our study area, means it is possible to distinguish between even quite close sites (<50 m apart) with confidence, unlike many other contexts where artifact distributions are more spatially continuous (Thomas 1975, Foley 1981, Dunnell and Dancey 1983, Kuna 2000).
 
@@
 names(table1) <- c("Landform type", "Archaeological prediction")
-kable(table1, caption = "Landform types and predicted archaeological signature. Adapted from Hughes et al. (2011), Kinhill-Stearns Roger (1982) and Hughes and Sullivan (1984).")
-```
-
-Our study area at Olympic Dam, South Australia, offers a unique opportunity to study landscape interfaces for two key reasons.
+kable(table1, caption = "Landform types and predicted archaeological signature.
+Adapted from Hughes et al.
+(2011); Kinhill-Stearns Roger (1982); Hughes and Sullivan (1984); and Sullivan et al.
+(2014a).")
+```
+
+Our study area in the Roxby dunefield in northern South Australia, offers a unique opportunity to study landscape interfaces for two key reasons.
 First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (Table \@ref(tab:table1)).
@@
 The relatively high density of sites provides an opportunity to investigate fine-grained gradients of archaeological change over multiple landscapes and their interfaces.
-More specifically, we can test whether a boundary effect exists at the edge of landscape units by examining how the size and density of surface scatters of stone artefacts changes with their distance from the boundaries.
-
-## Olympic Dam and investigation of its archaeology
+More specifically, we can test whether a boundary effect exists at the edge of landscape units by examining how the size and density of surface clusters of stone artefacts changes with their distance from the boundaries.
+
+## The Roxby dunefield and investigation of its archaeology
 
@@
 
-Olympic Dam is located on the southeastern edge of the driest part of the Australian arid zone (Figure \@ref(fig:overviewmap)).
+The Roxby dunefield is in the driest part of the Australian arid zone (Figure \@ref(fig:overviewmap)).
 Archaeological material is found across the broader region and partly reflects the long period of human occupation, the wide range of locally available resources and the extremely low rates of sediment movement that might otherwise obscure surface artefacts.
-Optically stimulated luminescence (OSL) dating of sediments underling buried artefacts at one typical site we investigated indicates a terminus post quem of 5 k BP for the artifact bearing deposit (Sullivan, Field et al.
-2012).
-Excavations at a second site produced five OSL ages for an open-air artefact quarrying/knapping location.
-These ages indicate that the site was used over relatively short time period, at about 2.3–2.4 ka, and again within the last 1 ka (Sullivan et al.
-2014).
+Optically stimulated luminescence (OSL) dating of sediments underling buried artefacts in several sites demonstrate episodic human use of the area from the late Pleistocene deglacial period (from around 19 ka, Hughes et al.
+2014b; Sullivan, Field et al.
+2012; Sullivan et al.
+2014!!b).
 No organic materials have been found at any of the open sites or in the excavated deposits.
-Such ages are consistent with the majority of dates for human occupation of sand deserts elsewhere in Australia, where evidence has suggested a substantial expansion of human activity in the last few millennia (Veth, Smith et al.
-2001, Smith 2006).
-A pair of dated sites are insufficient to draw conclusions beyond the local level, however the distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact scatters recorded during the survey (Hughes and Hiscock 2005).
+Distinctive implement forms such as backed artefacts, unifacial points and tula adzes at many sites in the study area indicate a mid to late Holocene age for the main period of land use and for most of the surface artefact clusters recorded during the survey (Hughes and Hiscock 2005).
 A study of 78 OSL ages from geological contexts in the study area has shown that the mid to late Holocene period was a time of broad-scale dune stability (Hughes et al.
-2014).
+2014b).
 This means that site preservation and visibility is not strongly controlled by dune movement.
@@
 
-```{r mapofstudyarea1, fig.cap="Map of study area with landforms and terrain patterns", fig.width=fig_width_two_cols}
+```{r mapofstudyarea1, fig.cap="Map of our study area within the Roxby dunefield, showing landforms and terrain patterns", fig.width=fig_width_two_cols}
 
@@
 
-```{r mapofstudyarea2, fig.cap="Map of study area showing the distribution of archaeological sites used in this study", fig.width=fig_width_two_cols}
+```{r mapofstudyarea2, fig.cap="Map of study area showing the distribution of archaeological sites sampled for this study", fig.width=fig_width_two_cols}
 
@@
 
-The Olympic Dam area has been the target of mineral exploration and mining since the 1980s.
-The extensive records of archaeological sites at Olympic Dam have been created during research-orientated consultancy work related to this exploration and mining activity.
-In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the proposed Olympic Dam mining project.
+The Roxby dunefield has been the target of mineral exploration and mining since the 1980s.
+The extensive records of archaeological sites near the Olympic Dam mine site in the duenfield have been created during research-orientated consultancy work related to this exploration and mining activity.
+In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the then-proposed mining project.
 They used terrain patterns based on combining categories of different landforms and surface geology (Figure \@ref(fig:mapofstudyarea1)) to develop an environmentally-based model to predict the nature and distribution of archaeological sites (Hughes and Hiscock 1982).
@@
 The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites (Hiscock and Hughes 1983, Hiscock 1989).
-In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009.
+In 2007 a proposed expansion of the existing mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009.
 The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).
@@
 
-```
-
-Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 16,000 archaeological sites.
-We achieved close to 90% coverage of 353 km^2^.
-The sites consist of stone artefact scatters (`r artefact_scatters`%), artefact scatters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al.
+# how many sites?
+total_number_of_sites <- prettyNum(nrow(ODX_allsites_2010), big.mark = ",")
+
+```
+
+Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 17,000 archaeological sites (Sullivan et al.
+2014a).
+In this study we draw on a sample of `r total_number_of_sites` sites from a 353 km^2^ area within the Roxby dunefield (Figure \@ref(fig:mapofstudyarea2)).
+The sites consist of stone artefact clusters (`r artefact_scatters`%), artefact clusters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al.
 2011).
 Extended descriptions of site definitions and survey methods have previously been presented in Hughes et al.
-(2011).
+(2011) and Sullivan et al.
+2014a.
 
@@
 
-```{r methodschematic, fig.cap="Schematic of two approaches used to investigate boundary effects at Olympic Dam, South Australia.
+```{r methodschematic, fig.cap="Schematic of two approaches used to investigate boundary effects in the Roxby duenfields, South Australia.
 A: the straight-line distances from sites to the boundaries of the terrain patterns that they are contained it was measured.
@@
 
-We used two approaches to investigate the presence and character of boundary effects on the Olympic Dam landscape (Figure \@ref(fig:methodschematic)).
+We used two approaches (Figure \@ref(fig:methodschematic)) to investigate the presence and character of boundary effects in the Roxby dunefield – a landscape consisting predominantly of longitudinal sand dunes and stony desert (gibber plains).
 In the first approach, we determined the straight-line distance from every site to the nearest boundary of the terrain pattern in which it was contained.
@@
 All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository.
-In our SOM our code is released under the MIT licence, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).
+In our SOM our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).
 
@@
 
-The data from Olympic Dam allow a preliminary test of the predictions of the landuse model developed for the region in the early 1980s (Hughes, Hiscock et al.
+The data from the Roxby dunefield allow a test of the predictions of the landuse model developed for the region in the early 1980s (Hughes, Hiscock et al.
 2011).
@@
 
-We found that in there were both more archaeological sites per square kilometer, and more artefacts per square kilometer, in landforms with more widely spaced dunes.
-This is consistent with our model, and our field observations that more and denser sites tended to occur occur near exposures of good-quality flakeable stone that are often found in interdunal swales.
+We found both more archaeological sites per square kilometer, and more artefacts per square kilometer, in landforms with more widely spaced dunes.
+This is consistent with our model, and our field observations that more and denser sites tended to occur occur near exposures of good-quality flakeable stone that are often found in interdunal swales (Hughes et al.
+2014a).
 In areas where the sand dunes are more closely spaced, such as landform types 5 and 6, the dune flanks flow into each other, leaving few or no swales between the dunes where surface outcrops of flakable stone might be exposed.
-This differential availability of flakable stone may be an important factor that influenced ancient peoples' use of the landscape.
-
-Our analysis of boundary effects shows people were sensitive to subtle gradients of change in the landscape.
+This differential availability of flakable stone may be an important factor that influenced ancient peoples' use of this landscape.
+
+Our analysis of boundary effects shows people were sensitive to subtle gradients of change in the physical environment.
 We found that sites were generally located further from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
@@
 
-This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns.
+This analysis suggests that prehistoric people in the Roxby dunefield  area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns.
 The novel element of the data presented here is that the distribution of human activities varies substantially and predictably within as well as between terrain patterns.
@@
 
-In addition to these observations about landscape use, our survey activity in the Olympic Dam region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
+In addition to these observations about landscape use, our survey activity in the  Roxby dunefield region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
 Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM).
@@
 # print table
-kable(display_table, caption = "Summary of archaeological survey productivity at Olympic Dam.
+kable(display_table, caption = "Summary of archaeological survey productivity in the Roxby dunefield.
 The data for sites recorded per person per day for before 2007 come from a survey undertaken in 1982 by Philip Hughes and Peter Hiscock that recorded 176 sites in 50 person days.")
@@
 
-The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Olympic Dam area were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
+The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Roxby dunefield were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
 The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites the reveal how people adapted to the landscape (cf.
@@
 
-This work was undertaken under a contract with BHP Billiton that includes research support.
+The fieldwork component of this study was undertaken as part of a broader investigation in accordance with the Olympic Dam Agreement between BHP Billiton and the Aboriginal native title claimant groups under a contract with BHP Billiton that included research support.
 Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301), a grant from the Herbert H.
@@
 Bevan, A. and J. Conolly (2004). "GIS, Archaeological Survey, and Landscape Archaeology on the Island of Kythera, Greece." Journal of Field Archaeology 29(1-2): 123-138.
+
+Bird, M. I., O’Grady, D., & Ulm, S. (2016). "Humans, water, and the colonization of Australia." Proceedings of the National Academy of Sciences, 113(41), 11477-11482.
 
@@
 
+Gould, R. A., & Saggers, S. (1985). Lithic procurement in Central Australia: a closer look at Binford's idea of embeddedness in archaeology. American Antiquity, 117-136.
+Chicago
+
 Gumerman, G.
@@
 Hughes, P., M.
+Sullivan, P.
+Hiscock and A.
+Neyland (2014a) The Need to Have Understood Your Local Geology: Nature and Sources of Materials Used to Manufacture Stone Artefacts at Olympic Dam, South Australia.
+Journal of the Anthropological Society of South Australia 34: 108-135.
+
+Hughes, P., M.
 Sullivan, F.
@@
 Spooner and D.
-Questiaux (2014).	Sand dune formation over the last 140 ka in the Roxby dunefield, arid northern South Australia.
+Questiaux (2014b).	Sand dune formation over the last 140 ka in the Roxby dunefield, arid northern South Australia.
 Quaternary Australasia 31(1):11.
@@
 
+Sullivan, M., P.
+Hiscock, and P.
+Hughes (2014) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia.
+Journal of the Anthropological Society of South Australia 38: 85-107.
+
 Sullivan, M., P. Hughes, A. Mosig Way and N. Spooner (2014).	Prehistoric mining at Olympic Dam in arid South Australia. Archaeology in Oceania 49(1):43-55.

### 2017-02-04 analysis/paper/paper.Rmd
make citations and references more consistent, remove data files from repo

@@
 ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes.
-Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972).
+Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968; Glassow and Leone 1972).
 This increased variety and density of plants and animals, often described as an ‘edge effect’, has been a focus of archaeological research.
@@
 
-These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists had been overly simplistic in their use of many ecological concepts (Rhoades 1978, Hardesty 1980, King and Graham 1981).
+These uses of the ecotone and edge effect concepts declined in the archaeological literature after several critiques pointed out that archaeologists had been overly simplistic in their use of many ecological concepts (Rhoades 1978; Hardesty 1980; King and Graham 1981).
 Problems with these previous approaches include ignoring controversies among biologists about the definition of ecotones, unjustified assumptions about the uniformity of all ecotones, and unverified assumptions about edge effects always increasing the amount of food species available to humans.
@@
 (1982), who used distribution of artefacts to see how culture group areas related to landform boundaries.
-However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these can be easier to measure and interpret than landscape boundaries (Daniel 2001; Jones, Beck et al.
+However, much recent work on archaeological settlement patterns of hunter-gatherers has eschewed boundaries in favour of analysing straight line distances from key resources such as raw materials for stone artefacts, as these can be easier to measure and interpret than landscape boundaries (Daniel 2001; Jones et al.
 2003).
@@
 The step-wise approaches are useful because they make comparative landscape analysis more tractable in many contexts and at different scales.
-For example, the spatial scales of much previous work on prehistoric arid zone foragers has ranged from continental (Williams, Ulm et al.
-2013; Bird et al 2016) to ecological regions (Holdaway, Douglass et al.
+For example, the spatial scales of much previous work on prehistoric arid zone foragers has ranged from continental (Williams et al.
+2013; Bird et al 2016) to ecological regions (Holdaway et al.
 2013).
@@
 
-Our work is motivated by previous archaeological work on cultural landscapes of foragers in Australian arid and semi-arid zones that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982, Veth 1993, Napton and Greathouse 1996, Holdaway 1998, Barton 2003, Smith 2006, Fanning, Holdaway et al.
-2008, Smith, Williams et al.
-2008, Veth, Smith et al.
-2008, Williams, Ulm et al.
-2013, Davies, Holdaway et al.
+Our work is motivated by previous archaeological work on cultural landscapes of foragers in Australian arid and semi-arid zones that emphasizes the relationship between settlement organization and both exploitable resources and prominent landscape features such as dunes, occurrences of flakeable rocks, pans, springs and salt lakes (Hughes and Hiscock 1982; Veth 1993; Napton and Greathouse 1996; Holdaway 1998; Barton 2003; Smith 2006; Fanning et al.
+2008; Smith et al.
+2008; Veth et al.
+2008; Williams et al.
+2013; Davies et al.
 2015; Gould and Saggers 1985).
@@
 Optically stimulated luminescence (OSL) dating of sediments underling buried artefacts in several sites demonstrate episodic human use of the area from the late Pleistocene deglacial period (from around 19 ka, Hughes et al.
-2014b; Sullivan, Field et al.
+2014b; Sullivan et al.
 2012; Sullivan et al.
-2014!!b).
+2014b).
 No organic materials have been found at any of the open sites or in the excavated deposits.
@@
 The Roxby dunefield has been the target of mineral exploration and mining since the 1980s.
-The extensive records of archaeological sites near the Olympic Dam mine site in the duenfield have been created during research-orientated consultancy work related to this exploration and mining activity.
+The extensive records of archaeological sites near the Olympic Dam mine site in the dunefield have been created during research-orientated consultancy work related to this exploration and mining activity.
 In 1980 Hughes, Hiscock and colleagues commenced a range of archaeological investigations for the then-proposed mining project.
@@
 Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al.
-2011; Hughes and Hiscock 1982; Hughes and Sullivan 1984).
+2011; Hughes and Hiscock 1982).
 In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts.
@@
 These descriptions were initially tested in the 1980s and 1990s by archaeological surveys of the mining area and infrastructure corridors.
-The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites (Hiscock and Hughes 1983, Hiscock 1989).
+The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites (Hiscock and Hughes 1983; Hiscock 1989).
 In 2007 a proposed expansion of the existing mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009.
@@
 In this study we draw on a sample of `r total_number_of_sites` sites from a 353 km^2^ area within the Roxby dunefield (Figure \@ref(fig:mapofstudyarea2)).
-The sites consist of stone artefact clusters (`r artefact_scatters`%), artefact clusters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al.
+The sites consist of stone artefact clusters (`r artefact_scatters`%), artefact clusters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes et al.
 2011).
@@
 (2011) and Sullivan et al.
-2014a.
+(2014a).
 
 To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information.
-Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003, Bevan and Conolly 2004, Tripcevich 2004, 2004, Given and Hyla 2006, Wagtendonk and De Jeu 2007, de la Vega and Agulla 2010, Tripcevich and Wernke 2010, Fei 2011, Scianna and Villa 2011, Traviglia 2011, Fee, Pettegrew et al.
-2013, Newhard, Levine et al.
-2013, Cascalheira, Gonçalves et al.
-2014, Banning and Hitchings 2015).
+Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003; Bevan and Conolly 2004; Tripcevich 2004a; 2004b; Given and Hyla 2006; Wagtendonk and De Jeu 2007; de la Vega and Agulla 2010; Tripcevich and Wernke 2010; Fei 2011; Scianna and Villa 2011, Traviglia 2011, Fee et al.
+2013; Newhard et al.
+2013; Cascalheira et al.
+2014; Banning and Hitchings 2015).
 Our mobile GIS configuration files, recording forms, and scripts are archived online at <http://doi.org/10.5281/zenodo.11833>.
@@
 To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/10.17605/OSF.IO/RMKGE>.
-Also in this version-controlled compendium are the raw data for all the tests and visualizations reported here, as well as additional regression diagnostics.
-All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository.
-In our SOM our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).
+Also in this version-controlled compendium are additional regression diagnostics.
+To respect the wishes of our stakeholders and protect the site locations from damage we have not made our data openly available.
+In our SOM our code is released under the MIT license, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).
 
@@
 
-The data from the Roxby dunefield allow a test of the predictions of the landuse model developed for the region in the early 1980s (Hughes, Hiscock et al.
+The data from the Roxby dunefield allow a test of the predictions of the landuse model developed for the region in the early 1980s (Hughes et al.
 2011).
@@
   xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
-  ylab("z-score of site density in buffer area")  +
-  labs(shape="Terrain\nPattern", colour="Terrain\nPattern")
+  ylab("z-score of site density \nin buffer area")  +
+  labs(shape="Terrain\nPattern", colour="Terrain\nPattern") +
+  theme(legend.position = c(0.4, 1.01), 
+        legend.justification = c(0.01, 1.05))
 
@@
 This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.
+
 Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.
+Our gain of `r percentage_increase`% is within the range documented by other studies using mobile data collection technologies.
+For example, Poehler and Ellis (2012) report a 371% increase in efficiency using iPads at Pompeii, and Austin (2014) reports a 21-32% increase when using Open Data Kit (Hartung et al.
+2010) to record skeletal remains.
+This wide variation in efficiency gains suggests that the magnitude of the gain is highly dependent on the context in which the mobile technologies are employed.
 
@@
 Attwell and Fletcher 1987).
-These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007, Kowalewski 2008: 233).
+These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007; Kowalewski 2008: 233).
 Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with relatively simple behavioral ecological models.
@@
 Attwell, M. R. and M. Fletcher (1987). "An analytical technique for investigating spatial relationships." Journal of Archaeological Science 14(1): 1-11.
+
+Austin, A. 2014 Mobilizing Archaeologists: Increasing the Quantity and Quality of Data Collected in the Field with Mobile Technology. Advances in Archaeological Practice 2(1):13-23.
 
@@
 Krek: 35-46.
+
 Dunnell, R. C. and W. S. Dancey (1983). "The Siteless Survey: A Regional Scale Data Collection Strategy." Advances in Archaeological Method and Theory 6: 267-287.
@@
 Harris, D. R., P. J. Ucko and G. Dimbleby (1969). "Agricultural systems, ecosystems and the origins of agriculture." The domestication and exploitation of plants and animals.: 3-15.
+
+Hartung, C., A.
+Lerer, Y.
+Anokwa, C.
+Tseng, W.
+Brunette and G.
+Borriello 2010 Open data kit: Tools to build information services for developing regions.
+Proceedings of the 4th ACM/IEEE International Conference on Information and Communication Technologies and Development: ACM.
 
@@
 
+Hughes, P. J. and P. Hiscock (1982). Annex 5. Olympic Dam Draft Environmental Impact Statement. Unpublished Report to Roxby Management Services Pty Ltd through Kinhill Stearns Roger Pty Ltd, Adelaide.
+
+Hughes, P., P.
+Hiscock, M.
+Sullivan and B.
+Marwick (2011).
+"An outline of archaeological investigations for the Olympic Dam Expansion in arid northeast South Australia." Journal of the Anthropological Society of South Australia 34: 21-37
+
 Hughes, P., M.
@@
 Quaternary Australasia 31(1):11.
-
-Hughes, P., P.
-Hiscock, M.
-Sullivan and B.
-Marwick (2011).
-"An outline of archaeological investigations for the Olympic Dam Expansion in arid northeast South Australia." Journal of the Anthropological Society of South Australia 34: 21-37
-
-Hughes, P. J. and P. Hiscock (1982). Annex 5. Olympic Dam Draft Environmental Impact Statement. Unpublished Report to Roxby Management Services Pty Ltd through Kinhill Stearns Roger Pty Ltd, Adelaide.
 
@@
 
+Marwick, B.
+(2016).
+Computational reproducibility in archaeological research: Basic principles and a case study of their implementation.
+Journal of Archaeological Method and Theory, 1-27, doi: 10.1007/s10816-015-9272-9
 Napton, L.
@@
 
+Poehler, E.E. and S.J.R. Ellis 2012 The 2011 Season of the Pompeii Quadriporticus Project: The Southern and Northern Sides. Fasti Online Documents and Research 249:1-12.
+
 Rhoades, R. E. (1978). "Archaeological Use and Abuse of Ecological Concepts and Studies: The Ecotone Example." American Antiquity 43(4): 608-614.
@@
 Struever, S. (1968). Woodland subsistence-settlement systems in the lower Illinois Valley. New perspectives in archaeology. S. R. Binford and L. R. Binford. Chicago, Aldine: 285-312.
-
-Sullivan, M., P.
-Hiscock, and P.
-Hughes (2014) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia.
-Journal of the Anthropological Society of South Australia 38: 85-107.
-
-Sullivan, M., P. Hughes, A. Mosig Way and N. Spooner (2014).	Prehistoric mining at Olympic Dam in arid South Australia. Archaeology in Oceania 49(1):43-55.
 
@@
 
+Sullivan, M., P.
+Hiscock, and P.
+Hughes (2014a) Three Scales: GIS, GPS and Digital Site and Data Recording Technology in Archaeological Salvage at Olympic Dam in Arid South Australia.
+Journal of the Anthropological Society of South Australia 38: 85-107.
+
+Sullivan, M., P. Hughes, A. Mosig Way and N. Spooner (2014b). Prehistoric mining at Olympic Dam in arid South Australia. Archaeology in Oceania 49(1):43-55.
+
 Thomas, D. H. (1975). Nonsite sampling in archaeology: up the creek without a site. Sampling in archaeology. J. W. Mueller. Tucson, AZ, University of Arizona Press: 61-81.
+
 Traviglia, A. (2011). Integrated Archaeological Investigations for the Study of the Greater Aquileia Area. The New Technologies for Aquileia, Aquileia, Italy.
 
-Tripcevich, N. (2004). "Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey." Cartography and Geographic Information Science 31(3): 137-151.
-
-Tripcevich, N. (2004). "Mobile GIS in archaeological survey." The SAA Archaeological Record(May).
+Tripcevich, N. (2004a). "Flexibility by Design: How Mobile GIS Meets the Needs of Archaeological Survey." Cartography and Geographic Information Science 31(3): 137-151.
+
+Tripcevich, N. (2004b). "Mobile GIS in archaeological survey." The SAA Archaeological Record(May).
 

### 2017-02-16 analysis/paper/paper.Rmd
prepare to submit to JASR

@@
 - Ben Marwick (corresponding author)
+- Marjorie Sullivan
 - Peter Hiscock 
 - Philip Hughes 
-- Marjorie Sullivan
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
+Marjorie Sullivan   
+The University of Sydney, Sydney NSW, Australia      
+Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
+
 Philip Hughes  
 Flinders University, Adelaide, South Australia, Australia     
-Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
-
-Marjorie Sullivan   
-The University of Sydney, Sydney NSW, Australia
 Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia
@@
 Adapted from Hughes et al.
-(2011); Kinhill-Stearns Roger (1982); Hughes and Sullivan (1984); and Sullivan et al.
+(2011), Kinhill-Stearns Roger (1982), Hughes and Sullivan (1984), and Sullivan et al.
 (2014a).")
@@
 2014b).
-This means that site preservation and visibility is not strongly controlled by dune movement.
+This means that site preservation and visibility is not strongly controlled by dune movement during that period.
 In 2013 we returned to sites that had been recorded 25 years earlier and found them to be in the same state as originally recorded, indicating that site visibility has changed little over 25 years.
@@
 In 2007 a proposed expansion of the existing mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009.
-The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).
+The data collected during that survey is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).
 
@@
 Also in this version-controlled compendium are additional regression diagnostics.
-To respect the wishes of our stakeholders and protect the site locations from damage we have not made our data openly available.
+To respect the wishes of our stakeholders and protect the site locations from potential damage we have not made our data openly available.
 In our SOM our code is released under the MIT license, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).

### 2017-02-16 README.Rmd
update DOI
update readme

@@
 
-http://dx.doi.org/xxxxxxx
+http://dx.doi.org/10.17605/OSF.IO/RMKGE
 
 The files at the URL above will generate the results as found in the publication.
-The files hosted at github.com/benmarwick/olympicdamboundaries are the development versions and may have changed since the report was published
+The files hosted at <https://github.com/benmarwick/olympicdamboundaries> are the development versions and may have changed since the report was published
 
@@
 This repository is our research compendium for our analysis of xxxx.
-The compendium contains all data, code, and text associated with the publication.
+The compendium contains all code, and text associated with the publication (data are not included for reasons explained in the manuscript).
 The `Rmd` files in the `analysis/paper/` directory contain details of how all the analyses reported in the paper were conducted, as well as instructions on how to rerun the analysis to reproduce the results.
@@
 - the manuscript as submitted (in MS Word format) and it's Rmd source file
-- all the data files (in CSV format, in the `data/` directory)    
 - supplementary information source files (in R markdown format) and executed versions

### 2017-05-25 analysis/paper/paper.Rmd
dealing with AE's comments

@@
 - Ben Marwick (corresponding author)
+- Peter Hiscock      
 - Marjorie Sullivan     
-- Peter Hiscock 
 - Philip Hughes       
@@
 
-The second reason why our data is uniquely suited to investigating landscape boundary effects is that we have a large number of sites spread across several landscape types.
+The second reason why our data are uniquely suited to investigating landscape boundary effects is that we have a large number of sites spread across several landscape types.
 The relatively high density of sites provides an opportunity to investigate fine-grained gradients of archaeological change over multiple landscapes and their interfaces.
@@
 ## Results
+
+### Efficiency of a mobile GIS compared to paper systems for field survey
+
+Our survey activity in the Roxby dunefield region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
+Here we briefly explore the question of whether our mobile GIS enabled faster site recording compared to when we used a paper-pen-GPS method (these methods are described in more detail in our SOM).
+Table \@ref(tab:mobilegisefficiency) shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS.
+Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day.
+This metric is useful because it is easy to measure without burdening field workers with additional data-logging tasks.
+However, we need to take into account possible confounding factors, such as travel time to the day’s survey area, the size and density of the sites in the day’s survey area, skill and efficiency of the individual archaeologist and the sampling strategy.
+
+```{r mobilegisefficiency}
+
+# checking benefit of mobile GIS
+dat <- read.table(text = "
+Year Pre.2007	2007	2008	2009
+Area.surveyed.(km2)	35	95	275	145
+Rate.of.survey.(km2/person/day)	0	0.3	0.3	0.3
+Total.no.persondays	0 317	917	483
+Approximate.number.of.sites.recorded	665	3900	6600	5500
+No.sites.recorded/person/day	 3.5	12.3	7.2	11.3
+Frequency.of.occurrence.of.sites/km2	19	41	24	38", row.names = 1, header = TRUE)
+
+display_table <- dat
+display_table$"" <- row.names(display_table)
+row.names(display_table) <- NULL
+display_table <- display_table[ , c(5, 1:4)]
+names(display_table) <- c("", "pre-2007", "2007", "2008", "2009")
+display_table[ , 1] <- gsub("\\.", " ", display_table[ , 1])
+display_table[ , 1] <- gsub("km2", "km^2^", display_table[ , 1])
+display_table[ , 1] <- gsub("persondays", "person-days", display_table[ , 1])
+
+# print table
+kable(display_table, caption = "Summary of archaeological survey productivity in the Roxby dunefield.
+The data for sites recorded per person per day for before 2007 come from a survey undertaken in 1982 by Philip Hughes and Peter Hiscock that recorded 176 sites in 50 person days.")
+
+# modelling these data
+
+dat_t <- data.frame(t(dat))
+
+# rate for pre-2007 and 2007-2009
+rate_for_pre_2007 <- dat_t$No.sites.recorded.person.day[1]  
+rate_for_2007_thru_2009 <-  round(mean(dat_t$No.sites.recorded.person.day[c(2:4)]),1)
+
+# the rate of recording is strongly correlated with the density of sites 
+# linear model for sites recorded by density of sites
+mobile_gis_efficiency_lm <- 
+  summary(mobile_gis_efficiency_lm_out <- lm(dat_t$No.sites.recorded.person.day ~ dat_t$Frequency.of.occurrence.of.sites.km2))
+mobile_gis_efficiency_lm_r_squared <- round(mobile_gis_efficiency_lm$adj.r.squared, 3)
+mobile_gis_efficiency_lm_p_value <- round(mobile_gis_efficiency_lm$coefficients[2, 4]  , 3)
+mobile_gis_efficiency_lm_df <- round(mobile_gis_efficiency_lm$df[2] , 3)
+mobile_gis_efficiency_lm_F <- round(mobile_gis_efficiency_lm$fstatistic[1] , 3)
+
+# a linear model for the 2007-2009 data
+# subset 2007-2009 data
+dat_t_s <- dat_t[2:4,]
+# linear model for sites recorded by density of sites, 2007-2009
+mobile_gis_efficiency_lm_2007_2009 <-
+ summary(mobile_gis_efficiency_lm_2007_2009_out <- 
+ lm(dat_t_s$No.sites.recorded.person.day ~ dat_t_s$Frequency.of.occurrence.of.sites.km2))
+
+# display model
+mobile_gis_efficiency_lm_2007_2009_est <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,1], 3)
+mobile_gis_efficiency_lm_2007_2009_est_error <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,2], 3)
+mobile_gis_efficiency_lm_2007_2009_inter <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,1], 3)
+mobile_gis_efficiency_lm_2007_2009_inter_error <-   round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,2], 3)
+mobile_gis_efficiency_lm_2007_2009_r_sq_adj <- round(mobile_gis_efficiency_lm_2007_2009$adj.r.squared, 3)
+mobile_gis_efficiency_lm_2007_2009_p_value <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2, 4], 3)
+
+# predict given 19 sites/km2 of pre-2007
+resulting_rate <- round(19 * mobile_gis_efficiency_lm_2007_2009_est + mobile_gis_efficiency_lm_2007_2009_inter, 2)
+
+# difference in rates, pre-2007 and 2007-2009
+difference_in_rates_pre_2007_and_2007_2009 <- 
+resulting_rate - rate_for_pre_2007
+
+# percentage increase
+percentage_increase <- round((resulting_rate - rate_for_pre_2007) / rate_for_pre_2007 * 100, 0)
+```
+
+Our pre-2007 data were collected by more experienced archaeologists than the 2007-2009 data, using a recording form with less variables than our mobile GIS recording forms, and the pre-2007 survey areas were all closer to the camp, with less travel time compared to the 2007-2009 data.
+Further, the average site size for the pre-2007 data is less than the 2007-2009 data.
+These conditions should favor a higher rate of recording for the 2007-2009 data.
+Table \@ref(tab:mobilegisefficiency) shows that prior to adopting the mobile GIS in 2007 we recorded `r rate_for_pre_2007` sites per person per day.
+With the mobile GIS our rate increased substantially to an average of `r rate_for_2007_thru_2009` sites per person per day for 2007-2009.
+
+While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS.
+This is because the rate of recording is strongly predicted by the density of sites (F(1, `r mobile_gis_efficiency_lm_df`) = `r mobile_gis_efficiency_lm_F`, R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`).
+To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) + `r mobile_gis_efficiency_lm_2007_2009_est`(site density) (`r mobile_gis_efficiency_lm_2007_2009_est_error`) +  (standard errors in parentheses).
+This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate using a mobile GIS given the density of 19 sites/km^2^ observed in the pre-2007 data.
+The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^.
+The difference of `r difference_in_rates_pre_2007_and_2007_2009` sites/person/day between the actual pre-2007 average rate of `r rate_for_pre_2007` and the predicted average rate of `r resulting_rate` is most likely due to the use of mobile GIS.
+This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.
+
+Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.
+Our gain of `r percentage_increase`% is within the range documented by other studies using mobile data collection technologies.
+For example, Poehler and Ellis (2012) report a 371% increase in efficiency using iPads at Pompeii, and Austin (2014) reports a 21-32% increase when using Open Data Kit (Hartung et al.
+2010) to record skeletal remains.
+This wide variation in efficiency gains suggests that the magnitude of the gain is highly dependent on the context in which the mobile technologies are employed.
 
@@
 
-### Efficiency of a mobile GIS compared to paper systems for field survey
-
-In addition to these observations about landscape use, our survey activity in the Roxby dunefield region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods.
-Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM).
-Table \@ref(tab:mobilegisefficiency) shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS.
-Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day.
-This metric is useful because it is easy to measure without burdening field workers with additional data-logging tasks.
-However, we need to take into account possible confounding factors, such as travel time to the day’s survey area, the size and density of the sites in the day’s survey area, skill and efficiency of the individual archaeologist and the sampling strategy.
-
-```{r mobilegisefficiency}
-
-# checking benefit of mobile GIS
-dat <- read.table(text = "
-Year Pre.2007	2007	2008	2009
-Area.surveyed.(km2)	35	95	275	145
-Rate.of.survey.(km2/person/day)	0	0.3	0.3	0.3
-Total.no.persondays	0 317	917	483
-Approximate.number.of.sites.recorded	665	3900	6600	5500
-No.sites.recorded/person/day	 3.5	12.3	7.2	11.3
-Frequency.of.occurrence.of.sites/km2	19	41	24	38", row.names = 1, header = TRUE)
-
-display_table <- dat
-display_table$"" <- row.names(display_table)
-row.names(display_table) <- NULL
-display_table <- display_table[ , c(5, 1:4)]
-names(display_table) <- c("", "pre-2007", "2007", "2008", "2009")
-display_table[ , 1] <- gsub("\\.", " ", display_table[ , 1])
-display_table[ , 1] <- gsub("km2", "km^2^", display_table[ , 1])
-display_table[ , 1] <- gsub("persondays", "person-days", display_table[ , 1])
-
-# print table
-kable(display_table, caption = "Summary of archaeological survey productivity in the Roxby dunefield.
-The data for sites recorded per person per day for before 2007 come from a survey undertaken in 1982 by Philip Hughes and Peter Hiscock that recorded 176 sites in 50 person days.")
-
-# modelling these data
-
-dat_t <- data.frame(t(dat))
-
-# rate for pre-2007 and 2007-2009
-rate_for_pre_2007 <- dat_t$No.sites.recorded.person.day[1]  
-rate_for_2007_thru_2009 <-  round(mean(dat_t$No.sites.recorded.person.day[c(2:4)]),1)
-
-# the rate of recording is strongly correlated with the density of sites 
-# linear model for sites recorded by density of sites
-mobile_gis_efficiency_lm <- 
-  summary(mobile_gis_efficiency_lm_out <- lm(dat_t$No.sites.recorded.person.day ~ dat_t$Frequency.of.occurrence.of.sites.km2))
-mobile_gis_efficiency_lm_r_squared <- round(mobile_gis_efficiency_lm$adj.r.squared, 3)
-mobile_gis_efficiency_lm_p_value <- round(mobile_gis_efficiency_lm$coefficients[2, 4]  , 3)
-mobile_gis_efficiency_lm_df <- round(mobile_gis_efficiency_lm$df[2] , 3)
-mobile_gis_efficiency_lm_F <- round(mobile_gis_efficiency_lm$fstatistic[1] , 3)
-
-# a linear model for the 2007-2009 data
-# subset 2007-2009 data
-dat_t_s <- dat_t[2:4,]
-# linear model for sites recorded by density of sites, 2007-2009
-mobile_gis_efficiency_lm_2007_2009 <-
- summary(mobile_gis_efficiency_lm_2007_2009_out <- 
- lm(dat_t_s$No.sites.recorded.person.day ~ dat_t_s$Frequency.of.occurrence.of.sites.km2))
-
-# display model
-mobile_gis_efficiency_lm_2007_2009_est <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,1], 3)
-mobile_gis_efficiency_lm_2007_2009_est_error <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,2], 3)
-mobile_gis_efficiency_lm_2007_2009_inter <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,1], 3)
-mobile_gis_efficiency_lm_2007_2009_inter_error <-   round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,2], 3)
-mobile_gis_efficiency_lm_2007_2009_r_sq_adj <- round(mobile_gis_efficiency_lm_2007_2009$adj.r.squared, 3)
-mobile_gis_efficiency_lm_2007_2009_p_value <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2, 4], 3)
-
-# predict given 19 sites/km2 of pre-2007
-resulting_rate <- round(19 * mobile_gis_efficiency_lm_2007_2009_est + mobile_gis_efficiency_lm_2007_2009_inter, 2)
-
-# difference in rates, pre-2007 and 2007-2009
-difference_in_rates_pre_2007_and_2007_2009 <- 
-resulting_rate - rate_for_pre_2007
-
-# percentage increase
-percentage_increase <- round((resulting_rate - rate_for_pre_2007) / rate_for_pre_2007 * 100, 0)
-```
-
-Our pre-2007 data were collected by more experienced archaeologists than the 2007-2009 data, using a recording form with less variables than our mobile GIS recording forms, and the pre-2007 survey areas were all closer to the camp, with less travel time compared to the 2007-2009 data.
-Further, the average site size for the pre-2007 data is less than the 2007-2009 data.
-These conditions should favor a higher rate of recording for the 2007-2009 data.
-Table \@ref(tab:mobilegisefficiency) shows that prior to adopting the mobile GIS in 2007 we recorded `r rate_for_pre_2007` sites per person per day.
-With the mobile GIS our rate increased substantially to an average of `r rate_for_2007_thru_2009` sites per person per day for 2007-2009.
-
-While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS.
-This is because the rate of recording is strongly predicted by the density of sites (F(1, `r mobile_gis_efficiency_lm_df`) = `r mobile_gis_efficiency_lm_F`, R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`).
-To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) + `r mobile_gis_efficiency_lm_2007_2009_est`(site density) (`r mobile_gis_efficiency_lm_2007_2009_est_error`) +  (standard errors in parentheses).
-This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate using a mobile GIS given the density of 19 sites/km^2^ observed in the pre-2007 data.
-The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^.
-The difference of `r difference_in_rates_pre_2007_and_2007_2009` sites/person/day between the actual pre-2007 average rate of `r rate_for_pre_2007` and the predicted average rate of `r resulting_rate` is most likely due to the use of mobile GIS.
-This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.
-
-Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.
-Our gain of `r percentage_increase`% is within the range documented by other studies using mobile data collection technologies.
-For example, Poehler and Ellis (2012) report a 371% increase in efficiency using iPads at Pompeii, and Austin (2014) reports a 21-32% increase when using Open Data Kit (Hartung et al.
-2010) to record skeletal remains.
-This wide variation in efficiency gains suggests that the magnitude of the gain is highly dependent on the context in which the mobile technologies are employed.
-
 ## Conclusion
@@
 Attwell and Fletcher 1987).
+Although Australian deserts are home to some unique flora and fauna, the general patterns of gradients of human behaviour observed here are not contingent on any specific taxa.
+We expect that beyond the sand dunes of Australia, gradients might also be observed in the archaeological record of the U.S.
+Great Basin, the coastal and high altitude deserts of South America, and the core deserts of Africa.
+Gradients are likely to be an indicator of a generic adaptation to landscape variation that can be found in many ecosystems.
 These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007; Kowalewski 2008: 233).

### 2017-06-05 analysis/paper/paper.Rmd
working on peer reviewer's comments

@@
 We investigate how Holocene forager populations adapted to this variation in the linear sand dune desert of arid South Australia.
-We use data from surface scatters of stone artefacts collected during pedestrian survey to compare behaviors at landform boundaries to behaviours at the centers of landforms.
-We propose a model of human use of the landscape that predicts the prehistoric occupants of the study are were sensitive to the different economic potential of subtly dissimilar landscapes.
+We use data from surface scatters of stone artefacts collected during pedestrian survey to compare behaviours at landform boundaries to behaviours at the centers of landforms.
+We propose a model of human use of the landscape that predicts the prehistoric occupants of the study were sensitive to the different economic potential of subtly dissimilar landscapes.
 In evaluating the model we find that there are different densities of archaeological sites in each landscape type.
@@
 
-A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the environmental variations they encountered.
+A key step in understanding past human use of landscapes is investigating how people adapted their behaviours to the environmental variations they encountered.
 When analysing the residues of these behaviours, archaeologists often depend on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g.
 Barton 2003).
-Variations in human behavior at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences.
-Previous work on past human behaviors at ecological interfaces has focused on regions of highly contrasting biomes.
+Variations in human behaviour at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences.
+Previous work on past human behaviours at ecological interfaces has focused on regions of highly contrasting biomes.
 For example, Epp (1985) found that archaeological sites were more concentrated in an ecotone (i.e.
@@
 Here we pursue a new approach to understanding boundaries by focusing on subtle variations in landscape characteristics within a biome, rather than between them.
-By focusing on continuous variation, rather than step-wise shifts in the archaeological record across ecological interfaces, we introduce a new approach to studying behaviors of prehistoric forager groups at interface zones.
+By focusing on continuous variation, rather than step-wise shifts in the archaeological record across ecological interfaces, we introduce a new approach to studying behaviours of prehistoric forager groups at interface zones.
 
@@
 2013).
-In this paper, we introduce a novel focus on human behavior in the arid zone: the interface between two types of subtly different landscapes.
+In this paper, we introduce a novel focus on human behaviour in the arid zone: the interface between two types of subtly different landscapes.
 We show that prehistoric forager groups' decisions about how they occupied the study area were sensitive to subtle differences in local landscape characteristics.
@@
 
+Our study area in the Roxby dunefield in northern South Australia (Figure \@ref(fig:overviewmap)), offers a unique opportunity to study landscape interfaces for two key reasons.
+First, in our study area these interfaces are relatively well-defined by landform attributes, such as the surface geology (Table \@ref(tab:table2)) and physical features such as the spacing between sand dune ridges or the presence of water bodies (Table \@ref(tab:table1)).
+The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes and bounaries other physical features, making them convenient for spatial analysis of archaeological features (Figure \@ref(fig:aerialphoto)).
+With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.
+
+The second reason why our data are uniquely suited to investigating landscape boundary effects is that we have a large number of sites spread across several landscape types.
+The relatively high density of sites provides an opportunity to investigate fine-grained gradients of archaeological change over multiple landscapes and their interfaces.
+More specifically, we can test whether a boundary effect exists at the edge of landscape units by examining how the size and density of surface clusters of stone artefacts changes with their distance from the boundaries.
+
+```{r overviewmap, fig.cap="Map of region including the study area", fig.width=fig_width_one_col}
+
+knitr::include_graphics("../figures/regional_map.png")
+
+```
+
+##### . 
+<!-- page break -->
+
+```{r table2}
+table2 <- read.csv("../data/table_2.csv")
+names(table2) <- c("Geological regime", "Period", "Description", "Materials for artefact manufacture")
+kable(table2, caption = "Geological regimes in the study area and predicted archaeological importance. Adapted from Hughes et al. (2011).")
+```
+
+##### . 
+<!-- page break -->
+
 ```{r table1}
@@
 
-Our study area in the Roxby dunefield in northern South Australia, offers a unique opportunity to study landscape interfaces for two key reasons.
-First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (Table \@ref(tab:table1)).
-The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes, making them convenient for spatial analysis of archaeological features.
-With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.
-
-The second reason why our data are uniquely suited to investigating landscape boundary effects is that we have a large number of sites spread across several landscape types.
-The relatively high density of sites provides an opportunity to investigate fine-grained gradients of archaeological change over multiple landscapes and their interfaces.
-More specifically, we can test whether a boundary effect exists at the edge of landscape units by examining how the size and density of surface clusters of stone artefacts changes with their distance from the boundaries.
+##### . 
+<!-- page break -->
+
+```{r aerialphoto, fig.cap="Aerial photograph showing a sample of landscape types and their boundaries in the study area.", fig.width=fig_width_two_cols}
+
+knitr::include_graphics("../figures/aerial_photo.jpg")
+
+```
+
+##### . 
+<!-- page break -->
+
+```{r tp-q4, fig.cap="Aerial photograph showing Terrain Pattern Q4 in the foreground, consisting of widely spaced sand dunes with only poorly developed pans and no rock outcrop in the swales.
+The more closely spaced dunes in the background are in Terrain Pattern Q5", fig.width=fig_width_one_col}
+
+knitr::include_graphics("../figures/tp_q4_q5.jpg")
+
+```
+
+```{r tp-qs4, fig.cap="Aerial photograph showing Terrain Pattern Qs4, consisting of widely spaced dunes with well-developed pans and patchy areas of bare rocky ground exposed in the swales.
+The exposed rock is predominantly silcrete.", fig.width=fig_width_two_cols}
+
+knitr::include_graphics("../figures/tp_qs5.jpg")
+
+```
+
+##### . 
+<!-- page break -->
 
 ## The Roxby dunefield and investigation of its archaeology
-
-```{r overviewmap, fig.cap="Map of region including the study area", fig.width=fig_width_one_col}
-
-knitr::include_graphics("../figures/regional_map.png")
-
-```
 
@@
 
-```{r mapofstudyarea1, fig.cap="Map of our study area within the Roxby dunefield, showing landforms and terrain patterns", fig.width=fig_width_two_cols}
+```{r mapofstudyarea1, fig.cap="Map of our study area within the Roxby dunefield, showing landforms and terrain patterns.
+Each polygon is labelled with 1-3 letters that denote the surface geological regime (see Table \\@ref(tab:table2) for more details), and a number that denotes the landform type (see Table \\@ref(tab:table1) for more details).", fig.width=fig_width_two_cols}
 
@@
 
-```{r table2}
-table2 <- read.csv("../data/table_2.csv")
-names(table2) <- c("Geological regime", "Period", "Description", "Materials for artefact manufacture")
-kable(table2, caption = "Geological regimes in the study area and predicted archaeological importance. Adapted from Hughes et al. (2011).")
-```
-
 Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al.
@@
 The model results in a series of predictions about the archaeological record expected for each of six landform types (Table \@ref(tab:table1)) and five geological regimes (Table \@ref(tab:table2)).
-These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.
+These predictions specify the frequency of sites and the density and diversity of artefacts which were produced for each major landscape unit.
 
@@
 
-To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information.
+To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record background environmental, cadastral, and archaeological information.
 Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003; Bevan and Conolly 2004; Tripcevich 2004a; 2004b; Given and Hyla 2006; Wagtendonk and De Jeu 2007; de la Vega and Agulla 2010; Tripcevich and Wernke 2010; Fei 2011; Scianna and Villa 2011, Traviglia 2011, Fee et al.
@@
 
-We investigated behaviors at landform boundaries by analyzing the contrast between archaeological sites close to the boundary of a landform and the archaeology deeper within the landform.
+We investigated behaviours at landform boundaries by analyzing the contrast between archaeological sites close to the boundary of a landform and the archaeology deeper within the landform.
 Our test of boundary effects focused on the archaeological evidence from resource-rich landform types/terrain patterns, such as Czs4, bordered by landscapes with lower resource levels, such as landform types 5 or 6.
@@
 ```{r methodschematic, fig.cap="Schematic of two approaches used to investigate boundary effects in the Roxby duenfields, South Australia.
-A: the straight-line distances from sites to the boundaries of the terrain patterns that they are contained it was measured.
+A: the straight-line distances from sites to the boundaries of the terrain patterns that they are contained in was measured.
 B: the number of sites per square kilometer was measured in a series of buffer areas, an example of a buffer area is indicated here by a grey region at the exterior of the terrain pattern area.", fig.width=fig_width_two_cols}
@@
 In the first approach, we determined the straight-line distance from every site to the nearest boundary of the terrain pattern in which it was contained.
-Descriptive statistics for these distances allows testing of the prediction that sites tend to be closer to the edge of some landforms than other landforms.
+By comparing the distributions of these distances and using statistical tests we can see if there are diffrences in the patterns of site proximity to landform boundaries.
+These data will allow us to test the prediction that sites tend to be closer to the edge of some landforms than other landforms.
 
@@
 This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis.
-
-Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, data volume was high, a highly structured data collection form was used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.
-Our gain of `r percentage_increase`% is within the range documented by other studies using mobile data collection technologies.
-For example, Poehler and Ellis (2012) report a 371% increase in efficiency using iPads at Pompeii, and Austin (2014) reports a 21-32% increase when using Open Data Kit (Hartung et al.
-2010) to record skeletal remains.
-This wide variation in efficiency gains suggests that the magnitude of the gain is highly dependent on the context in which the mobile technologies are employed.
 
@@
 In total these areas contained about `r all_sites_in_study_area` archaeological sites.
-Most of the study area consists of widely spaced dunes – landform type 4 (`r LF4`%), moderately spaced dunes – landform type 5 (`r LF5`%) and closely spaced dunes – landform type 6 (`r LF6`%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P).
+Most of the study area consists of widely spaced dunes – landform type 4 (`r LF4`%), moderately spaced dunes – landform type 5 (`r LF5`%) and closely spaced dunes – landform type 6 (`r LF6`%), mainly overlying Tertiary stony-plains (Czs4 and Czs5) but some overlying limestone (A) or quartzite surfaces (K and P).
 Other terrain patterns make up less than 10% of the 353 km^2^ study area (Figure \@ref(fig:percentcoverageofeachterrainpattern)).
@@
 ```{r sitesinterrainpatterns, fig.cap="Site density in each terrain pattern.
-Annotation shows the result of a t-test to compare site densities in the two largest terrain patters, Czs4 and Czs5", fig.width=fig_width_two_cols}
+Annotation shows the result of a t-test to compare site densities in the two largest terrain patterns, Czs4 and Czs5", fig.width=fig_width_two_cols}
 ## sites in each TP
@@
 When considered as homogeneous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
-As predicted (Table \@ref(tab:table1)), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km^2^ compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km^2^ (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns).
+As predicted (Table \@ref(tab:table1)), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km^2^ compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km^2^ (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns)).
 Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km^2^, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km^2^.
@@
 ```{r distanceofsitetoboundary, fig.cap="Distribution of distances of archaeological sites to terrain pattern bounaries.
+Distance is shows the vertical axis with a logarithmic scale.
 Data points show distance values for individual sites.", fig.width=fig_width_one_col}
@@
 These gradients allow for more sophisticated interpretations of prehistoric behaviour beyond the obvious attraction of prominent landscape features (Wilson 2007; Kowalewski 2008: 233).
-Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with relatively simple behavioral ecological models.
+Given a large enough dataset, such as the one presented here, these patterns can be quickly quantified using a geographical information system and explained with relatively simple behavioural ecological models.
 We hope that this data-intensive approach to using gradients of changes in archaeological site attributes across the landscape might suggest a new direction for investigating past human behaviour at ecological interfaces, avoiding the problematic assumptions of previous work on edge effects.
+
+The methodological contribution of this study has been to demonstrate the advantage of using a mobile GIS for field data collection in a context where data collection is intensive, data volume is high, a highly structured data collection form is used, and the start-up, purchasing, and training costs could be amortized over multiple years of data collection.
+Our gain of `r percentage_increase`% is within the range documented by other studies using mobile data collection technologies.
+For example, Poehler and Ellis (2012) report a 371% increase in efficiency using iPads at Pompeii, and Austin (2014) reports a 21-32% increase when using Open Data Kit (Hartung et al.
+2010) to record skeletal remains.
+This wide variation in efficiency gains suggests that the magnitude of the gain is highly dependent on the context in which the mobile technologies are employed.
 

### 2017-06-05 analysis/paper/paper.Rmd
do ratio of observed site distance to max-min possible dist

@@
                message = FALSE,
-               dpi = 900)
+               dpi = 900,
+               cache = TRUE)
 options(warn = -1)
@@
 
-```{r distanceofsitetoboundary, fig.cap="Distribution of distances of archaeological sites to terrain pattern bounaries.
-Distance is shows the vertical axis with a logarithmic scale.
-Data points show distance values for individual sites.", fig.width=fig_width_one_col}
+```{r}
+# get shapefiles
+library(units)
+library(sf)
+odx_polygons <- read_sf("../data/ODX_TP_2010_data/CLipped_ODO_terrain_patterns_2010_Ja.shp")
+```
+
+```{r tpelongationandlength,  fig.cap="Upper: Distribution of elongation ratios of terrain pattern polygons in the study area.
+The ANOVA output in the upper right of the plot indicates that there is no significant difference in the elongation ratios of the terrain patterns.
+Lower: Distribution of calliper lengths (longest distance between two points) of the terrain patterns in the study area.
+The Krukall-Wallis test in the upper right of the plot indicates that there is no significant difference in the lengths of the terrain patterns.", fig.width=fig_width_two_cols}
+
+# compute elongation of each landform area
+
+# subset to get only TPs 1 km or more accross
+
+odx_polygons <- 
+  filter(odx_polygons,  GREGLANDF %in% c("A4", "Czs4", "Q4", "K4", "Czs5", "Q5", "Q6"))
+
+# simplify to reduce slivers
+
+library(rmapshaper)
+
+odx_polygons_simpl <- vector("list", length = nrow(odx_polygons))
+for(i in seq_len(nrow(odx_polygons))) {
+  t1 <- ms_simplify(as(odx_polygons[i, ], 'Spatial'), keep = 0.2)
+  t2 <- st_as_sf(t1)
+  odx_polygons_simpl[[i]] <- t2
+  # print(i)
+}
+
+# get coords of polygons to put into Momocs
+
+odx_polygons_coords <- apply(odx_polygons, 1,  function(i) st_coordinates(i$geometry))
+
+odx_polygons_coords <- lapply(odx_polygons_simpl, function(i) st_coordinates(i$geometry))
+
+# create Momocs outlines object
+
+library(Momocs)
+
+odx_polygons_outline <- Out(odx_polygons_coords,
+                            fac = data.frame(TP_NO = odx_polygons$TP_NO))
+
+# compute elongation for each shape
+
+odx_polygons_elongation <- 
+  data.frame(GEOLREGIME = odx_polygons$GEOLREGIME,
+             LF = odx_polygons$LANDFORM,
+             GREGLANDF = odx_polygons$GREGLANDF,
+             ID = odx_polygons$TP_NO,
+             elongation = sapply(odx_polygons_outline$coo, coo_elongation),
+             calliper_length = sapply(odx_polygons_outline$coo, coo_calliper))
+
+# how varied are the shapes of each TP?
+
+#  hist(odx_polygons_elongation$elong) it's normal
+tp_elong_anova <- summary(aov(elongation ~ GREGLANDF, data = odx_polygons_elongation))
+# no signifant difference
+
+fval <- round(tp_elong_anova[[1]]$`F value`[1], 3)
+pval <- round(tp_elong_anova[[1]]$`Pr(>F)`[1], 3)
+df <-  paste0(tp_elong_anova[[1]]$Df, collapse = ", ")
+
+anova_string <- paste0("F(", df, ") = ", fval, ", p = ", pval)
+
+library(ggplot2)
+elong <- ggplot(odx_polygons_elongation,
+       aes(reorder(GREGLANDF, -elongation),
+           elongation)) +
+  geom_boxplot() +
+  theme_bw(base_size = text_size-1) +
+  xlab("") +
+  ylab("Elongation ratio") +
+  annotate("text",  x = "Q5", y = 0.85, label = anova_string)
+
+# similar analysis for maximum dimension (calliper length or Feret's diameter)
+
+#  hist(odx_polygons_elongation$calliper_length) it's NOT normal
+tp_length_kruskal <- kruskal.test(calliper_length ~ GREGLANDF, data = odx_polygons_elongation)
+# no signifant difference
+
+hval <- unname(round(tp_length_kruskal$statistic, 3))
+pval <- round(tp_length_kruskal$p.value, 3)
+df <-  unname(tp_length_kruskal$parameter)
+
+kruskal_string <- paste0("H(", df, ") = ", hval, ", p = ", pval)
+
+library(ggplot2)
+calliper <- ggplot(odx_polygons_elongation,
+       aes(reorder(GREGLANDF, -calliper_length),
+           calliper_length/1000)) +
+  geom_boxplot() +
+  theme_bw(base_size = text_size-1) +
+  xlab("Terrain Pattern") +
+  ylab("Calliper length (km)") +
+  annotate("text",  x = "Q4", y = 18, label = kruskal_string)
+
+library(gridExtra)
+grid.arrange(elong, calliper, ncol = 1)
+
+```
+
+```{r distanceofsitetoboundary}
 
@@
 library(ggforce)
-ggplot(ODX_allsites_2010.s, 
+d2b <- ggplot(ODX_allsites_2010.s, 
        aes(reorder(GREGLANDF, -RASTERVALU),
@@
   ylab("Distance of site to boundary (m)")
+```
+
+```{r maxmindistancetoboundary, fig.cap="Left: Distributions of distances of archaeological sites to terrain pattern bounaries.
+Right: Distributions of ratios of observed distance of site to terrain pattern boundary and the maximum possible distance to boundary for each terrain pattern.
+Distance is shown on the vertical axis with a logarithmic scale.
+Data points show distance values for individual sites.", fig.width=fig_width_two_cols}
+
+# "a new variable should be created for each site which is a ratio of 
+# minimum distance to boundary [of that site] to maximum possible 
+# minimum distance to boundary for any location within a polygon."
+
+library(dplyr)
+library(tidyr)
+
+# make a fn to ut random points on a polygon, and compute areas to boundary for all points in that polygon
+
+library(sp)
+library(rgeos)
+
+random_points_to_boundary <- function(poly, n = 1000, ...){
+  # convert st to sp poly <- odx_polygons[2,]
+  t1 <- as(poly, 'Spatial')
+  # generate random points in that polygon
+  t2 <- spsample(t1, n = n,type = "random")
+  # compute distance of points to polygon boundary
+  gd <- gDistance(t2, as(t1, "SpatialLines"), byid = TRUE) 
+  return(as.vector(unname(gd)))
+}
+
+# run that fn for each poly in the study area
+odx_polygons_dists <- vector("list", length = nrow(odx_polygons))
+for(i in seq_len(length(odx_polygons_dists))){
+  odx_polygons_dists[[i]] <-random_points_to_boundary(odx_polygons[i,])
+  # print(i)
+}
+
+names(odx_polygons_dists) <- odx_polygons$TP_NO
+
+odx_polygons_dists <- as.data.frame(do.call('cbind', odx_polygons_dists))
+
+odx_polygons_dists <- stack(odx_polygons_dists)
+
+odx_polygons_dists <- separate(odx_polygons_dists, 
+                               ind, c("TP", "LF"), 
+                               sep = "/",
+                               remove = FALSE)
+
+odx_polygons_dists_max <- 
+odx_polygons_dists %>% 
+  group_by(TP) %>% 
+  dplyr::summarise(values = max(values))
+
+# ratio of actual site distance to boundary and max TP distance to boundary
+
+odx_polygons_dists_LF_max <- 
+  odx_polygons_dists %>% 
+  group_by(ind) %>% 
+  dplyr::summarise(values = max(values)) %>% 
+  mutate(TP_NO = ind)
+
+ODX_allsites_2010.s_with_rand <- 
+  ODX_allsites_2010.s %>% 
+    left_join(odx_polygons_dists_LF_max) %>% 
+    mutate(ratio = RASTERVALU/values)
+
+library(ggplot2)
+library(ggforce)
+r2b <- ggplot(ODX_allsites_2010.s_with_rand, 
+       aes(reorder(GREGLANDF, -ratio),
+           ratio)) + 
+  geom_boxplot() + 
+  scale_y_log10() + 
+  theme_bw(base_size = text_size-1) +
+  theme(axis.title.y = element_text(angle = 90))  +
+  geom_sina(alpha = 0.03) +
+  xlab("Terrain pattern") +
+  ylab("Ratio of site distance to boundary \nto maximum possible distance to boundary")
+
+grid.arrange(d2b, r2b, ncol = 2)
+
 ```

### 2017-06-06 analysis/paper/paper.Rmd
something about git2r
adding new figures and text in response to reviewers

@@
 First, in our study area these interfaces are relatively well-defined by landform attributes, such as the surface geology (Table \@ref(tab:table2)) and physical features such as the spacing between sand dune ridges or the presence of water bodies (Table \@ref(tab:table1)).
-The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes and bounaries other physical features, making them convenient for spatial analysis of archaeological features (Figure \@ref(fig:aerialphoto)).
+The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes and boundaries other physical features, making them convenient for spatial analysis of archaeological features (Figure \@ref(fig:aerialphoto)).
 With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.
@@
 This is because the rate of recording is strongly predicted by the density of sites (F(1, `r mobile_gis_efficiency_lm_df`) = `r mobile_gis_efficiency_lm_F`, R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`).
-To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) + `r mobile_gis_efficiency_lm_2007_2009_est`(site density) (`r mobile_gis_efficiency_lm_2007_2009_est_error`) +  (standard errors in parentheses).
+To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data, which resulted in a rate of site recording that can by predicted by the following model (standard errors in parentheses):
+
+rate = `r mobile_gis_efficiency_lm_2007_2009_est` (`r mobile_gis_efficiency_lm_2007_2009_est_error`) * site density + `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`)
+
 This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate using a mobile GIS given the density of 19 sites/km^2^ observed in the pre-2007 data.
@@
 Lower: Distribution of calliper lengths (longest distance between two points) of the terrain patterns in the study area.
-The Krukall-Wallis test in the upper right of the plot indicates that there is no significant difference in the lengths of the terrain patterns.", fig.width=fig_width_two_cols}
+The Kruskall-Wallis test in the upper right of the plot indicates that there is no significant difference in the lengths of the terrain patterns.", fig.width=fig_width_two_cols}
 
@@
 odx_polygons <- 
-  filter(odx_polygons,  GREGLANDF %in% c("A4", "Czs4", "Q4", "K4", "Czs5", "Q5", "Q6"))
+  dplyr::filter(odx_polygons,  GREGLANDF %in% c("A4", "Czs4", "Q4", "K4", "Czs5", "Q5", "Q6"))
 
@@
 # get coords of polygons to put into Momocs
-
-odx_polygons_coords <- apply(odx_polygons, 1,  function(i) st_coordinates(i$geometry))
 
@@
 ```
+
+Measurements of the distance from a site to the boundary of its containing terrain pattern could be strongly affected by the size and shape of the terrain pattern areas in our study area.
+For example, the minimum distance that any site could be the a boundary in a long, thin terrain pattern area would be much less than the minimum distance on a round or square area of the area.
+To ensure we can provide a meaningful comparison of boundary effects in the different terrain patterns, we tested to see if there were any significant differences in the elongation ratios and calliper lengths (also known as Feret diameters, or longest distance between two points of the shape) of all terrain pattern units in the study area.
+Figure \@ref(fig:tpelongationandlength) shows the distributions of elongation ratios and calliper lengths for each terrain pattern.
+The statistical hypothesis tests summarised in Figure \@ref(fig:tpelongationandlength) indicate that there are no signficant differences between the terrain patterns in the shapes of their polygons.
+Based on this result we conclude that the effect of the shapes of the terrain patterns on the distributions of distances from sites to boundaries does not greatly vary between different terrain patterns.
+This means that any boundary effects apparently in these data are not dependent simply on the shapes of the terrain patterns.
 
@@
 
-```{r maxmindistancetoboundary, fig.cap="Left: Distributions of distances of archaeological sites to terrain pattern bounaries.
+```{r maxmindistancetoboundary, fig.cap="Left: Distributions of distances of archaeological sites to terrain pattern boundaries.
+Distance is shown on the vertical axis with a logarithmic scale.
 Right: Distributions of ratios of observed distance of site to terrain pattern boundary and the maximum possible distance to boundary for each terrain pattern.
-Distance is shown on the vertical axis with a logarithmic scale.
 Data points show distance values for individual sites.", fig.width=fig_width_two_cols}
@@
 for(i in seq_len(length(odx_polygons_dists))){
-  odx_polygons_dists[[i]] <-random_points_to_boundary(odx_polygons[i,])
+  odx_polygons_dists[[i]] <- random_points_to_boundary(odx_polygons[i,])
   # print(i)
@@
   left_join(odx_polygons_dists_LF_max) %>% 
-    mutate(ratio = RASTERVALU/values)
+  mutate(ratio = RASTERVALU/values) %>% 
+  group_by(GREGLANDF) %>% 
+  mutate(z = scale(ratio)[,1])
 
@@
   scale_y_log10() + 
-  theme_bw(base_size = text_size-1) +
+  theme_bw(base_size = text_size - 1) +
   theme(axis.title.y = element_text(angle = 90))  +
@@
 
-Figure \@ref(fig:distanceofsitetoboundary) depicts distances of sites from the boundaries of the terrain pattern that contains those sites.
+The left panel of Figure \@ref(fig:maxmindistancetoboundary) depicts distances of sites from the boundaries of the terrain pattern that contains those sites.
+The right panel shows a ratio the observed distance of each site to its boundary and the maximum possible distance to boundary for any location within a its terrain pattern.
+We computed this maximum possible distance by simulating 1000 points randomly placed over each terrain pattern, measuring the distance to the boundary for each point, and taking the maximum distance from the boundary from this set of randomly distributed points.
+This ratio variable is relative to terrain pattern polygon shape and size for every site, and so attmpets the control for differences between the shapes of the terrain patterns.
 The sample of sites in this figure is limited to terrain pattern areas that are on average more than 1 km across (cf.
@@
 This means that sites tend to be nearer to the centers of landform type 4, but nearer to the boundaries of landform type 5.
-Visual inspection of Figure \@ref(fig:mapofstudyarea1) shows that this relationship is not dependent simply on the size or shape of the terrain patterns.
 
 ```{r sitesinbuffers, fig.cap="Site density by distance to the boundary of the terrain pattern for Czs4 and Czs5.
-Data points show site density values for each buffer zone.", fig.width=fig_width_two_cols}
+Data points show site density values for each buffer zone.
+Shaded region indicates the 95% confidence interval of a Loess regression line.", fig.width=fig_width_two_cols}
 
@@
                                       !inner_outer$Buf %in% c("-100-000", "000-100"), ]
+two_way_anova_data <- inner_outer[inner_outer$GREGLANDF %in% c('Czs5','Czs4') &
+                                      !inner_outer$SiteDens > 200, ]
+
 two_way_anova_summary <- summary(fit <- aov(data = two_way_anova_data, 
@@
 
+```{r disttobdrybyneighbour1, fig.cap="Distribution of distances of archaeological sites to terrain pattern boundaries for sites located in Czs5, by the nearest neighbouring Terrain Pattern.
+Data points show distance values for individual sites.", fig.width=fig_width_two_cols}
+# how the sites were positioned in the Czs5 relative to which terrain patterns surrounded them?
+
+library(rgeos)
+
+# convert sites data frame to SPDF
+ODX_allsites_spdf <- 
+SpatialPointsDataFrame(cbind(ODX_allsites_2010.s_with_rand$POINT_X,
+                             ODX_allsites_2010.s_with_rand$POINT_Y),
+                       data = as.data.frame(ODX_allsites_2010.s_with_rand),
+                       proj4string = CRS(st_crs(odx_polygons)$proj4string))
+
+# convert polys to sp
+odx_polygons_spdf <- as(odx_polygons, 'Spatial')
+
+# for each point, get distance to each polygon
+xx <- gDistance(odx_polygons_spdf,
+                ODX_allsites_spdf, 
+                byid=TRUE)
+
+# now find the closest poly(the one that contains the point), and the
+# second closest (the one on the other side of the closest boundary)
+
+xx <- data.frame(xx)
+names(xx) <- odx_polygons$GREGLANDF
+
+# get details of closest and second closest polygons
+xx_1st_min <- unlist(lapply(1:nrow(xx), function(i) order(xx[i,])[1]))
+xx_2nd_min <- unlist(lapply(1:nrow(xx), function(i) order(xx[i,])[2]))
+
+xx_1st_min_LF <- odx_polygons$GREGLANDF[xx_1st_min]
+xx_2nd_min_LF <- odx_polygons$GREGLANDF[xx_2nd_min]
+
+containing_and_neareest_poly <- 
+  data.frame(xx_1st_min_LF, 
+             xx_2nd_min_LF)
+
+# What about developing a new variable for each site within the Czs5 which describes the nature of the boundary (or the nearest other terrain pattern), and looking at how distance to boundary and density of sites in different buffer zones vary within and between these new groups?
+
+ODX_allsites_min_poly <- 
+  cbind(as.data.frame(ODX_allsites_spdf), 
+      containing_and_neareest_poly)
+
+ODX_allsites_min_poly_Czs5 <- 
+ODX_allsites_min_poly %>% 
+  filter(GREGLANDF == "Czs5")
+
+# see how distance to boundary varies by the LF of the neighbouring polygon
+
+#  hist(ODX_allsites_min_poly_Czs5$RASTERVALU) it's NOT normal
+dt_length_kruskal <- kruskal.test(ratio ~ xx_2nd_min_LF, 
+                                  data = ODX_allsites_min_poly_Czs5)
+# is a signifant difference
+
+hval <- unname(round(dt_length_kruskal$statistic, 3))
+pval <- round(dt_length_kruskal$p.value, 3)
+pval <- ifelse(pval == 0, "< 0.001", pval)
+df <-  unname(dt_length_kruskal$parameter)
+
+kruskal_string <- paste0("H(", df, ") = ", hval, ", p = ", pval)
+
+#  there is a sig result, so let's do a post-hoc test
+library(FSA)
+dt <- dunnTest(RASTERVALU ~ xx_2nd_min_LF,
+               data = ODX_allsites_min_poly_Czs5,
+               method = "bonferroni")
+
+what_pair_is_different <- dplyr::filter(dt$res, P.adj < 0.05)
+
+kruskal_string <- paste0(kruskal_string, 
+                         " (", what_pair_is_different$Comparison, ")")
+
+ggplot(ODX_allsites_min_poly_Czs5,
+       aes(reorder(xx_2nd_min_LF, 
+                   -RASTERVALU),
+           RASTERVALU)) +
+  geom_boxplot()  + 
+  geom_sina(alpha = 0.04) + 
+  theme_bw(base_size = text_size - 1) +
+  xlab("Neighbouring Terrain Pattern for sites located in Czs5") +
+  ylab("Distance from boundary (m)") +
+  # annotate("text",  x = "Q5", y = 1300, label = kruskal_string) +
+  scale_y_log10(breaks = seq(0,1400,200))
+
+```
+
 Figure \@ref(fig:sitesinbuffers) shows that in Czs4 landscapes the density of sites increases toward the center of the landform while in Czs5 landforms site density decreases towards the center of the landform.
-The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding the 0-100 m zone).
+The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding one outlier buffer (0-100m) in a Czs4 terrain pattern where the site density is anomalously high, more than twice the value of the next highest density).
 
@@
 We found both more archaeological sites per square kilometer, and more artefacts per square kilometer, in landforms with more widely spaced dunes.
-This is consistent with our model, and our field observations that more and denser sites tended to occur occur near exposures of good-quality flakeable stone that are often found in interdunal swales (Hughes et al.
+This is consistent with our model, and our field observations that more and denser sites tended to occur near exposures of good-quality flakeable stone that are often found in interdunal swales (Hughes et al.
 2014a).
 In areas where the sand dunes are more closely spaced, such as landform types 5 and 6, the dune flanks flow into each other, leaving few or no swales between the dunes where surface outcrops of flakable stone might be exposed.
-This differential availability of flakable stone may be an important factor that influenced ancient peoples' use of this landscape.
+This differential availability of flakable stone may be an important factor that influenced ancient people's use of this landscape.
 
@@
 We found that sites were generally located further from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
-This indicates that people preferred to be located close to the center of areas with widely-spared dunes, but only on the edges of the areas with closely spaced dunes.
+This indicates that people preferred to be located close to the center of areas with widely-spaced dunes, but only on the edges of the areas with closely spaced dunes.
 Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.
@@
 The substantive archaeological contribution of this study has been to show that when choosing habitations, prehistoric people in the Roxby dunefield were sensitive to relatively small variations in the landscape, such as the spacing of dunes.
-The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites the reveal how people adapted to the landscape (cf.
+The implication of this finding is that even in relatively homogeneous landscapes, such as the desert investigated here, there are gradient-like patterns in the distribution of archaeological sites that reveal how people adapted to the landscape (cf.
 Attwell and Fletcher 1987).
@@
 # what commit is this file at?
-library(git2r)
+library("git2r")
 repo <- repository(path = "../..")
@@
 
-The current git commit of this file is `r last_commit@sha`, which is on the `r branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r when(last_commit)`.
+The current git commit of this file is `r last_commit@sha`, which is on the `r git2r::branches(repo)[[1]]@name` branch and was made by `r last_commit@committer@name` on `r git2r::when(last_commit)`.
 The current commit message is "`r last_commit@summary`".

### 2017-06-07 analysis/paper/paper.Rmd
resubmit

@@
 University of Washington
-Denny Hall 117, Box 353100
+Denny Hall, Box 353100
 Seattle, WA 98195-3100 USA
@@
 In this study we use the results from a sample of hand-held GIS-enabled archaeological site recordings to ask if it is possible to observe gradients of change in archaeological materials over distance to understand forager choices of the optimum locations for their activities in respect to environmental boundaries.
-We predict that prehistoric foragers were sensitive to subtle differences in resource structure as gradients across the landscape.
+We predict that prehistoric foragers were sensitive to subtle differences in resource structure as gradients across landscape boundaries.
 This approach may be contrasted with concepts of linear boundaries with step-wise shifts in habitability.
@@
 Our study area in the Roxby dunefield in northern South Australia (Figure \@ref(fig:overviewmap)), offers a unique opportunity to study landscape interfaces for two key reasons.
-First, in our study area these interfaces are relatively well-defined by landform attributes, such as the surface geology (Table \@ref(tab:table2)) and physical features such as the spacing between sand dune ridges or the presence of water bodies (Table \@ref(tab:table1)).
+First, in our study area these interfaces are relatively well-defined by landform attributes, such as the surface geology (Table \@ref(tab:table2)) and physical features such as the presence of water bodies or the spacing between sand dune ridges (Table \@ref(tab:table1)).
+Although clearly visible, the causes of variation in dune spacing are poorly understood.
+Previous work has associated sand transport rates and directions, wind regime characteristics, grain size patterns, topography, and substrates with dune spacing, however there are no simple relationships (Lancaster 1988; Wasson and Hyde 1983; Wasson et al.
+1988).
 The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes and boundaries other physical features, making them convenient for spatial analysis of archaeological features (Figure \@ref(fig:aerialphoto)).
@@
 
-```{r overviewmap, fig.cap="Map of region including the study area", fig.width=fig_width_one_col}
+```{r overviewmap, fig.cap="Map of Australia showing the location of the Roxby dunefield.", fig.width=fig_width_one_col}
 
@@
 
-```{r tp-q4, fig.cap="Aerial photograph showing Terrain Pattern Q4 in the foreground, consisting of widely spaced sand dunes with only poorly developed pans and no rock outcrop in the swales.
-The more closely spaced dunes in the background are in Terrain Pattern Q5", fig.width=fig_width_one_col}
+```{r tp-q4, fig.cap="Aerial photograph showing terrain pattern Q4 in the foreground, consisting of widely spaced sand dunes with only poorly developed pans and no rock outcrop in the swales.
+The more closely spaced dunes in the background are in terrain pattern Q5", fig.width=fig_width_one_col}
 
@@
 
-```{r tp-qs4, fig.cap="Aerial photograph showing Terrain Pattern Qs4, consisting of widely spaced dunes with well-developed pans and patchy areas of bare rocky ground exposed in the swales.
+```{r tp-qs4, fig.cap="Aerial photograph showing terrain pattern Qs4, consisting of widely spaced dunes with well-developed pans and patchy areas of bare rocky ground exposed in the swales.
 The exposed rock is predominantly silcrete.", fig.width=fig_width_two_cols}
@@
 
-```{r mapofstudyarea1, fig.cap="Map of our study area within the Roxby dunefield, showing landforms and terrain patterns.
+```{r mapofstudyarea1, fig.cap="Map of the study area within the Roxby dunefield, showing landforms and terrain patterns.
 Each polygon is labelled with 1-3 letters that denote the surface geological regime (see Table \\@ref(tab:table2) for more details), and a number that denotes the landform type (see Table \\@ref(tab:table1) for more details).", fig.width=fig_width_two_cols}
@@
 
-```{r mapofstudyarea2, fig.cap="Map of study area showing the distribution of archaeological sites sampled for this study", fig.width=fig_width_two_cols}
+```{r mapofstudyarea2, fig.cap="Map of the study area showing the distribution of archaeological sites sampled for this study", fig.width=fig_width_two_cols}
 
@@
 When considered as homogeneous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns.
-As predicted (Table \@ref(tab:table1)), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km^2^ compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km^2^ (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns)).
+As predicted in Table \@ref(tab:table1), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km^2^ compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km^2^ (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns)).
 Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km^2^, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km^2^.
@@
 Lower: Distribution of calliper lengths (longest distance between two points) of the terrain patterns in the study area.
-The Kruskall-Wallis test in the upper right of the plot indicates that there is no significant difference in the lengths of the terrain patterns.", fig.width=fig_width_two_cols}
+The Kruskal-Wallis test in the upper right of the plot indicates that there is no significant difference in the lengths of the terrain patterns.", fig.width=fig_width_two_cols}
 
@@
 Measurements of the distance from a site to the boundary of its containing terrain pattern could be strongly affected by the size and shape of the terrain pattern areas in our study area.
-For example, the minimum distance that any site could be the a boundary in a long, thin terrain pattern area would be much less than the minimum distance on a round or square area of the area.
+For example, the minimum distance that any site could be from the a boundary in a long, thin terrain pattern area would be much less than the minimum distance on a round or square area of the area.
 To ensure we can provide a meaningful comparison of boundary effects in the different terrain patterns, we tested to see if there were any significant differences in the elongation ratios and calliper lengths (also known as Feret diameters, or longest distance between two points of the shape) of all terrain pattern units in the study area.
@@
 Based on this result we conclude that the effect of the shapes of the terrain patterns on the distributions of distances from sites to boundaries does not greatly vary between different terrain patterns.
-This means that any boundary effects apparently in these data are not dependent simply on the shapes of the terrain patterns.
+This means that any boundary effects apparent in these data are not dependent simply on the shapes of the terrain patterns.
 
@@
 
+##### . 
+<!-- page break -->
+
+```{r tablepairwisecomparisons, dependson='distanceofsitetoboundary'}
+# set dependson=distanceofsitetoboundary so that this chunk updates
+# when the previous one is updated, if no dependson, then the cache will not update
+
+#  hist(ODX_allsites_2010.s$RASTERVALU) it's NOT normal
+dst_length_kruskal <- kruskal.test(RASTERVALU ~ GREGLANDF, 
+                                  data = ODX_allsites_2010.s)
+# is a signifant difference
+
+hval <- unname(round(dst_length_kruskal$statistic, 3))
+pval <- round(dst_length_kruskal$p.value, 3)
+pval <- ifelse(pval == 0, "< 0.001", paste0("= ", pval))
+df <-  unname(dst_length_kruskal$parameter)
+
+kruskal_string <- paste0("H(", df, ") = ", hval, ", p ", pval)
+
+#  there is a sig result, so let's do a post-hoc test
+library(FSA)
+dt_dst <- dunnTest(RASTERVALU ~ GREGLANDF,
+               data = ODX_allsites_2010.s,
+               method = "bonferroni")
+
+all_pair_dst <- dt_dst$res
+
+# make a grid of pairwise differnces
+all_pair_grid_dst <- 
+all_pair_dst %>% 
+  separate(Comparison, c("pairA", "pairB"), sep = " - ") %>% 
+  select(pairA, pairB,  P.adj) %>% 
+  spread(pairA, P.adj)
+
+tp_column_dst <- all_pair_grid_dst$pairB
+
+all_pair_grid_dst <- all_pair_grid_dst[ , names(all_pair_grid_dst) != 'pairB']
+
+# and for the z stat
+all_pair_grid_dst_z <- 
+all_pair_dst %>% 
+  separate(Comparison, c("pairA", "pairB"), sep = " - ") %>% 
+  select(pairA, pairB,  Z) %>% 
+  spread(pairA, Z) %>% 
+  select(-pairB)
+
+# update the values suitable for a table to display
+# replace tiny values with <0.05
+pair_wide_comparison_grid_dst <- 
+data.frame(sapply(all_pair_grid_dst[,1:ncol(all_pair_grid_dst)],  
+                  function(i) ifelse(i < 0.001, 
+                                      "<0.001", 
+                                       as.character(round(i, 3)))),  
+           stringsAsFactors = FALSE)
+
+# add in z values
+dunn_test_table_dst <- 
+as.data.frame(do.call(cbind, 
+                      lapply(1:ncol(all_pair_grid_dst_z), 
+                                    function(i)
+                                    paste0(round(all_pair_grid_dst_z[ , i], 3), 
+                                     " (", 
+                                     pair_wide_comparison_grid_dst[ , i], 
+                                     ")"  ) )), 
+              stringsAsFactors = FALSE)
+
+# remove NA and replace with blank
+dunn_test_table_dst[dunn_test_table_dst == "NA (NA)"] <- ""
+
+# add laels columns and move this label col to front
+dunn_test_table_dst_labeled <- 
+  cbind(" " = tp_column_dst, 
+        dunn_test_table_dst)
+
+names(dunn_test_table_dst_labeled) <- c("", names(all_pair_grid_dst))
+
+kable(dunn_test_table_dst_labeled,
+      align = 'r',
+      caption = "Summary of Dunn's Kruskal-Wallis multiple comparisons test results for the distributions of distances of sites to boundaries for each terrain pattern.
+The cell values are the Z statistic, with p-values in parentheses.")
+
+```
+
+```{r tablepairwisecomparisonsratio}
+#  hist(ODX_allsites_2010.s_with_rand$ratio) it's NOT normal
+dstr_length_kruskal <- kruskal.test(ratio ~ GREGLANDF, 
+                                    data = ODX_allsites_2010.s_with_rand)
+# is a signifant difference
+
+hvalr <- unname(round(dstr_length_kruskal$statistic, 3))
+pvalr <- round(dstr_length_kruskal$p.value, 3)
+pvalr <- ifelse(pvalr == 0, "< 0.001", paste0("= ", pvalr))
+dfr <-  unname(dstr_length_kruskal$parameter)
+
+kruskalr_string <- paste0("H(", dfr, ") = ", hvalr, ", p ", pvalr)
+
+#  there is a sig result, so let's do a post-hoc test
+library(FSA)
+dt_dstr <- dunnTest(ratio ~ GREGLANDF,
+               data = ODX_allsites_2010.s_with_rand,
+               method = "bonferroni")
+
+all_pair <- dt_dstr$res
+
+# make a grid of pairwise differnces
+all_pair_grid <- 
+all_pair %>% 
+  separate(Comparison, c("pairA", "pairB"), sep = " - ") %>% 
+  select(pairA, pairB,  P.adj) %>% 
+  spread(pairA, P.adj)
+
+tp_column <- all_pair_grid$pairB
+
+all_pair_grid <- all_pair_grid[ , names(all_pair_grid) != 'pairB']
+
+all_pair_grid_z <- 
+all_pair %>% 
+  separate(Comparison, c("pairA", "pairB"), sep = " - ") %>% 
+  select(pairA, pairB,  Z) %>% 
+  spread(pairA, Z) %>% 
+  select(-pairB)
+
+# update the values suitable for a table to display
+# replace tiny values with <0.05
+pair_wide_comparison_grid_ratios <- 
+data.frame(sapply(all_pair_grid[,1:ncol(all_pair_grid)],  
+                  function(i) ifelse(i < 0.001, 
+                                      "<0.001", 
+                                       as.character(round(i, 3)))),    
+           stringsAsFactors = FALSE)
+
+# add in z values
+dunn_test_table <- 
+as.data.frame(do.call(cbind, 
+                      lapply(1:ncol(all_pair_grid_z), 
+                                    function(i)
+                                    paste0(round(all_pair_grid_z[ , i], 3), 
+                                     " (", 
+                                     pair_wide_comparison_grid_ratios[ , i], 
+                                     ")"  ) )), 
+              stringsAsFactors = FALSE)
+
+# remove NA and replace with blank
+dunn_test_table[dunn_test_table == "NA (NA)"] <- ""
+
+# add laels columns and move this label col to front
+dunn_test_table_labeled <- 
+  cbind(" " = tp_column, 
+        dunn_test_table)
+
+names(dunn_test_table_labeled) <- c("", names(all_pair_grid))
+
+kable(dunn_test_table_labeled,
+      align = 'r',
+      caption = "Summary of Dunn's Kruskal-Wallis multiple comparisons test results for the distributions of minimum distance ratios of sites to boundaries for each terrain pattern.
+The cell values are the Z statistic, with p-values in parentheses.")
+```
+
+```{r k4exception}
+n_and_percent_of_sites <- 
+ODX_allsites_2010.s_with_rand %>% 
+  dplyr::count(GREGLANDF) %>%
+  mutate(prop = n / sum(n),
+         perc = n / sum(n) * 100)
+
+k4_percent_of_all_sites <- 
+ n_and_percent_of_sites %>% 
+  filter(GREGLANDF == "K4")
+
+k4_n <- k4_percent_of_all_sites$n
+k4_perc <- round(k4_percent_of_all_sites$perc, 1)
+
+k4 <- paste0("(n = ", k4_n, ", ", k4_perc, "%)" )
+
+```
+
 The left panel of Figure \@ref(fig:maxmindistancetoboundary) depicts distances of sites from the boundaries of the terrain pattern that contains those sites.
+There are statistically significant differences between the terrain patterns (`r kruskal_string`).
 The right panel shows a ratio the observed distance of each site to its boundary and the maximum possible distance to boundary for any location within a its terrain pattern.
-We computed this maximum possible distance by simulating 1000 points randomly placed over each terrain pattern, measuring the distance to the boundary for each point, and taking the maximum distance from the boundary from this set of randomly distributed points.
+The differences is the distributions of ratios between the terrain patterns are also statistically significant (`r kruskalr_string`).
+We computed the maximum possible distance by simulating 1000 points randomly placed over each terrain pattern, then measuring the distance to the boundary for each point, and then taking the maximum distance from the boundary from this set of randomly distributed points.
 This ratio variable is relative to terrain pattern polygon shape and size for every site, and so attmpets the control for differences between the shapes of the terrain patterns.
-The sample of sites in this figure is limited to terrain pattern areas that are on average more than 1 km across (cf.
+Tables \@ref(tab:tablepairwisecomparisons) and \@ref(tab:tablepairwisecomparisonsratio) show the bonferroni-adjusted p-values from Dunn's (1964) test of multiple comparisons following the significant Kruskal-Wallis tests.
+These tables indicate that the differences in the values observed in Figure \@ref(fig:maxmindistancetoboundary) are unikely to result from chance alone The sample of sites in this figure is limited to terrain pattern areas that are on average more than 1 km across (cf.
 Figure \@ref(fig:mapofstudyarea1)).
@@
 A4, Czs4 and Q4) than in more closely spaced dunefields such as landform types 5 and 6 (e.g.
-Czs5, Q5and Q5).
+Czs5, Q5 and Q5).
 This means that sites tend to be nearer to the centers of landform type 4, but nearer to the boundaries of landform type 5.
+K4 appears to be the exception to this, probably because it has the smallest number of sites  in our sample `r k4`, an order of magnitude less than Czs4 and Czs5, so we do not have a robust indication of the true distribution of sites in K4.
 
@@
 
-```{r zscoresonsitedensitybyzones, fig.cap="Z-scores for site density in interior buffer zones for Czs4 and Czs5", fig.width=fig_width_one_col}
+```{r zscoresonsitedensitybyzones, fig.cap="Z-scores for site density in interior buffer zones for Czs4 and Czs5.
+The values for the interior buffer zones indicate the distance in m from the boundary of the terrain pattern to the center of the terrain pattern.", fig.width=fig_width_one_col}
 
@@
   theme(axis.title.y = element_text(angle = 90)) + 
-  xlab("interior buffer region (distance in m from \nboundary to the center of the terrain pattern)") + 
+  xlab("Interior buffer region (m)") + 
   ylab("z-score of site density \nin buffer area")  +
@@
 
-```{r disttobdrybyneighbour1, fig.cap="Distribution of distances of archaeological sites to terrain pattern boundaries for sites located in Czs5, by the nearest neighbouring Terrain Pattern.
+```{r disttobdrybyneighbour1, fig.cap="Distribution of distances of archaeological sites to terrain pattern boundaries for sites located in Czs5, by the nearest neighbouring terrain pattern.
 Data points show distance values for individual sites.", fig.width=fig_width_two_cols}
@@
 pval <- round(dt_length_kruskal$p.value, 3)
-pval <- ifelse(pval == 0, "< 0.001", pval)
+pval <- ifelse(pval == 0, "< 0.001", paste0("= ", pval))
 df <-  unname(dt_length_kruskal$parameter)
 
-kruskal_string <- paste0("H(", df, ") = ", hval, ", p = ", pval)
+kruskal_string <- paste0("H(", df, ") = ", hval, ", p ", pval)
 
@@
 what_pair_is_different <- dplyr::filter(dt$res, P.adj < 0.05)
-
-kruskal_string <- paste0(kruskal_string, 
-                         " (", what_pair_is_different$Comparison, ")")
 
@@
   theme_bw(base_size = text_size - 1) +
-  xlab("Neighbouring Terrain Pattern for sites located in Czs5") +
+  xlab("Neighbouring terrain pattern for sites located in Czs5") +
   ylab("Distance from boundary (m)") +
@@
 These analyses give support to the prediction that resource-rich landscapes such as Csz4 have sites located away from their margin and far inside the terrain pattern areas while landscapes with lower resource levels such as Csz5 commonly have sites closer to their margins and fewer sites further in their interiors.
+If we consider just sites within Czs5 (Figure \@ref(fig:disttobdrybyneighbour1)), we see that sites tend to be closer to the boundary when the neighbouring terrain pattern is landform four, such as Czs4, however this is not a statistically significant difference (`r kruskal_string`) .
 
@@
 We found that sites were generally located further from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6.
-This indicates that people preferred to be located close to the center of areas with widely-spaced dunes, but only on the edges of the areas with closely spaced dunes.
-Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.
+This indicates that preferred to be located away from the boundary of areas with widely spaced dunes, but only on the edges of the areas with closely spaced dunes.
+Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas further away from the boundary of the terrain patterns with areas 'closer to from the boundary of the terrain patterns.
 
@@
 We speculate that this is due to the relationship between dune spacing and the availability of flakable stone, as well as other resources.
-The causes of variation in dune spacing are poorly understood.
-Previous work has associated sand transport rates and directions, wind regime characteristics, grain size patterns, topography, and substrates with dune spacing, however there are no simple relationships (Lancaster 1988; Wasson and Hyde 1983; Wasson et al.
-1988).
 
@@
 
+Dunn, O.J. 1964. "Multiple comparisons using rank sums". Technometrics 6:241-252.
+
 Dunnell, R. C. and W. S. Dancey (1983). "The Siteless Survey: A Regional Scale Data Collection Strategy." Advances in Archaeological Method and Theory 6: 267-287.
@@
 Computational reproducibility in archaeological research: Basic principles and a case study of their implementation.
-Journal of Archaeological Method and Theory, 1-27, doi: 10.1007/s10816-015-9272-9
+Journal of Archaeological Method and Theory, 1-27, doi: 10.1007/s10816-015-9272-9.
+
 Napton, L.

### 2017-07-07 README.Rmd
update readme

@@
 ```
+
+## Research compendium for a report on Landform boundary effects on Holocene forager landscape use in arid South Australia
+
 [![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](https://github.com/benmarwick/olympicdamboundaries/commits/master) 
@@
 [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0001--7879--4531-green.svg)](http://orcid.org/0000-0001-7879-4531)
-
-## Research compendium for a report on xxxx
 
@@
 
-Marwick, B,  xxxxx
+> Marwick, B, P.
+Hiscock, M.
+Sullivan, P.
+Hughes.
+2017.
+Landform boundary effects on Holocene forager landscape use in arid South Australia.
+_Journal of Archaeological Science: Reports_ http://dx.doi.org/10.1016/j.jasrep.2017.07.004
 
@@
 
-This repository is our research compendium for our analysis of xxxx.
+This repository is our research compendium for our analysis of Landform boundary effects on Holocene forager landscape use in arid South Australia.
 The compendium contains all code, and text associated with the publication (data are not included for reasons explained in the manuscript).
